---
title: "16S_analysis_&_visualization"
author: "Amelia Slusser"
date: "2026-05-07"
output: html_document
---
Load packages
```{r}
library(dplyr)
library(purrr)
library(readr)
library(tidyr)
library(phyloseq)
library(stringr)
library(tibble)
library(ggplot2)
library(scales)
library(viridis)
library(RColorBrewer)
library(vegan)
library(ape)
library(Maaslin2)
library(rlang)
library(microbiome)
library(ggh4x)
library(writexl)
```

Load in the data
```{r}
# List all EMU output files
files <- list.files(base_dir, pattern = ".tsv$", full.names = TRUE)
samples <- gsub(".tsv", "", basename(files))

# Loop through files and rename estimated counts column
for (file in files) {
  
  df <- read.delim(file, sep = "\t", check.names = FALSE)
  
  # Rename the column if it exists
  colnames(df)[colnames(df) == "estimated counts"] <- "estimated_counts"
  
  # Overwrite the file with updated column name
  write.table(df,
              file = file,
              sep = "\t",
              quote = FALSE,
              row.names = FALSE)
}

#upload meta data to merge with samples and reads
meta <- read.table("meta_data.txt", header = TRUE, sep = "\t", stringsAsFactors = FALSE)
```

Clean up the data
```{r}
# Read all files into a single long-format data frame (memory-efficient)
long_df1 <- map2_dfr(files, samples, function(file, samp) {
  read_tsv(file, col_types = cols()) %>%
    select(tax_id, lineage, estimated_counts) %>%
    mutate(sample = samp)
})
# Remove unclassified/unmapped
long_df2 <- long_df1 %>%
  filter(!tax_id %in% c("mapped_unclassified", "unmapped"))
# Replace NA with 0
long_df2$estimated_counts[is.na(long_df2$estimated_counts)] <- 0
# Keep only one row per tax_id per sample (first occurrence)
long_df3 <- long_df2 %>%
  distinct(tax_id, lineage, sample, .keep_all = TRUE)
# Convert to wide format (samples as columns)
merged_df <- long_df3 %>%
  pivot_wider(names_from = sample, values_from = estimated_counts, values_fill = 0)
# Optional: sort by tax_id
merged_df <- merged_df %>% arrange(tax_id)
#collapse repeated taxIDs by deleting any repeated rows
merged_df_unique <- merged_df %>%
  distinct(tax_id, .keep_all = TRUE)
# Extract abundance columns
counts_cols <- setdiff(names(merged_df_unique), c("tax_id", "lineage"))
# Step 2: Create OTU matrix
otu_mat <- as.matrix(merged_df_unique %>% select(-tax_id, -lineage))
# Step 3: Assign rownames from unique tax_ids
rownames(otu_mat) <- merged_df_unique$tax_id
```

Relative Abundance for MaAsLin2 analysis
```{r}
# Read all files into a single long-format data frame (memory-efficient)
a_long_df1 <- map2_dfr(files, samples, function(file, samp) {
  read_tsv(file, col_types = cols()) %>%
    select(tax_id, lineage, abundance) %>%
    mutate(sample = samp)
})
# Remove unclassified/unmapped
a_long_df2 <- a_long_df1 %>%
  filter(!tax_id %in% c("mapped_unclassified", "unmapped"))
# Replace NA with 0
a_long_df2$abundance[is.na(a_long_df2$abundance)] <- 0
# Keep only one row per tax_id per sample (first occurrence)
a_long_df3 <- a_long_df2 %>%
  distinct(tax_id, lineage, sample, .keep_all = TRUE)
# Convert to wide format (samples as columns)
a_merged_df <- a_long_df3 %>%
  pivot_wider(names_from = sample, values_from = abundance, values_fill = 0)
# Optional: sort by tax_id
a_merged_df <- a_merged_df %>% arrange(tax_id)
#collapse repeated taxIDs by deleting any repeated rows
a_merged_df_unique <- a_merged_df %>%
  distinct(tax_id, .keep_all = TRUE)
# Extract abundance columns
a_counts_cols <- setdiff(names(a_merged_df_unique), c("tax_id", "lineage"))
# Step 2: Create OTU matrix
a_otu_mat <- as.matrix(a_merged_df_unique %>% select(-tax_id, -lineage))
# Step 3: Assign rownames from unique tax_ids
rownames(a_otu_mat) <- a_merged_df_unique$tax_id
```

Save merged files
```{r}
write_tsv(merged_df_unique, "merged_df_final.tsv")
```

Save merged files (relative abundance)
```{r}
write_tsv(a_merged_df_unique, "merged_df_final_abundance.tsv")
```

Create a phyloseq object
```{r}
# Define the 6 (mouse database) main ranks
ranks <- c("Kingdom","Phylum","Order","Family","Genus","Species")

tax_mat <- merged_df_unique %>%
  select(tax_id, lineage) %>%
  mutate(
    # Split lineage by ';' and remove any empty elements
    lineage_split = map(lineage, ~ {
      pieces <- strsplit(.x, ";")[[1]]
      pieces <- pieces[pieces != ""]   # remove empty strings
      # Take last 6 elements if longer
      vals <- tail(pieces, 6)
      # Pad with NA on left if shorter than 6
      if(length(vals) < 6) vals <- c(rep(NA, 6 - length(vals)), vals)
      rev(vals)  # reverse so Kingdom is first
    })
  ) %>%
  
  # Expand list-column into separate columns
  unnest_wider(lineage_split, names_sep = "_") %>%
  rename_with(~ ranks, starts_with("lineage_split_")) %>%
  mutate(across(everything(), ~ trimws(.))) %>%
  as.data.frame()

tax_mat <- tax_mat %>%
  select(-lineage)

tax_mat <- tax_mat %>%
  mutate(Species = ifelse(
    !is.na(Species) & !str_detect(Species, "\\s"),
    paste0(Species, " sp."),
    Species
  ))

rownames(tax_mat) <- merged_df_unique$tax_id

tax_mat_phy <- as.matrix(tax_mat)

# Set rownames to tax_id
rownames(otu_mat) <- merged_df_unique$tax_id

tax_mat_phy <- as.matrix(tax_mat)

# Prepare metadata
# Set sample_ID as rownames (for phyloseq)
meta <- meta %>% column_to_rownames("sample_ID")

# Check order against your OTU matrix to make sure they match
all(rownames(meta) %in% colnames(otu_mat))  # should be TRUE

# Check that rownames for OTU and TAX are identical
identical(
  rownames(otu_mat),
  rownames(tax_mat)
)

#now make your phyloseq object
OTU <- otu_table(otu_mat, taxa_are_rows = TRUE)   # TRUE if rows are taxa, FALSE if samples
TAX <- tax_table(as.matrix(tax_mat_phy))
META <- sample_data(meta)

ps <- phyloseq(OTU, TAX, META)

#check a few things 
ps
sample_names(ps)[1:5]
taxa_names(ps)[1:5]
sample_data(ps)[1:5, ]

ntaxa(ps)
```

Determine read depth of each sample as quality control measure
```{r}
# ============================================================
# DETERMINE READ DEPTH OF EACH SAMPLE BY SEX
# ============================================================

# Get total sequencing depth for every sample
sums <- sample_sums(ps)

# Overall summary
summary(sums)
hist(
  sums,
  breaks = 50,
  main = "Sequencing depth per sample",
  xlab = "Reads"
)

# Overall useful numbers
min(sums)
median(sums)
quantile(sums, probs = c(0.05, 0.10, 0.25))


# ------------------------------------------------------------
# Create sequencing-depth data frame and add metadata
# ------------------------------------------------------------

depth_df <- data.frame(
  sample_ID = sample_names(ps),
  Reads = as.numeric(sample_sums(ps)),
  stringsAsFactors = FALSE
)

# Extract metadata as a normal data frame
meta_df <- as.data.frame(sample_data(ps))
meta_df$sample_ID <- rownames(meta_df)

# Join sequencing depth with metadata
depth_df <- depth_df %>%
  left_join(meta_df, by = "sample_ID")

# Standardize Sex as a factor
depth_df <- depth_df %>%
  mutate(
    Sex = factor(
      str_to_title(as.character(Sex)),
      levels = c("F", "M")
    )
  )

# Check that all samples received a sex assignment
table(depth_df$Sex, useNA = "ifany")

# Inspect final data
head(depth_df)

# Export sample read depth
write_xlsx(depth_df, "sample_read_depth.xlsx")
# ------------------------------------------------------------
# Split sequencing-depth data by sex
# ------------------------------------------------------------

female_depth_df <- depth_df %>%
  filter(Sex == "F")

male_depth_df <- depth_df %>%
  filter(Sex == "M")


# ------------------------------------------------------------
# Sex-specific sequencing-depth summaries
# ------------------------------------------------------------

female_depth_summary <- female_depth_df %>%
  summarize(
    Number_of_samples = n(),
    Minimum_reads = min(Reads, na.rm = TRUE),
    Percentile_5 = quantile(Reads, 0.05, na.rm = TRUE),
    Percentile_10 = quantile(Reads, 0.10, na.rm = TRUE),
    Percentile_25 = quantile(Reads, 0.25, na.rm = TRUE),
    Median_reads = median(Reads, na.rm = TRUE),
    Mean_reads = mean(Reads, na.rm = TRUE),
    Maximum_reads = max(Reads, na.rm = TRUE)
  )

male_depth_summary <- male_depth_df %>%
  summarize(
    Number_of_samples = n(),
    Minimum_reads = min(Reads, na.rm = TRUE),
    Percentile_5 = quantile(Reads, 0.05, na.rm = TRUE),
    Percentile_10 = quantile(Reads, 0.10, na.rm = TRUE),
    Percentile_25 = quantile(Reads, 0.25, na.rm = TRUE),
    Median_reads = median(Reads, na.rm = TRUE),
    Mean_reads = mean(Reads, na.rm = TRUE),
    Maximum_reads = max(Reads, na.rm = TRUE)
  )

female_depth_summary
male_depth_summary

# ------------------------------------------------------------
# Use a common y-axis maximum for direct comparison
# ------------------------------------------------------------

depth_break_interval <- 20000

depth_y_max <- ceiling(
  max(depth_df$Reads, na.rm = TRUE) / depth_break_interval
) * depth_break_interval


# ------------------------------------------------------------
# Female sequencing-depth plot
# ------------------------------------------------------------

female_reads_plot <- ggplot(
  female_depth_df,
  aes(x = Group, y = Reads, fill = Group)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.5
  ) +
  geom_jitter(
    aes(color = Group),
    width = 0.2,
    size = 2,
    alpha = 0.8
  ) +
  theme_bw() +
  labs(
    title = "Sequencing Depth per Sample - Female Large Intestine",
    x = "Group",
    y = "Read Counts"
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(
    labels = comma,
    breaks = seq(
      0,
      depth_y_max,
      by = depth_break_interval
    ),
    limits = c(0, depth_y_max),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

female_reads_plot

ggsave(
  "female_reads_plot.png",
  plot = female_reads_plot,
  width = 8,
  height = 6,
  dpi = 600
)


# ------------------------------------------------------------
# Male sequencing-depth plot
# ------------------------------------------------------------

male_reads_plot <- ggplot(
  male_depth_df,
  aes(x = Group, y = Reads, fill = Group)
) +
  geom_boxplot(
    outlier.shape = NA,
    alpha = 0.5
  ) +
  geom_jitter(
    aes(color = Group),
    width = 0.2,
    size = 2,
    alpha = 0.8
  ) +
  theme_bw() +
  labs(
    title = "Sequencing Depth per Sample - Male Large Intestine",
    x = "Group",
    y = "Read Counts"
  ) +
  scale_fill_brewer(palette = "Set2") +
  scale_color_brewer(palette = "Set2") +
  scale_y_continuous(
    labels = comma,
    breaks = seq(
      0,
      depth_y_max,
      by = depth_break_interval
    ),
    limits = c(0, depth_y_max),
    expand = expansion(mult = c(0, 0.05))
  ) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

male_reads_plot

ggsave(
  "male_reads_plot.png",
  plot = male_reads_plot,
  width = 8,
  height = 6,
  dpi = 600
)


# ============================================================
# RAREFACTION CURVES BY SEX
# ============================================================

# Extract OTU table and convert to samples x taxa
otu_df <- as.data.frame(t(as(otu_table(ps), "matrix")))

# Make sample IDs an explicit column
otu_df$Sample <- rownames(otu_df)

head(otu_df)


# ------------------------------------------------------------
# Rarefaction function for an individual sample
# ------------------------------------------------------------

rarefaction_points <- function(x, step = 5000) {
  
  # Ensure counts are whole, nonnegative numbers
  x <- round(as.numeric(x))
  x[is.na(x)] <- 0
  x[x < 0] <- 0
  
  total_reads <- sum(x)
  
  # Return no rows for samples with no reads
  if (total_reads == 0) {
    return(
      data.frame(
        Depth = numeric(0),
        Observed = numeric(0)
      )
    )
  }
  
  # Generate depths and make sure the total depth is included
  seqs <- unique(
    c(
      seq(1, total_reads, by = step),
      total_reads
    )
  )
  
  observed <- sapply(seqs, function(n) {
    
    sampled_taxa <- sample(
      rep(seq_along(x), times = x),
      size = n,
      replace = FALSE
    )
    
    length(unique(sampled_taxa))
  })
  
  data.frame(
    Depth = seqs,
    Observed = observed
  )
}


# Set seed so rarefaction results are reproducible
set.seed(123)


# ------------------------------------------------------------
# Calculate rarefaction curves for all samples
# ------------------------------------------------------------

taxa_columns <- setdiff(colnames(otu_df), "Sample")

rarefaction_df <- map_dfr(
  seq_len(nrow(otu_df)),
  function(i) {
    
    sample_counts <- as.numeric(
      otu_df[i, taxa_columns, drop = TRUE]
    )
    
    df <- rarefaction_points(
      sample_counts,
      step = 5000
    )
    
    df$Sample <- otu_df$Sample[i]
    
    df
  }
)

table(rarefaction_df$Sample)


# ------------------------------------------------------------
# Add metadata
# ------------------------------------------------------------

metadata <- as.data.frame(sample_data(ps))
metadata$Sample <- rownames(metadata)

rarefaction_df <- rarefaction_df %>%
  left_join(metadata, by = "Sample") %>%
  mutate(
    Sex = factor(
      str_to_title(as.character(Sex)),
      levels = c("F", "M")
    )
  )

# Check metadata assignments
table(
  rarefaction_df$Sex,
  rarefaction_df$Treatment,
  useNA = "ifany"
)


# ------------------------------------------------------------
# Split rarefaction data by sex
# ------------------------------------------------------------

female_rarefaction_df <- rarefaction_df %>%
  filter(Sex == "F")

male_rarefaction_df <- rarefaction_df %>%
  filter(Sex == "M")


# ------------------------------------------------------------
# Use the same x-axis maximum for male and female plots
# ------------------------------------------------------------

rarefaction_break_interval <- 25000

rarefaction_x_max <- ceiling(
  max(rarefaction_df$Depth, na.rm = TRUE) /
    rarefaction_break_interval
) * rarefaction_break_interval


# ============================================================
# INDIVIDUAL-SAMPLE RAREFACTION CURVES
# ============================================================

# ------------------------------------------------------------
# Female individual-sample rarefaction plot
# ------------------------------------------------------------

female_rar_sample <- ggplot(
  female_rarefaction_df,
  aes(
    x = Depth,
    y = Observed,
    color = Group,
    group = Sample
  )
) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    title = "Female Rarefaction Curves",
    x = "Sequencing Depth",
    y = "Observed Species",
    color = "Group"
  ) +
  scale_color_brewer(
    palette = "Set2",
    breaks = c("Control", "Stress"),
    labels = c("Control", "Stress")
  ) +
  scale_x_continuous(
    labels = comma,
    breaks = seq(
      0,
      rarefaction_x_max,
      by = rarefaction_break_interval
    ),
    limits = c(0, rarefaction_x_max),
    expand = expansion(mult = c(0, 0.02))
  ) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

female_rar_sample

ggsave(
  "female_rarefaction_per_sample.png",
  plot = female_rar_sample,
  height = 8,
  width = 10,
  dpi = 600
)


# ------------------------------------------------------------
# Male individual-sample rarefaction plot
# ------------------------------------------------------------

male_rar_sample <- ggplot(
  male_rarefaction_df,
  aes(
    x = Depth,
    y = Observed,
    color = Group,
    group = Sample
  )
) +
  geom_line(linewidth = 1) +
  theme_bw() +
  labs(
    title = "Male Rarefaction Curves",
    x = "Sequencing Depth",
    y = "Observed Species",
    color = "Group"
  ) +
  scale_color_brewer(
    palette = "Set2",
    breaks = c("Control", "Stress"),
    labels = c("Control", "Stress")
  ) +
  scale_x_continuous(
    labels = comma,
    breaks = seq(
      0,
      rarefaction_x_max,
      by = rarefaction_break_interval
    ),
    limits = c(0, rarefaction_x_max),
    expand = expansion(mult = c(0, 0.02))
  ) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

male_rar_sample

ggsave(
  "male_rarefaction_per_sample.png",
  plot = male_rar_sample,
  height = 8,
  width = 10,
  dpi = 600
)


# ============================================================
# TREATMENT-AVERAGED RAREFACTION CURVES BY SEX
# ============================================================

# Calculate treatment averages separately within each sex
rarefaction_summary <- rarefaction_df %>%
  group_by(Sex, Treatment, Depth) %>%
  summarize(
    MeanObserved = mean(Observed, na.rm = TRUE),
    SDObserved = sd(Observed, na.rm = TRUE),
    Number_of_samples = n_distinct(Sample),
    .groups = "drop"
  )


# Split treatment summaries by sex
female_rarefaction_summary <- rarefaction_summary %>%
  filter(Sex == "F")

male_rarefaction_summary <- rarefaction_summary %>%
  filter(Sex == "M")

# ------------------------------------------------------------
# Female treatment-average rarefaction plot
# ------------------------------------------------------------

female_rar_avg <- ggplot() +
  geom_line(
    data = female_rarefaction_df,
    aes(
      x = Depth,
      y = Observed,
      group = Sample,
      color = Treatment
    ),
    alpha = 0.3,
    linewidth = 0.8
  ) +
  geom_line(
    data = female_rarefaction_summary,
    aes(
      x = Depth,
      y = MeanObserved,
      color = Treatment
    ),
    linewidth = 1.5
  ) +
  theme_bw() +
  labs(
    title = "Female Rarefaction Curves by Treatment",
    x = "Sequencing Depth",
    y = "Observed Species"
  ) +
  scale_x_continuous(
    labels = comma,
    breaks = seq(
      0,
      rarefaction_x_max,
      by = rarefaction_break_interval
    ),
    limits = c(0, rarefaction_x_max),
    expand = expansion(mult = c(0, 0.02))
  ) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

female_rar_avg

ggsave(
  "female_rarefaction_avg.png",
  plot = female_rar_avg,
  height = 8,
  width = 10,
  dpi = 600
)


# ------------------------------------------------------------
# Male treatment-average rarefaction plot
# ------------------------------------------------------------

male_rar_avg <- ggplot() +
  geom_line(
    data = male_rarefaction_df,
    aes(
      x = Depth,
      y = Observed,
      group = Sample,
      color = Treatment
    ),
    alpha = 0.3,
    linewidth = 0.8
  ) +
  geom_line(
    data = male_rarefaction_summary,
    aes(
      x = Depth,
      y = MeanObserved,
      color = Treatment
    ),
    linewidth = 1.5
  ) +
  theme_bw() +
  labs(
    title = "Male Rarefaction Curves by Treatment",
    x = "Sequencing Depth",
    y = "Observed Species"
  ) +
  scale_x_continuous(
    labels = comma,
    breaks = seq(
      0,
      rarefaction_x_max,
      by = rarefaction_break_interval
    ),
    limits = c(0, rarefaction_x_max),
    expand = expansion(mult = c(0, 0.02))
  ) +
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(size = 14)
  )

male_rar_avg

ggsave(
  "male_rarefaction_avg.png",
  plot = male_rar_avg,
  height = 8,
  width = 10,
  dpi = 600
)
```

################# Relative abundance workflow #################
Create the relative abundance phyloseq object
```{r}
ps_rel <- transform_sample_counts(ps, function(x) x / sum(x))
```

--------------------- Stacked barplots for taxonomy (Fig. 2B & C) ---------------------
Stacked bar plot function
```{r}
plot_taxa_barplot <- function(ps_obj,
                              tax_level = "Genus",
                              sex_label = "F",
                              output_file = NULL,
                              excel_file = NULL,
                              top_n = 20,
                              relative = TRUE,
                              palette = "Paired") {
  
  # Collapse to chosen taxonomic level
  ps_tax <- tax_glom(
    ps_obj,
    taxrank = tax_level
  )
  
  # Convert to long format
  df <- psmelt(ps_tax)
  
  # Ensure Group order
  df$Group <- factor(
    df$Group,
    levels = c("Control", "Stress")
  )
  
  # Filter by sex
  df_sex <- df %>%
    filter(Sex == sex_label)
  
  # Convert to relative abundance if desired
  if (relative) {
    df_sex <- df_sex %>%
      group_by(Sample) %>%
      mutate(
        Abundance = Abundance / sum(Abundance)
      ) %>%
      ungroup()
  }
  
  # Aggregate abundances
  df_agg <- df_sex %>%
    group_by(
      Sample,
      Group,
      !!sym(tax_level)
    ) %>%
    summarise(
      Abundance = sum(Abundance),
      .groups = "drop"
    )
  
    # Aggregate abundances
  df_agg <- df_sex %>%
    group_by(
      Sample,
      Group,
      !!sym(tax_level)
    ) %>%
    summarise(
      Abundance = sum(Abundance),
      .groups = "drop"
    )

  # Export abundance data to Excel
  if (!is.null(excel_file)) {

    # Long-format data
    abundance_long <- df_agg %>%
      rename(
        Taxon = !!sym(tax_level)
      ) %>%
      arrange(Group, Sample, desc(Abundance))

    # Wide-format data: one row per taxon, one column per sample
    abundance_wide <- abundance_long %>%
      select(Taxon, Sample, Abundance) %>%
      tidyr::pivot_wider(
        names_from = Sample,
        values_from = Abundance,
        values_fill = 0
      )

    writexl::write_xlsx(
      list(
        Relative_abundance_long = abundance_long,
        Relative_abundance_wide = abundance_wide
      ),
      path = excel_file
    )
  }
  
  # Identify top taxa
  top_taxa <- df_agg %>%
    group_by(!!sym(tax_level)) %>%
    summarise(
      TotalAbundance = sum(Abundance),
      .groups = "drop"
    ) %>%
    arrange(desc(TotalAbundance)) %>%
    slice_head(n = top_n) %>%
    pull(!!sym(tax_level))
  
  # Collapse lower-abundance taxa into "Other"
  df_top <- df_agg %>%
    mutate(
      Taxon = ifelse(
        !!sym(tax_level) %in% top_taxa,
        as.character(!!sym(tax_level)),
        "Other"
      )
    ) %>%
    group_by(Sample, Group, Taxon) %>%
    summarise(
      Abundance = sum(Abundance),
      .groups = "drop"
    )
  
  # Put Other last
  tax_levels <- c(
    sort(setdiff(unique(df_top$Taxon), "Other")),
    "Other"
  )
  
  df_top$Taxon <- factor(
    df_top$Taxon,
    levels = tax_levels
  )
  
  # Generate taxon colors
  n_tax <- length(tax_levels) - 1
  
  tax_colors <- brewer.pal(
    min(n_tax, 12),
    palette
  )
  
  if (n_tax > 12) {
    tax_colors <- colorRampPalette(
      brewer.pal(12, palette)
    )(n_tax)
  }
  
  # Add gray for the Other category
  tax_colors <- c(
    tax_colors,
    "gray60"
  )
  
  names(tax_colors) <- tax_levels
  
  # Define group colors
  group_colors <- c(
    "Control" = "#4D4D4D",
    "Stress"  = "#FE8A13"
  )
  
  # Plot
  p <- ggplot(
    df_top,
    aes(
      x = Sample,
      y = Abundance,
      fill = Taxon
    )
  ) +
    
    geom_bar(
      stat = "identity",
      color = "black",
      linewidth = 0.2
    ) +
    
    # Invisible layer used to generate the Group legend
    geom_point(
      data = df_top %>%
        distinct(Sample, Group),
      aes(
        x = Sample,
        y = 0,
        color = Group
      ),
      inherit.aes = FALSE,
      alpha = 0,
      show.legend = TRUE
    ) +
    
    # Group-specific facet ribbons
    ggh4x::facet_wrap2(
      ~ Group,
      nrow = 1,
      scales = "free_x",
      strip = ggh4x::strip_themed(
        
        background_x = ggh4x::elem_list_rect(
          fill = c(
            "#4D4D4D",
            "#FE8A13"
          ),
          color = "black",
          linewidth = 0.5
        ),
        
        # Keep the colored ribbons but hide their labels
        text_x = ggh4x::elem_list_text(
          color = "transparent",
          size = 20
        )
      )
    ) +
    
    scale_fill_manual(
      values = tax_colors,
      name = tax_level
    ) +
    
    scale_color_manual(
      values = group_colors,
      name = "Group"
    ) +
    
    ylab(
      ifelse(
        relative,
        "Relative Abundance",
        "Estimated Abundance"
      )
    ) +
    
    xlab("Sample") +
    
    ggtitle(
      paste0(
        tax_level,
        "-level Abundance - LI ",
        ifelse(
          sex_label == "F",
          "F",
          "M"
        )
      )
    ) +
    
    theme_bw(base_size = 18) +
    
    theme(
      # X-axis formatting
      axis.text.x = element_blank(),
      axis.ticks.x = element_blank(),
      axis.title.x = element_text(
        size = 20,
        face = "bold",
        margin = margin(t = 10)
      ),
      
      # Y-axis formatting
      axis.text.y = element_text(
        size = 17,
        color = "black"
      ),
      axis.title.y = element_text(
        size = 20,
        face = "bold",
        margin = margin(r = 10)
      ),
      
      # Plot title
      plot.title = element_text(
        hjust = 0.5,
        face = "bold",
        size = 22,
        margin = margin(b = 12)
      ),
      
      # Legend formatting
      legend.position = "right",
      legend.direction = "vertical",
      legend.box = "vertical",
      legend.title = element_text(
        size = 19,
        face = "bold"
      ),
      legend.text = element_text(size = 16),
      legend.key.size = grid::unit(0.7, "cm")
    ) +
    
    guides(
        # Group legend with thin black borders
        color = guide_legend(
          order = 1,
          override.aes = list(
            alpha = 1,
            shape = 22,
            size = 7,
            fill = unname(group_colors),
            color = "black",
            stroke = 0.6
          )
        ),
        
        # Taxon legend with thin black borders
        fill = guide_legend(
          order = 2,
          ncol = 1,
          override.aes = list(
            color = "black",
            linewidth = 0.4
          )
        )
      )
  
  # Save if filename is provided
  if (!is.null(output_file)) {
    ggsave(
      filename = output_file,
      plot = p,
      width = 10,
      height = 12,
      dpi = 600
    )
  }
  
  return(p)
}
```

Export relative abundance files
```{r}
write_xlsx(df_agg, "relative_abundance_genus.tsv")
```

Plot stacked bars
```{r}
# Plot females
stacked_bar_female <- plot_taxa_barplot(
  ps_obj = ps_rel,
  tax_level = "Genus",
  top_n = 20,
  sex_label = "F",
  output_file = "genus_female.png",
  excel_file = "genus_female_relative_abundance.xlsx"
)

# Plot males
stacked_bar_male <- plot_taxa_barplot(
  ps_obj = ps_rel,
  tax_level = "Genus",
  top_n = 20,
  sex_label = "M",
  output_file = "genus_male.png",
  excel_file = "genus_male_relative_abundance.xlsx"
)

# Optionally, display plots in R
print(stacked_bar_female)
print(stacked_bar_male)
```

--------------------- Beta diversity (Fig. 2D & E) ---------------------
Beta diversity female
```{r}
# Subset female samples
ps_f <- subset_samples(
  ps_rel,
  Sex == "F"
)

# Remove taxa that have zero abundance across all female samples
ps_f <- prune_taxa(
  taxa_sums(ps_f) > 0,
  ps_f
)

# Extract female metadata
meta_f <- as(
  sample_data(ps_f),
  "data.frame"
)

meta_f$sample_ID <- rownames(meta_f)

# Set group order
meta_f$Group <- factor(
  meta_f$Group,
  levels = c("Control", "Stress")
)

# Calculate Bray-Curtis distances
bray_dist_f <- phyloseq::distance(
  ps_f,
  method = "bray"
)

# Ensure metadata rows follow the same sample order as the distance object
meta_f <- meta_f[
  labels(bray_dist_f),
  ,
  drop = FALSE
]

# Confirm that sample ordering matches
stopifnot(
  identical(
    rownames(meta_f),
    labels(bray_dist_f)
  )
)

# -------------------------
# PCoA
# -------------------------

ord_bray_f <- ape::pcoa(
  bray_dist_f
)

# Extract the first two PCoA axes
ord_df_f <- as.data.frame(
  ord_bray_f$vectors[, 1:2]
)

colnames(ord_df_f) <- c(
  "Axis.1",
  "Axis.2"
)

ord_df_f$sample_ID <- rownames(
  ord_df_f
)

# Add sample metadata to the PCoA plotting data
ord_df_f <- merge(
  ord_df_f,
  meta_f,
  by = "sample_ID",
  sort = FALSE
)

# Restore the original PCoA sample order after merging
ord_df_f <- ord_df_f[
  match(
    rownames(ord_bray_f$vectors),
    ord_df_f$sample_ID
  ),
]

rownames(ord_df_f) <- NULL

# -------------------------
# PERMANOVA
# -------------------------

set.seed(123)

adonis_female <- vegan::adonis2(
  bray_dist_f ~ Group,
  data = meta_f,
  permutations = 9999
)

# Extract PERMANOVA p-value for plotting
pval_female <- adonis_female$`Pr(>F)`[1]

# View complete PERMANOVA results
adonis_female

# -------------------------
# Beta dispersion
# -------------------------

# Calculate distances to group centroids
bd_f <- vegan::betadisper(
  bray_dist_f,
  meta_f$Group
)

# Save the standard beta-dispersion ANOVA as an object
bd_anova_f <- anova(
  bd_f
)

# Save the permutation-based beta-dispersion test as an object
set.seed(123)

bd_permutest_f <- vegan::permutest(
  bd_f,
  permutations = 9999
)
```

Beta diversity males
```{r}
# Subset male samples
ps_m <- subset_samples(
  ps_rel,
  Sex == "M"
)

# Remove taxa that have zero abundance across all male samples
ps_m <- prune_taxa(
  taxa_sums(ps_m) > 0,
  ps_m
)

# Extract male metadata
meta_m <- as(
  sample_data(ps_m),
  "data.frame"
)

meta_m$sample_ID <- rownames(meta_m)

# Set group order
meta_m$Group <- factor(
  meta_m$Group,
  levels = c("Control", "Stress")
)

# Calculate Bray-Curtis distances
bray_dist_m <- phyloseq::distance(
  ps_m,
  method = "bray"
)

# Ensure metadata rows follow the same sample order as the distance object
meta_m <- meta_m[
  labels(bray_dist_m),
  ,
  drop = FALSE
]

# Confirm that sample ordering matches
stopifnot(
  identical(
    rownames(meta_m),
    labels(bray_dist_m)
  )
)

# -------------------------
# PCoA
# -------------------------

ord_bray_m <- ape::pcoa(
  bray_dist_m
)

# Extract the first two PCoA axes
ord_df_m <- as.data.frame(
  ord_bray_m$vectors[, 1:2]
)

colnames(ord_df_m) <- c(
  "Axis.1",
  "Axis.2"
)

ord_df_m$sample_ID <- rownames(
  ord_df_m
)

# Add sample metadata to the PCoA plotting data
ord_df_m <- merge(
  ord_df_m,
  meta_m,
  by = "sample_ID",
  sort = FALSE
)

# Restore the original PCoA sample order after merging
ord_df_m <- ord_df_m[
  match(
    rownames(ord_bray_m$vectors),
    ord_df_m$sample_ID
  ),
]

rownames(ord_df_m) <- NULL

# -------------------------
# PERMANOVA
# -------------------------

set.seed(123)

adonis_male <- vegan::adonis2(
  bray_dist_m ~ Group,
  data = meta_m,
  permutations = 9999
)

# Extract PERMANOVA p-value for plotting
pval_male <- adonis_male$`Pr(>F)`[1]

# View complete PERMANOVA results
adonis_male

# -------------------------
# Beta dispersion
# -------------------------

# Calculate distances to group centroids
bd_m <- vegan::betadisper(
  bray_dist_m,
  meta_m$Group
)

# Save the standard beta-dispersion ANOVA as an object
bd_anova_m <- anova(
  bd_m
)

# Save the permutation-based beta-dispersion test as an object
set.seed(123)

bd_permutest_m <- vegan::permutest(
  bd_m,
  permutations = 9999
)
```

Stats and plotting data export function
```{r}
prepare_beta_export <- function(
    ord_df,
    ord_obj,
    distance_obj,
    adonis_result,
    betadisper_obj,
    betadisper_anova,
    betadisper_permutest,
    sex_label
) {
  
  sex_name <- ifelse(sex_label == "F", "Female", "Male")
  
  # -------------------------------------------------------
  # 1. Exact data used in the PCoA plot
  # -------------------------------------------------------
  
  plotting_data <- ord_df %>%
    mutate(
      Sex_analysis = sex_name,
      Group = as.character(Group)
    ) %>%
    relocate(
      sample_ID,
      Sex_analysis,
      Group,
      Axis.1,
      Axis.2
    ) %>%
    arrange(Group, sample_ID)
  
  # -------------------------------------------------------
  # 2. PERMANOVA results
  # -------------------------------------------------------
  
  permanova_results <- as.data.frame(adonis_result) %>%
    rownames_to_column("Term") %>%
    mutate(
      Sex_analysis = sex_name,
      Distance = "Bray-Curtis",
      Permutations = 9999,
      .before = 1
    )
  
  # -------------------------------------------------------
  # 3. Beta-dispersion ANOVA results
  # -------------------------------------------------------
  
  dispersion_anova_results <- as.data.frame(
    betadisper_anova
  ) %>%
    rownames_to_column("Term") %>%
    mutate(
      Sex_analysis = sex_name,
      Distance = "Bray-Curtis",
      .before = 1
    )
  
  # -------------------------------------------------------
  # 4. Permutation test of beta dispersion
  # -------------------------------------------------------
  
  dispersion_perm_results <- as.data.frame(
    betadisper_permutest$tab
  ) %>%
    rownames_to_column("Term") %>%
    mutate(
      Sex_analysis = sex_name,
      Distance = "Bray-Curtis",
      .before = 1
    )
  
  # -------------------------------------------------------
  # 5. Distance to group centroid for each sample
  # -------------------------------------------------------
  
  dispersion_sample_data <- data.frame(
    sample_ID = names(betadisper_obj$distances),
    Group = as.character(betadisper_obj$group),
    Distance_to_centroid = unname(
      betadisper_obj$distances
    ),
    Sex_analysis = sex_name
  ) %>%
    relocate(
      sample_ID,
      Sex_analysis,
      Group,
      Distance_to_centroid
    ) %>%
    arrange(Group, sample_ID)
  
  # -------------------------------------------------------
  # 6. PCoA eigenvalues and variance explained
  # -------------------------------------------------------
  
  variance_explained <- as.data.frame(
    ord_obj$values
  ) %>%
    mutate(
      Axis = paste0("PCoA_", row_number()),
      Sex_analysis = sex_name,
      .before = 1
    )
  
  # Convert proportions to percentages for readability
  if ("Relative_eig" %in% names(variance_explained)) {
    variance_explained <- variance_explained %>%
      mutate(
        Percent_explained = Relative_eig * 100
      )
  }
  
  # -------------------------------------------------------
  # 7. Bray-Curtis distance matrix
  # -------------------------------------------------------
  
  distance_matrix <- as.matrix(distance_obj) %>%
    as.data.frame(check.names = FALSE) %>%
    rownames_to_column("sample_ID")
  
  # -------------------------------------------------------
  # 8. Bray-Curtis distances in long format
  # -------------------------------------------------------
  
  distance_long <- as.matrix(distance_obj) %>%
    as.data.frame(check.names = FALSE) %>%
    rownames_to_column("Sample_1") %>%
    pivot_longer(
      cols = -Sample_1,
      names_to = "Sample_2",
      values_to = "Bray_Curtis_distance"
    ) %>%
    # Remove self-comparisons and duplicate pairs
    filter(Sample_1 < Sample_2) %>%
    mutate(
      Sex_analysis = sex_name,
      .before = 1
    ) %>%
    arrange(Sample_1, Sample_2)
  
  return(
    list(
      PCoA_Plotting_Data = plotting_data,
      PERMANOVA_Results = permanova_results,
      Dispersion_ANOVA = dispersion_anova_results,
      Dispersion_Permutation = dispersion_perm_results,
      Distances_to_Centroid = dispersion_sample_data,
      Variance_Explained = variance_explained,
      Bray_Curtis_Long = distance_long,
      Bray_Curtis_Matrix = distance_matrix
    )
  )
}
```

Run function to create file outputs
```{r}
female_beta_export <- prepare_beta_export(
  ord_df = ord_df_f,
  ord_obj = ord_bray_f,
  distance_obj = bray_dist_f,
  adonis_result = adonis_female,
  betadisper_obj = bd_f,
  betadisper_anova = bd_anova_f,
  betadisper_permutest = bd_permutest_f,
  sex_label = "F"
)

male_beta_export <- prepare_beta_export(
  ord_df = ord_df_m,
  ord_obj = ord_bray_m,
  distance_obj = bray_dist_m,
  adonis_result = adonis_male,
  betadisper_obj = bd_m,
  betadisper_anova = bd_anova_m,
  betadisper_permutest = bd_permutest_m,
  sex_label = "M"
)

write_xlsx(
  female_beta_export,
  path = "beta_diversity_female_data_and_statistics.xlsx"
)

write_xlsx(
  male_beta_export,
  path = "beta_diversity_male_data_and_statistics.xlsx"
)
```

Beta diversity plots 
```{r}
# --- Metadata ---
meta <- as(sample_data(ps_rel), "data.frame")
meta$sample_ID <- rownames(meta)

# --- Bray–Curtis distance ---
bray_dist <- phyloseq::distance(ps_rel, method = "bray")

# --- PCoA ---
ord_bray <- ape::pcoa(bray_dist)

# --- Make plotting data frame ---
ord_df <- as.data.frame(ord_bray$vectors[, 1:2])
colnames(ord_df) <- c("Axis.1", "Axis.2")
ord_df$sample_ID <- rownames(ord_df)
ord_df <- merge(ord_df, meta, by = "sample_ID")

# --- PERMANOVA ---
meta_f <- meta %>% filter(Sex == "F")
keep_f <- intersect(rownames(meta_f), rownames(as.matrix(bray_dist)))
bray_dist_f <- as.dist(as.matrix(bray_dist)[keep_f, keep_f])
meta_f <- meta_f[keep_f, ]
meta_f$Group <- factor(meta_f$Group, levels = c("Control", "Stress"))
adonis_female <- adonis2(bray_dist_f ~ Group, data = meta_f, permutations = 9999)
pval_female <- adonis_female$`Pr(>F)`[1]

meta_m <- meta %>% filter(Sex == "M")
keep_m <- intersect(rownames(meta_m), rownames(as.matrix(bray_dist)))
bray_dist_m <- as.dist(as.matrix(bray_dist)[keep_m, keep_m])
meta_m <- meta_m[keep_m, ]
meta_m$Group <- factor(meta_m$Group, levels = c("Control", "Stress"))
adonis_male <- adonis2(bray_dist_m ~ Group, data = meta_m, permutations = 9999)
pval_male <- adonis_male$`Pr(>F)`[1]

# --- Plot function ---
plot_beta_by_sex <- function(df, sex_label, pval) {
  
  df_sex <- df %>% filter(Sex == sex_label)
  eig1 <- round(ord_bray$values$Relative_eig[1] * 100, 1)
  eig2 <- round(ord_bray$values$Relative_eig[2] * 100, 1)
  
  # Colors
  fill_colors <- c("Control" = "#4D4D4D", "Stress" = "#FE8A13")
  line_colors <- c("Control" = "#4D4D4D", "Stress" = "#FE8A13")
  
  # Significance label
  sig_label <- ifelse(pval < 0.001, "***",
                      ifelse(pval < 0.01, "**",
                             ifelse(pval < 0.05, "*", "ns")))
  
  # Text color logic
  text_color <- ifelse(pval < 0.05, "black", "gray50")
  
  # Plot
  ggplot(df_sex, aes(x = Axis.1, y = Axis.2)) +
    geom_point(aes(fill = Group), shape = 21, size = 5, color = "gray50", alpha = 1.0) +
    stat_ellipse(aes(group = Group, fill = Group),
                 geom = "polygon", alpha = 0.25, color = NA,
                 level = 0.8, type = "norm") +
    #stat_ellipse(aes(group = Group, color = Group),
    #geom = "path", linetype = "solid", size = 0.5,
    #level = 0.9, type = "norm") +
    scale_fill_manual(values = fill_colors) +
    #scale_color_manual(values = line_colors) +
    annotate("text",
             x = Inf, y = Inf,
             label = paste0("PERMANOVA, ", sig_label, "  p=", signif(pval, 3)),
             hjust = 1.1, vjust = 1.5,
             size = 5, fontface = "bold", color = text_color) +
    theme_bw() +
    labs(
      title = paste("Beta Diversity (PCoA - Bray–Curtis) - Large Intestine", 
                    ifelse(sex_label == "F", "Females", "Males")),
      x = paste0("PCoA 1 (", eig1, "%)"),
      y = paste0("PCoA 2 (", eig2, "%)")
    ) +
    theme(
      plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
      axis.text = element_text(size = 12),
      axis.title = element_text(size = 14, face = "bold"),
      legend.title = element_text(size = 12, face = "bold"),
      legend.text = element_text(size = 11)
    )
}


# --- Generate plots ---
p_beta_female <- plot_beta_by_sex(ord_df, "F", pval_female)
p_beta_male   <- plot_beta_by_sex(ord_df, "M", pval_male)

# --- Display ---
p_beta_female
p_beta_male

ggsave("beta_female.png", plot = p_beta_female, width = 9, height = 6, dpi = 600)
ggsave("beta_males.png", plot = p_beta_male, width = 8, height = 6, dpi = 600)
```

--------------------- Alpha diversity (Fig. S2C & D) ---------------------
Compute alpha diversity metrics
```{r}
otu_alpha <- as(otu_table(ps_rel), "matrix")

if(taxa_are_rows(ps_rel)) {
  otu_alpha <- t(otu_alpha)
}

shannon <- vegan::diversity(otu_alpha, index = "shannon")
simpson <- vegan::diversity(otu_alpha, index = "simpson")

```

Build alpha diversity data frame
```{r}
alpha_div <- data.frame(
  sample_ID = names(shannon),
  Shannon = shannon,
  Simpson = simpson
)
```

Add metadata
```{r}
meta <- data.frame(sample_data(ps_rel))

meta$sample_ID <- rownames(meta)

alpha_div <- alpha_div %>%
  left_join(meta, by = "sample_ID")

# Keep only samples with valid metadata
alpha_div <- alpha_div %>%
  filter(!is.na(Group),
         !is.na(Sex))
```

Factor ordering
```{r}
alpha_div$Group <- factor(
  alpha_div$Group,
  levels = c("Control", "Stress")
)

alpha_div$Sex <- factor(
  alpha_div$Sex,
  levels = c("F", "M")
)
```

Format for plotting
```{r}
alpha_long <- alpha_div %>%
  pivot_longer(
    cols = c("Shannon", "Simpson"),
    names_to = "Metric",
    values_to = "Value"
  ) %>%
  mutate(
    Metric = factor(
      Metric,
      levels = c("Shannon", "Simpson")
    )
  )
```

Function to prepare plotting data and statistical results, by sex
```{r}
prepare_alpha_export <- function(alpha_long, sex_label) {
  
  # Exact data used for plotting
  plotting_data <- alpha_long %>%
    filter(Sex == sex_label) %>%
    mutate(
      Sex = as.character(Sex),
      Group = factor(Group, levels = c("Control", "Stress")),
      Metric = factor(Metric, levels = c("Shannon", "Simpson"))
    ) %>%
    arrange(Metric, Group, sample_ID) %>%
    mutate(
      Group = as.character(Group),
      Metric = as.character(Metric)
    )
  
  # Wilcoxon test
  statistical_results <- lapply(
    c("Shannon", "Simpson"),
    function(metric_name) {
      
      df_metric <- plotting_data %>%
        filter(Metric == metric_name)
      
      control_values <- df_metric %>%
        filter(Group == "Control") %>%
        pull(Value)
      
      stress_values <- df_metric %>%
        filter(Group == "Stress") %>%
        pull(Value)
      
      # Return a row only when both groups have sufficient observations
      if (length(control_values) > 1 &&
          length(stress_values) > 1) {
        
        test_result <- wilcox.test(
          control_values,
          stress_values,
          exact = FALSE
        )
        
        p_value <- test_result$p.value
        
        significance <- case_when(
          p_value < 0.001 ~ "***",
          p_value < 0.01  ~ "**",
          p_value < 0.05  ~ "*",
          TRUE            ~ "ns"
        )
        
        data.frame(
          Sex = ifelse(sex_label == "F", "Female", "Male"),
          Metric = metric_name,
          Comparison = "Control vs Stress",
          Test = test_result$method,
          W_statistic = unname(test_result$statistic),
          p_value = p_value,
          Significance = significance,
          Control_n = length(control_values),
          Stress_n = length(stress_values),
          Control_mean = mean(control_values, na.rm = TRUE),
          Stress_mean = mean(stress_values, na.rm = TRUE),
          Control_SD = sd(control_values, na.rm = TRUE),
          Stress_SD = sd(stress_values, na.rm = TRUE),
          Control_median = median(control_values, na.rm = TRUE),
          Stress_median = median(stress_values, na.rm = TRUE),
          Control_IQR = IQR(control_values, na.rm = TRUE),
          Stress_IQR = IQR(stress_values, na.rm = TRUE)
        )
        
      } else {
        NULL
      }
    }
  ) %>%
    bind_rows()
  
  return(
    list(
      Plotting_Data = plotting_data,
      Wilcoxon_Results = statistical_results
    )
  )
}
```

Prepare and export results
```{r}
# Prepare female and male results
female_alpha_export <- prepare_alpha_export(
  alpha_long = alpha_long,
  sex_label = "F"
)

male_alpha_export <- prepare_alpha_export(
  alpha_long = alpha_long,
  sex_label = "M"
)

# Export each sex as a separate Excel workbook
write_xlsx(
  female_alpha_export,
  path = "alpha_diversity_female_data_and_statistics.xlsx"
)

write_xlsx(
  male_alpha_export,
  path = "alpha_diversity_male_data_and_statistics.xlsx"
)
```

Plotting function
```{r}
plot_alpha_box_signif <- function(alpha_long, sex_label) {
  
# Filter by sex
  df <- alpha_long %>%
    filter(Sex == sex_label)
  
 # Ensure factors
  df$Group  <- factor(df$Group,
                      levels = c("Control", "Stress"))
  
  df$Metric <- factor(df$Metric,
                      levels = c("Shannon", "Simpson"))
  
  df <- df %>%
    mutate(Group_num = as.numeric(Group))
  
# Stats
  sig_list <- lapply(levels(df$Metric), function(metric) {
    
    df_metric <- df %>%
      filter(Metric == metric)
    
    val1 <- df_metric$Value[df_metric$Group == "Control"]
    val2 <- df_metric$Value[df_metric$Group == "Stress"]
    
    if(length(val1) > 1 & length(val2) > 1) {
      
      test <- wilcox.test(
        val1,
        val2,
        exact = FALSE
      )
      
      p <- test$p.value
      
      sig <- ifelse(
        p < 0.001, "***",
        ifelse(
          p < 0.01, "**",
          ifelse(
            p < 0.05, "*", "ns"
          )
        )
      )
      
      p_label <- ifelse(
        p < 0.001,
        paste0("Wilcoxon, ", sig, "  p<0.001"),
        paste0("Wilcoxon, ", sig, "  p=", signif(p, 2))
      )
      
      return(
        data.frame(
          Metric = metric,
          x_start = 1,
          x_end = 2,
          label = p_label,
          sig = sig,
          p_value = p
        )
      )
    }
    
    return(NULL)
  })
  
  sig_df <- bind_rows(sig_list)
  
# Base plot
  
  p <- ggplot(
    df,
    aes(x = Group_num,
        y = Value,
        fill = Group)
  ) +
    
    geom_violin(
      trim = FALSE,
      alpha = 0.7
    ) +
    
    geom_boxplot(
      width = 0.1,
      outlier.shape = NA,
      alpha = 0.5,
      color = "black"
    ) +
    
    geom_jitter(
      width = 0.15,
      alpha = 0.7,
      size = 2,
      color = "black"
    ) +
    
    facet_wrap(
      ~ Metric,
      scales = "free_y"
    ) +
    
    scale_x_continuous(
      breaks = 1:2,
      labels = levels(df$Group)
    ) +
    
    scale_fill_manual(
      values = c(
        "Control" = "#4D4D4D",
        "Stress"  = "#FE8A13"
      )
    ) +
    
    theme_bw() +
    
    theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        hjust = 0.5
      ),
      
      axis.text.x = element_text(
        size = 12,
        face = "bold"
      ),
      
      axis.text.y = element_text(size = 10),
      
      strip.text = element_text(
        size = 14,
        face = "bold"
      ),
      
      axis.title.y = element_text(
        size = 14,
        face = "bold"
      ),
      
      legend.position = "none"
    ) +
    
    labs(
      x = NULL,
      y = "Alpha Diversity",
      title = paste(
        "Alpha Diversity - Large Intestine",
        ifelse(
          sex_label == "F",
          "Females",
          "Males"
        )
      )
    )
  
# Add significance bars

  if(!is.null(sig_df) && nrow(sig_df) > 0) {
    
    gb <- ggplot_build(p)
    
    y_ranges <- gb$layout$panel_params %>%
      lapply(function(x)
        c(x$y.range[1], x$y.range[2]))
    
    sig_df <- sig_df %>%
      group_by(Metric) %>%
      mutate(
        y = y_ranges[[which(
          levels(df$Metric) == Metric
        )]][2] * 1.05
      ) %>%
      ungroup()
    
    p <- p +
      
      geom_segment(
        data = sig_df,
        aes(
          x = x_start,
          xend = x_end,
          y = y,
          yend = y
        ),
        inherit.aes = FALSE,
        linewidth = 0.8
      ) +
      
      geom_text(
        data = sig_df,
        aes(
          x = (x_start + x_end)/2,
          y = y,
          label = label,
          color = sig
        ),
        inherit.aes = FALSE,
        vjust = -0.6,
        size = 4.5,
        show.legend = FALSE
      ) +
      
      scale_color_manual(
        values = c(
          "ns" = "gray50",
          "*" = "black",
          "**" = "black",
          "***" = "black"
        )
      )
  }
  
  return(p)
}
```

Generate plots
```{r}
alpha_sig_females <- plot_alpha_box_signif(
  alpha_long,
  "F"
)

alpha_sig_males <- plot_alpha_box_signif(
  alpha_long,
  "M"
)

# View plots
alpha_sig_females
alpha_sig_males
```

Save plots
```{r}
ggsave(
  "alpha_sig_females.png",
  plot = alpha_sig_females,
  width = 10,
  height = 6,
  dpi = 600
)

ggsave(
  "alpha_sig_males.png",
  plot = alpha_sig_males,
  width = 10,
  height = 6,
  dpi = 600
)
```

################# Differential abundance workflow #################
Load in relative abundance data for MaAsLin2 (males and females separately; Fig. S2E & F)
```{r}
input_data <- read_tsv("merged_df_final_abundance_m.tsv")
input_data <- input_data %>% column_to_rownames(var = names(input_data)[1])
input_data <- t(input_data)
input_metadata <- read_tsv("meta_data_m.tsv")
input_metadata <- input_metadata %>% column_to_rownames(var = names(input_metadata)[1])
```

MaAsLin2 analysis
```{r}
Maaslin2(
        input_data,
        input_metadata,
        output = "maaslin2_results_m_li",
        min_abundance = 0.0001,
        min_prevalence = 0.50,
        normalization = "NONE",
        transform = "LOG",
        analysis_method = "LM",
        max_significance = 0.05,
        random_effects = NULL,
        fixed_effects = "Group",
        correction = "BH",
        standardize = TRUE,
        plot_heatmap = TRUE,
        heatmap_first_n = 50,
        plot_scatter = TRUE,
        max_pngs = 10,
        save_scatter = FALSE,
        save_models = FALSE,
        reference = NULL
    )
```

Plotting results from MaAsLin2
Load in MaAsLin2 results (males and females separately)
```{r}
path <- "~/Desktop/all_results.tsv"
maaslin2_res <- read_tsv(path)
```

Plot horizontal bar chart using ggplot2
```{r}
# Keep only significant features
sig_results <- maaslin2_res %>%
  filter(qval < 0.10)

# Order taxa by effect size
sig_results <- sig_results %>%
  arrange(coef) %>%
  mutate(feature = factor(feature, levels = feature))

# Match beta diversity palette
fill_colors <- c(
  "Decreased" = "#4D4D4D",
  "Increased" = "#FE8A13"
)

# Add direction column
sig_results <- sig_results %>%
  mutate(Direction = ifelse(coef > 0,
                            "Increased",
                            "Decreased"))

# Plot
ggplot(sig_results,
       aes(x = coef,
           y = feature,
           fill = Direction)) +
  
  geom_col(
    width = 0.8,
    color = "gray50"
  ) +
  
  geom_vline(
    xintercept = 0,
    linetype = "solid",
    color = "gray50",
    linewidth = 0.7
  ) +
  
  scale_fill_manual(values = fill_colors) +
  
  labs(
    title = "Differentially Abundant Species - Large Intestine Males",
    x = "MaAsLin2 Coefficient",
    y = NULL,
    fill = NULL
  ) +
  
  theme_bw() +
  
  theme(
    plot.title = element_text(
      size = 16,
      face = "bold",
      hjust = 0.5
    ),
    
    axis.text.y = element_text(face = "italic"),
    
    axis.text = element_text(size = 12),
    
    axis.title = element_text(
      size = 14,
      face = "bold"
    ),
    
    legend.title = element_text(
      size = 12,
      face = "bold"
    ),
    
    legend.text = element_text(size = 11),
    
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    
    axis.line = element_line(color = "black"),
    
    legend.position = "top"
  )
```

Save plot
```{r}
ggsave(
  "maaslin2_barplot_males.png",
  width = 12,
  height = 6,
  dpi = 600
)
```