---
title: "TF_&_receptor_activity_inference"
author: "Amelia Slusser"
date: "2026-05-19"
output: html_document
---
################################################################################
Tools: decoupleR (DoRothEA regulons) + ChEA3 (REST API)
Input: Hippocampal global proteomics from female mice
Output: Active TFs, active receptors, integrated candidate list
################################################################################
 
==============================================================================
0. INSTALL AND LOAD PACKAGES
==============================================================================

```{r}
library(tidyverse)
library(decoupleR)   # DoRothEA regulons + activity inference
library(OmnipathR)   # receptor annotations
library(httr)        # ChEA3 API calls
library(jsonlite)    # JSON parsing for ChEA3
library(readxl)
library(readr)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(ggplot2)
```

==============================================================================
1. LOAD AND PREPARE YOUR HIPPOCAMPAL PROTEOMICS DATA
==============================================================================
 --- What your input file should look like ---
 A tab- or comma-separated file with at minimum:
  - A column of mouse gene symbols (e.g. "Gene.names" from MaxQuant)
   - A column of quantitative values per sample
     (LFQ intensity, iBAQ, TMT ratio, or fold-change vs. control)

 Example structure (MaxQuant proteinGroups.txt):
   Gene.names | LFQ.intensity.sample1 | LFQ.intensity.sample2 | ...
   Camk2a     | 4.2e8                 | 3.9e8                 | ...

 Adjust the path and column names below to match your actual file.
```{r}
df_raw <- as.data.frame(read_xlsx("f_pfc_sig.xlsx"))
```

 --- Create row names ---
```{r}
rownames(df_raw) <- df_raw[,1]
df_raw <- df_raw[,-1]
```
 
 --- Handle duplicate gene symbols (optional if needed) ---
 Some proteins share gene symbols. Aggregate by taking the mean across duplicates.
```{r}
#mat_clean <- mat_raw %>%
  #rownames_to_column("Gene") %>%
  #group_by(Gene) %>%
  #summarise(across(everything(), ~ mean(.x, na.rm = TRUE)), .groups = "drop") %>%
  #column_to_rownames("Gene")
```

# --- Convert to matrix (required by decoupleR) ---
 Rows = genes, columns = samples
```{r}
pfc_mat <- as.matrix(df_raw)
```
 
 Quick sanity check: should print gene symbols and sample names
```{r}
cat("Matrix dimensions:", dim(pfc_mat), "\n")
cat("Sample names:", colnames(pfc_mat), "\n")
cat("First few gene symbols:", head(rownames(pfc_mat)), "\n")
```

 ==============================================================================
 2. DOROTHEA: INFER DIFFERENTIAL TF ACTIVITY
 ==============================================================================
   Per-sample: run decoupleR on all 12 samples, then test whether scores
       differ between groups. Gives richer statistics and per-sample variance

 --- Load DoRothEA mouse regulons ---
 get_collectri() is the recommended up-to-date regulon source in decoupleR.
 It supersedes the older DoRothEA v2 but uses the same confidence-tier logic.
 organism = 9606 is human; use 10090 for mouse.

```{r}
regulons <- get_collectri(organism = "10090", split_complexes = FALSE)
```

 Alternatively, use the classic DoRothEA network filtered to confidence A+B:
```{r}
#regulons <- get_dorothea(organism = "mouse", levels = c("A", "B"))
```
 
 Preview regulon structure:
 source = TF gene symbol, target = regulated gene, mor = mode of regulation (+1 activated, -1 repressed)
```{r}
head(regulons)
```

Export regulon table (only need this once)
```{r}
#write_tsv(regulons, "DoRothEA_regulons_all.tsv")
```

 --- Per-sample scores then group comparison ---
Define which columns belong to which group 
```{r}
control_cols <- c("CORT_1L",	"CORT_1N",	"CORT_1R",	"CORT_3L",	"CORT_3N",	"CORT_3R")
 
stress_cols  <- c("CORT_5",	"CORT_6",	"CORT_7",	"CORT_11",	"CORT_12",	"CORT_13")
 
all_sample_cols <- c(control_cols, stress_cols)
```

 Using WMean, run all samples together to preserve comparability with null distribution
```{r}
tf_per_sample <- run_wmean(
  mat     = pfc_mat,
  net     = regulons,
  .source = "source",
  .target = "target",
  .mor    = "mor",
  times   = 1000,
  minsize = 5
) %>%
  filter(statistic == "norm_wmean")
```

 Alternatively, run all samples together using MLM
```{r}
tf_per_sample <- run_mlm(
  mat     = pfc_mat,
  net     = regulons,
  .source = "source",
  .target = "target",
  .mor    = "mor",
  sparse  = FALSE,
  center  = FALSE,
  na.rm   = FALSE,
  minsize = 5
)
```

 Add group labels to per-sample scores
```{r}
group_labels <- tibble(
  condition = c(control_cols, stress_cols),
  group     = c(rep("control", length(control_cols)),
                rep("stress",  length(stress_cols)))
)
 
tf_per_sample_grouped <- tf_per_sample %>%
  left_join(group_labels, by = "condition")
```
 
 Test TF activity difference between groups (Welch t-test per TF)
```{r}
tf_group_test <- tf_per_sample_grouped %>%
  group_by(source) %>%
  summarise(
    mean_ctrl       = mean(score[group == "control"], na.rm = TRUE),
    mean_stress     = mean(score[group == "stress"],  na.rm = TRUE),
    sd_ctrl         = sd(score[group == "control"],   na.rm = TRUE),
    sd_stress       = sd(score[group == "stress"],    na.rm = TRUE),
    p_value         = tryCatch(
      t.test(score[group == "stress"],
             score[group == "control"])$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p_adj        = p.adjust(p_value, method = "BH"),
    delta_score  = mean_stress - mean_ctrl,
    direction    = if_else(delta_score > 0,
                           "more_active_in_stress", "less_active_in_stress")
  ) %>%
  filter(!is.na(p_value)) %>%
  arrange(p_adj)
 
sig_tf_group <- tf_group_test %>%
  filter(p_adj < 0.05, abs(delta_score) > 0.5) 

sig_tf_group <- sig_tf_group %>%
  rename(TF = source)
 
cat("\nDifferentially active TFs (per-sample method):", nrow(sig_tf_group), "\n")

write_tsv(tf_per_sample_grouped, "TF_per_sample_grouped.tsv")
write_tsv(tf_group_test, "TF_activity_per_group.tsv")
write_tsv(sig_tf_group, "Sig_only_TF_activity_per_group.tsv")
```

 ==============================================================================
 3. VISUALIZE DIFFERENTIAL TF ACTIVITY
 ==============================================================================
 
 --- Volcano-style plot of TF activity scores ---
```{r}
ggplot(tf_group_test %>% filter(!is.na(p_adj)),
       aes(x = delta_score, y = -log10(p_adj),
           color = p_adj < 0.05 & abs(delta_score) > 0.5)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", alpha = 0.5) +
  ggrepel::geom_text_repel(
    data = . %>% filter(p_adj < 0.05, abs(delta_score) > 0.5),
    aes(label = source), size = 3, max.overlaps = 20
  ) +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = "#1D9E75"),
                     guide = "none") +
  labs(title  = "CORT Differential TF activity: stress vs. control",
       x      = "Delta activity score (stress - control)",
       y      = "-log10(adjusted p-value)") +
  theme_minimal(base_size = 12)
 
ggsave("TF_activity_volcano.pdf", width = 8, height = 6)
```
 
  --- Heatmap of top TF activity scores per sample ---
```{r}
# ── Setup ──────────────────────────────────────────────────────────────

sample_order <- c(control_cols, stress_cols)

group_vector <- c(rep("Control", length(control_cols)),
                  rep("Stress",  length(stress_cols)))
names(group_vector) <- sample_order

ha_col <- HeatmapAnnotation(
  Group = factor(
    group_vector,
    levels = c("Control", "Stress")
  ),

  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress  = "#FE8A13"
    )
  ),

  border = TRUE
)

# ══════════════════════════════════════════════════════════════════════════════
# HEATMAP 1 - All TFs (per-sample scores, group comparison sig and non sig)
# ══════════════════════════════════════════════════════════════════════════════

# Pull all TFs — top 50 by absolute delta score
group_tf_names <- tf_group_test %>%
  slice_max(abs(delta_score), n = 50) %>%
  pull(source)

# Build the per-sample score matrix for these TFs
# tf_per_sample comes from run_wmean() on the full 12-sample matrix 
heatmap_data <- tf_per_sample %>%
  filter(source %in% group_tf_names, statistic == "norm_wmean") %>%
  select(source, sample = condition, score) %>%
  pivot_wider(names_from = sample, values_from = score) %>%
  column_to_rownames("source") %>%
  as.matrix()

# Reorder columns to match desired sample order
# (only keep columns that exist — guards against any missing samples)
heatmap_data <- heatmap_data[, intersect(sample_order,
                                              colnames(heatmap_data))]

# Row-scale so TFs with very different absolute activity ranges are comparable
heatmap_data_scaled <- t(scale(t(heatmap_data)))

# Create significance labels
sig_df <- tf_group_test %>%
  filter(source %in% rownames(heatmap_data_scaled)) %>%
  distinct(source, p_adj, delta_score) %>%
  mutate(

    sig_label = case_when(
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    )
  )
# Match annotation order to heatmap rows
sig_df <- sig_df[
  match(rownames(heatmap_data_scaled), sig_df$source),
]
# Create p-val color function
p_col_fun <- colorRamp2(
  c(1, 0.05, 0.01, 0),
  c("#F0F0F0", "#FDBBA1", "#FB6A4A", "#EF3B2C")
)

# Right-side annotation showing group-level statistics from sig_tf_group
delta_scores <- tf_group_test %>%
  filter(source %in% group_tf_names) %>%
  arrange(match(source, rownames(heatmap_data_scaled))) %>%
  pull(delta_score)

ha_row <- rowAnnotation(

  Adj_FDR = anno_simple(

    sig_df$p_adj,

    col = p_col_fun,

    pch = sig_df$sig_label,

    pt_gp = grid::gpar(
      fontsize = 9,
      fontface = "bold",
      col = "black"
    ),

    width = unit(3, "mm"),

    border = TRUE
  ),

  "Delta activity score\n(stress−ctrl)" = anno_barplot(

    sig_df$delta_score,

    gp = grid::gpar(
      fill = ifelse(sig_df$delta_score > 0,
                    "#FE8A13",
                    "#4D4D4D"),
      col = NA
    ),

    axis_param = list(
      gp = grid::gpar(fontsize = 7)
    ),

    width = unit(2.5, "cm")
  ),
  show_annotation_name = c(Adj_FDR = FALSE)
  )

ht1 <- Heatmap(
  heatmap_data_scaled,
  name               = "TF Activity\n(row-scaled)",
  top_annotation     = ha_col,
  right_annotation   = ha_row,
  col                = viridis(100),
  width = unit(8, "cm"),
  cluster_columns    = FALSE,
  cluster_rows       = TRUE,
  show_row_names     = TRUE,
  show_column_names  = TRUE,
  row_names_gp       = grid::gpar(fontsize = 9),
  column_names_gp    = grid::gpar(fontsize = 9),
  column_title       = "Cortex TF Activity"
)

pdf("TF_activity_heatmap_all.pdf", width = 8, height = 7)
draw(ht1, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

# ══════════════════════════════════════════════════════════════════════════════
# HEATMAP 2 - TFs (per-sample scores, group comparison significant only)
# ══════════════════════════════════════════════════════════════════════════════

# Pull the significant TFs — top 50 by absolute delta score
group_tf_names <- sig_tf_group %>%
  slice_max(abs(delta_score), n = 50) %>%
  pull(TF)

# Build the per-sample score matrix for these TFs
# tf_per_sample comes from run_wmean() on the full 12-sample matrix 
heatmap_data <- tf_per_sample %>%
  filter(source %in% group_tf_names,
         statistic == "norm_wmean") %>%
  select(TF = source,
         sample = condition,
         score) %>%
  pivot_wider(names_from = sample,
              values_from = score) %>%
  column_to_rownames("TF") %>%
  as.matrix()

# Reorder columns to match desired sample order
# (only keep columns that exist — guards against any missing samples)
heatmap_data <- heatmap_data[, intersect(sample_order,
                                              colnames(heatmap_data))]

# Row-scale so TFs with very different absolute activity ranges are comparable
heatmap_data_scaled <- t(scale(t(heatmap_data)))

# Right-side annotation showing group-level statistics from sig_tf_group
delta_scores <- sig_tf_group %>%
  filter(TF %in% group_tf_names) %>%
  arrange(match(TF, rownames(heatmap_data_scaled))) %>%
  pull(delta_score)

ha_row <- rowAnnotation(
  "Delta activity score\n(stress−ctrl)" = anno_barplot(
    delta_scores,
    gp        = grid::gpar(fill = ifelse(delta_scores > 0, "#FE8A13", "#4D4D4D"),
                           col  = NA),
    axis_param = list(gp = grid::gpar(fontsize = 7)),
    width      = unit(2.5, "cm")
  )
)

ht2 <- Heatmap(
  heatmap_data_scaled,
  name               = "TF Activity\n(row-scaled)",
  top_annotation     = ha_col,
  right_annotation   = ha_row,
  col                = viridis(100),
  width = unit(8, "cm"),
  cluster_columns    = FALSE,
  cluster_rows       = TRUE,
  show_row_names     = TRUE,
  show_column_names  = TRUE,
  row_names_gp       = grid::gpar(fontsize = 9),
  column_names_gp    = grid::gpar(fontsize = 9),
  column_title       = "Cortex TF activity, significant by group comparison"
)

pdf("TF_activity_heatmap_sig.pdf", width = 10, height = 7)
draw(ht2, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()
```

  --- Heatmap of top TF activity scores per sample MLM ---
```{r}
# ── Setup ──────────────────────────────────────────────────────────────

sample_order <- c(control_cols, stress_cols)

group_vector <- c(rep("Control", length(control_cols)),
                  rep("Stress",  length(stress_cols)))
names(group_vector) <- sample_order

# Only keep samples that are actually present in tf_per_sample
sample_order_present <- intersect(sample_order, unique(tf_per_sample$condition))
group_vector_present <- group_vector[sample_order_present]

ha_col <- HeatmapAnnotation(
  Group = factor(
    group_vector_present,
    levels = c("Control", "Stress")
  ),
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress  = "#FE8A13"
    )
  ),
  border = TRUE
)

# Helper function to safely build heatmap matrix from MLM output
make_tf_heatmap_matrix <- function(tf_names, tf_per_sample, sample_order_present) {
  
  heatmap_data <- tf_per_sample %>%
    filter(source %in% tf_names) %>%
    select(source, sample = condition, score) %>%
    pivot_wider(
      names_from = sample,
      values_from = score
    ) %>%
    column_to_rownames("source") %>%
    as.matrix()
  
  # Keep only expected samples, preserving order
  heatmap_data <- heatmap_data[
    ,
    intersect(sample_order_present, colnames(heatmap_data)),
    drop = FALSE
  ]
  
  # Remove rows with all NA values
  heatmap_data <- heatmap_data[rowSums(!is.na(heatmap_data)) > 0, , drop = FALSE]
  
  # Row-scale
  heatmap_data_scaled <- t(scale(t(heatmap_data)))
  
  # Replace NaN values caused by zero variance rows
  heatmap_data_scaled[is.nan(heatmap_data_scaled)] <- 0
  
  return(heatmap_data_scaled)
}

# ══════════════════════════════════════════════════════════════════════════════
# HEATMAP 1 - Top 50 TFs by absolute delta score
# ══════════════════════════════════════════════════════════════════════════════

group_tf_names <- tf_group_test %>%
  slice_max(abs(delta_score), n = 50) %>%
  pull(source)

heatmap_data_scaled <- make_tf_heatmap_matrix(
  tf_names = group_tf_names,
  tf_per_sample = tf_per_sample,
  sample_order_present = sample_order_present
)

# Stop early with a useful message if no TFs are available
if (nrow(heatmap_data_scaled) == 0) {
  stop("No TFs available for Heatmap 1. Check whether tf_per_sample$source matches tf_group_test$source.")
}

sig_df <- tf_group_test %>%
  filter(source %in% rownames(heatmap_data_scaled)) %>%
  distinct(source, p_adj, delta_score) %>%
  mutate(
    sig_label = case_when(
      p_adj < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

sig_df <- sig_df[
  match(rownames(heatmap_data_scaled), sig_df$source),
]

p_col_fun <- colorRamp2(
  c(1, 0.05, 0.01, 0),
  c("#F0F0F0", "#FDBBA1", "#FB6A4A", "#EF3B2C")
)

ha_row <- rowAnnotation(
  
  Adj_FDR = anno_simple(
    sig_df$p_adj,
    col = p_col_fun,
    pch = sig_df$sig_label,
    pt_gp = grid::gpar(
      fontsize = 9,
      fontface = "bold",
      col = "black"
    ),
    width = unit(3, "mm"),
    border = TRUE
  ),
  
  "Delta activity score\n(stress−ctrl)" = anno_barplot(
    sig_df$delta_score,
    gp = grid::gpar(
      fill = ifelse(sig_df$delta_score > 0,
                    "#FE8A13",
                    "#4D4D4D"),
      col = NA
    ),
    axis_param = list(
      gp = grid::gpar(fontsize = 7)
    ),
    width = unit(2.5, "cm")
  ),
  
  show_annotation_name = c(Adj_FDR = FALSE)
)

ht1 <- Heatmap(
  heatmap_data_scaled,
  name               = "TF Activity\n(row-scaled)",
  top_annotation     = ha_col,
  right_annotation   = ha_row,
  col                = viridis(100),
  width              = unit(8, "cm"),
  cluster_columns    = FALSE,
  cluster_rows       = TRUE,
  show_row_names     = TRUE,
  show_column_names  = TRUE,
  row_names_gp       = grid::gpar(fontsize = 9),
  column_names_gp    = grid::gpar(fontsize = 9),
  column_title       = "CORT TF Activity"
)

pdf("CORT_TF_activity_heatmap_all_MLM.pdf", width = 8, height = 8)
draw(ht1, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

# ══════════════════════════════════════════════════════════════════════════════
# HEATMAP 2 - Significant TFs only
# ══════════════════════════════════════════════════════════════════════════════

group_tf_names_sig <- sig_tf_group %>%
  slice_max(abs(delta_score), n = 50) %>%
  pull(TF)

# Only make this heatmap if significant TFs exist
if (length(group_tf_names_sig) == 0) {
  
  message("No significant TFs found using p_adj < 0.05 and abs(delta_score) > 0.5. Skipping significant-only heatmap.")
  
} else {
  
  heatmap_data_scaled_sig <- make_tf_heatmap_matrix(
    tf_names = group_tf_names_sig,
    tf_per_sample = tf_per_sample,
    sample_order_present = sample_order_present
  )
  
  if (nrow(heatmap_data_scaled_sig) == 0) {
    stop("No TFs available for Heatmap 2 after filtering significant TFs.")
  }
  
  delta_scores_sig <- sig_tf_group %>%
    filter(TF %in% rownames(heatmap_data_scaled_sig)) %>%
    arrange(match(TF, rownames(heatmap_data_scaled_sig))) %>%
    pull(delta_score)
  
  ha_row_sig <- rowAnnotation(
    "Delta activity score\n(stress−ctrl)" = anno_barplot(
      delta_scores_sig,
      gp = grid::gpar(
        fill = ifelse(delta_scores_sig > 0,
                      "#FE8A13",
                      "#4D4D4D"),
        col = NA
      ),
      axis_param = list(
        gp = grid::gpar(fontsize = 7)
      ),
      width = unit(2.5, "cm")
    )
  )
  
  ht2 <- Heatmap(
    heatmap_data_scaled_sig,
    name               = "TF Activity\n(row-scaled)",
    top_annotation     = ha_col,
    right_annotation   = ha_row_sig,
    col                = viridis(100),
    width              = unit(8, "cm"),
    cluster_columns    = FALSE,
    cluster_rows       = TRUE,
    show_row_names     = TRUE,
    show_column_names  = TRUE,
    row_names_gp       = grid::gpar(fontsize = 9),
    column_names_gp    = grid::gpar(fontsize = 9),
    column_title       = "CORT TF Activity, significant by group comparison"
  )
  
  pdf("CORT_TF_activity_heatmap_sig_MLM.pdf", width = 10, height = 8)
  draw(ht2, heatmap_legend_side = "right", annotation_legend_side = "right")
  dev.off()
}
```

Export the heat map data
```{r}
# Convert scaled heatmap matrix to a data frame
# and preserve TF names in the first column
heatmap_df <- heatmap_data_scaled %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "TF")

# Export to Excel
writexl::write_xlsx(
  heatmap_df,
  path = "CORT_TF_activity_heatmap_data_scaled.xlsx"
)
```

 ==============================================================================
 4. EXPLAIN WHICH PROTEINS DROVE EACH TF INFERENCE
 ==============================================================================
 For every significant TF, show which detected proteins contributed to its
 inferred activity score, and how that contribution differs between groups.
```{r}
explain_tf_drivers <- function(tf_name, regulons, pfc_mat,
                               ctrl_cols, stress_cols) {
  # Targets of this TF in the regulon
  tf_targets <- regulons %>%
    filter(source == tf_name) %>%
    select(target, mor)
 
  # Targets that appear in the data
  detected <- tf_targets %>%
    filter(target %in% rownames(pfc_mat)) %>%
    mutate(
      mean_ctrl   = rowMeans(pfc_mat[target, ctrl_cols,   drop = FALSE],
                             na.rm = TRUE),
      mean_stress = rowMeans(pfc_mat[target, stress_cols, drop = FALSE],
                             na.rm = TRUE),
      log2FC      = mean_stress - mean_ctrl,
      # Weighted contribution: direction of regulation × fold-change
      # Positive = supports TF being more active in stress
      # Negative = supports TF being less active in stress
      weighted_contribution = mor * log2FC
    ) %>%
    arrange(desc(abs(weighted_contribution)))
 
  detected
}
```

```{r}
# Run for all TFs and save a combined driver table
all_drivers <- map_dfr(sig_tf_group$TF, function(tf) {
  explain_tf_drivers(tf, regulons, pfc_mat, control_cols, stress_cols) %>%
    mutate(
      inferred_TF   = tf,
      TF_direction  = sig_tf_group$direction[sig_tf_group$TF == tf],
      delta_score   = sig_tf_group$delta_score[sig_tf_group$TF == tf]
    )
}, .progress = TRUE)

# Flag the most influential drivers: top contributors per TF
key_drivers <- all_drivers %>%
  group_by(inferred_TF) %>%
  slice_max(abs(weighted_contribution), n = 10) %>%
  ungroup() %>%
  arrange(inferred_TF, desc(abs(weighted_contribution)))

write_tsv(all_drivers, "TF_driver_proteins_all_group.tsv")
write_tsv(key_drivers, "TF_driver_proteins_top10_group.tsv")

cat("\nDriver tables saved.\n")
cat("Columns in driver table:\n")
cat("  target                = detected protein that contributed\n")
cat("  mor                   = mode of regulation (+1 activated, -1 repressed)\n")
cat("  mean_ctrl/mean_stress = log2 abundance in each group\n")
cat("  log2FC                = stress - control log2 abundance difference\n")
cat("  weighted_contribution = mor x log2FC (positive = supports TF being\n")
cat("                          more active in stress)\n")
cat("  inferred_TF           = which TF this protein helped infer\n")
cat("  TF_direction          = more/less_active_in_stress\n")
cat("  delta_score           = delta score\n")
```

 ==============================================================================
 6. CHEA3: GROUP-AWARE ANALYSIS
 ==============================================================================
 ChEA3 is list-based (no scores), so run it separately for each group.
 Load in gene/protein significance data
```{r}
df_t <- as.data.frame(read_excel("f_pfc_t.xlsx"))
```
 
 Compare which TFs are enriched in stress-enriched vs control-enriched genes.
```{r}
# Stress-enriched proteins (upregulated in stress)
stress_up_genes <- df_t %>%
  filter(t > 0) %>%
  pull(Gene)
stress_up_genes <- toupper(stress_up_genes)

# Control-enriched proteins (downregulated in stress)
ctrl_up_genes <- df_t %>%
  filter(t < 0) %>%
  pull(Gene)
ctrl_up_genes <- toupper(ctrl_up_genes)

cat("\nGenes upregulated in stress (for ChEA3):", length(stress_up_genes), "\n")
cat("Genes upregulated in control (for ChEA3):", length(ctrl_up_genes), "\n")
``` 

 ChEA3 query function
```{r}
chea3_query <- function(gene_list, query_name) {

  payload <- list(
    query_name = query_name,
    gene_set = gene_list
  )

  response <- httr::POST(
    url = "https://maayanlab.cloud/chea3/api/enrich/",
    body = payload,
    encode = "json"
  )

  httr::stop_for_status(response)

  jsonlite::fromJSON(
    httr::content(response, as = "text", encoding = "UTF-8"),
    flatten = TRUE
  )
}
```

 Run queries
```{r}
chea3_stress <- chea3_query(
  stress_up_genes,
  "stress_upregulated"
)

chea3_control <- chea3_query(
  ctrl_up_genes,
  "control_upregulated"
)
```

 Parse results
```{r}
parse_chea3 <- function(result, label) {

  if (is.null(result)) {
    return(tibble())
  }

  # Check available result tables
  print(names(result))

  # Integrated ranking table
  tbl <- result[["Integrated--meanRank"]]

  if (is.null(tbl)) {
    message("Integrated--meanRank table not found")
    return(tibble())
  }

  as_tibble(tbl) %>%
    mutate(
      mean_rank = as.numeric(Score),
      group = label
    ) %>%
    select(TF, mean_rank, group) %>%
    arrange(mean_rank)
}
```

 Final tables
```{r}
chea3_stress_tbl <- parse_chea3(
  chea3_stress,
  "stress"
)

chea3_control_tbl <- parse_chea3(
  chea3_control,
  "control"
)

# Preview
head(chea3_stress_tbl)
head(chea3_control_tbl)
```

 TFs preferentially enriched in stress vs. control gene sets
```{r}
if (nrow(chea3_stress_tbl) > 0 & nrow(chea3_control_tbl) > 0) {
  chea3_comparison <- full_join(
    chea3_stress_tbl  %>% rename(rank_stress  = mean_rank) %>% select(-group),
    chea3_control_tbl %>% rename(rank_control = mean_rank) %>% select(-group),
    by = "TF"
  ) %>%
    mutate(
      rank_stress  = replace_na(rank_stress,  max(rank_stress,  na.rm = TRUE)),
      rank_control = replace_na(rank_control, max(rank_control, na.rm = TRUE)),
      rank_diff    = rank_control - rank_stress,  # positive = preferentially in stress
      preference   = case_when(
        rank_diff >  50 ~ "stress_preferential",
        rank_diff < -50 ~ "control_preferential",
        TRUE            ~ "shared"
      )
    ) %>%
    arrange(rank_stress)
 
  write_tsv(chea3_comparison, "ChEA3_stress_vs_control_comparison.tsv")
  cat("\nChEA3 comparison saved.\n")
}
```

 ==============================================================================
 7. RECEPTOR ACTIVITY — DIRECT PROTEIN DETECTION APPROACH
 ==============================================================================
 DoRothEA infers TF activity; it does NOT infer receptor activity directly.
 For receptors, the most defensible approach with proteomics data is:
 (a) identify which receptors were actually detected in your data, and
 (b) use their abundance as a proxy for activity.

 We pull a curated receptor list from OmniPath (intercell database).
 
 --- Get mouse receptor annotations from OmniPath ---
```{r}
intercell <- intercell_network(
  interactions_param = list(datasets = "omnipath"),
  transmitter_param  = list(categories = "ligand"),
  receiver_param     = list(categories = "receptor")
)
```

 Extract unique receptor gene symbols (convert to mouse case: first-letter cap)
```{r}
human_receptors <- intercell %>%
  pull(target_genesymbol) %>%
  unique()
```

 OmniPath is human-centric; convert to mouse gene symbols by lowercasing all
 but the first character (standard mouse gene symbol convention)
```{r}
mouse_receptors <- str_to_title(tolower(human_receptors))
```

 --- Identify receptors detected in your hippocampal data ---
```{r}
detected_receptors <- intersect(rownames(pfc_mat), mouse_receptors)
cat("\nReceptors detected in cortex proteomics:", length(detected_receptors), "\n")
print(detected_receptors)
```

 --- Extract receptor abundances ---
```{r}
receptor_abundance <- pfc_mat[detected_receptors, , drop = FALSE] %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sample", values_to = "log2_abundance") %>%
  group_by(gene) %>%
  summarise(mean_log2_abundance = mean(log2_abundance, na.rm = TRUE),
            n_samples_detected  = sum(!is.na(log2_abundance)),
            .groups = "drop") %>%
  arrange(desc(mean_log2_abundance))
 
cat("\nTop detected receptors by mean abundance:\n")
print(head(receptor_abundance, 20))
 
write_tsv(receptor_abundance, "detected_receptors_f_pfc.tsv")
```

 ==============================================================================
 8. FINAL INTEGRATED OUTPUT
 ==============================================================================
 
 Build the final candidate table, now with group context
```{r}
tfs_final <- sig_tf_group %>%
  left_join(
    chea3_comparison %>%
      select(TF),
    by = "TF"
  ) %>%
  mutate(
    regulator_type = "TF",
    evidence       = "DoRothEA + ChEA3 (differential)"
  )
```
 
 Receptors: report mean abundance per group
```{r}
detected_receptor_names <- intersect(rownames(pfc_mat), mouse_receptors)
 
receptors_final <- tibble(gene = detected_receptor_names) %>%
  mutate(
    mean_log2_ctrl   = rowMeans(pfc_mat[gene, control_cols, drop = FALSE],
                                na.rm = TRUE),
    mean_log2_stress = rowMeans(pfc_mat[gene, stress_cols,  drop = FALSE],
                                na.rm = TRUE),
    log2FC_receptor  = mean_log2_stress - mean_log2_ctrl,
    regulator_type   = "receptor",
    evidence         = "direct detection"
  ) %>%
  filter(gene %in% rownames(pfc_mat)) 
```

 Putting it all together
```{r}
final <- bind_rows(
  tfs_final %>% rename(gene = TF),
  receptors_final
)
 
write_tsv(final, "TF_receptor_candidates_differential.tsv")
 
cat("\n========================================\n")
cat("Differential TF and receptor analysis complete.\n\n")
cat("Key output files:\n")
cat("  TF_activity_per_sample_group_method.tsv     — per-sample group comparison\n")
cat("  TF_driver_proteins_top10.tsv                — proteins driving each TF inference\n")
cat("  TF_activity_volcano.pdf                     — volcano plot\n")
cat("  TF_activity_heatmap.pdf                     — per-sample heatmap\n")
cat("  ChEA3_stress_vs_control_comparison.tsv      — ChEA3 group comparison\n")
cat("  TF_receptor_candidates_differential.tsv     — FINAL table\n")
cat("========================================\n") 
```

 ══════════════════════════════════════════════════════════════════════════════
 COMBINED TF + RECEPTOR HEATMAP
 ══════════════════════════════════════════════════════════════════════════════
 Pull TFs & receptors from final results table
```{r}
tf_names <- final %>%
  filter(regulator_type == "TF") %>%
  pull(gene)

receptor_names <- final %>%
  filter(regulator_type == "receptor") %>%
  pull(gene)
```

Build TF & receptor activity/abundance matrices and combine
```{r}
tf_matrix <- tf_per_sample %>%
  filter(source %in% tf_names,
         statistic == "norm_wmean") %>%
  select(source,
         sample = condition,
         score) %>%
  pivot_wider(
    names_from  = sample,
    values_from = score
  ) %>%
  column_to_rownames("source") %>%
  as.matrix()

# reorder samples
tf_matrix <- tf_matrix[, intersect(sample_order,
                                   colnames(tf_matrix))]

# row-scale TF activity
tf_matrix_scaled <- t(scale(t(tf_matrix)))

# receptor_mat should contain receptor abundances
# rows = genes/proteins
# cols = samples

receptor_matrix <- pfc_mat[
  intersect(receptor_names, rownames(pfc_mat)),
  sample_order,
  drop = FALSE
]

# row-scale receptors
receptor_matrix_scaled <- t(scale(t(receptor_matrix)))

# Combine
combined_matrix <- rbind(
  tf_matrix_scaled,
  receptor_matrix_scaled
)

# Force Control on left, Stress on right
combined_matrix <- combined_matrix[, sample_order]
```

Row annotation dataframe
```{r}
row_type <- c(
  rep("TF", nrow(tf_matrix_scaled)),
  rep("Receptor", nrow(receptor_matrix_scaled))
)

names(row_type) <- rownames(combined_matrix)
```

Left-side annotation block
```{r}
ha_row_left <- rowAnnotation(
  Type = row_type,

  col = list(
    Type = c(
      TF = "purple",
      Receptor = "gold2"
    )
  ),

  border = TRUE,

  show_annotation_name = TRUE,

  annotation_name_gp = grid::gpar(fontsize = 9)
)
```

Optional right-side delta score annotation (TFs only)
```{r}
delta_scores <- final %>%
  filter(regulator_type == "TF",
         gene %in% rownames(tf_matrix_scaled)) %>%
  arrange(match(gene, rownames(tf_matrix_scaled))) %>%
  pull(delta_score)

# pad receptors with NA
delta_scores_full <- c(
  delta_scores,
  rep(NA, nrow(receptor_matrix_scaled))
)

ha_row_right <- rowAnnotation(
  "Delta activity score\n(stress−ctrl)" = anno_barplot(
    delta_scores_full,

    gp = grid::gpar(
      fill = ifelse(delta_scores_full > 0,
                    "#FE8A13",
                    "#4D4D4D"),
      col = NA
    ),

    border = FALSE,

    axis_param = list(
      gp = grid::gpar(fontsize = 7)
    ),

    width = unit(2.5, "cm")
  )
)
```

Optional right-sided log2FC annotation (Receptors only)
```{r}
log2_fc <- final %>%
  filter(regulator_type == "Receptor",
         gene %in% rownames(tf_matrix_scaled)) %>%
  arrange(match(gene, rownames(tf_matrix_scaled))) %>%
  pull(log2FC_receptor)

# pad TFs with NA
log2_fc_full <- c(
  log2_fc,
  rep(NA, nrow(tf_matrix_scaled))
)

ha_row_right <- rowAnnotation(
  "Delta activity score\n(stress−ctrl)" = anno_barplot(
    delta_scores_full,

    gp = grid::gpar(
      fill = ifelse(delta_scores_full > 0,
                    "#FE8A13",
                    "#4D4D4D"),
      col = NA
    ),

    border = TRUE,

    axis_param = list(
      gp = grid::gpar(fontsize = 7)
    ),

    width = unit(2.5, "cm")
  )
)
```

Create combined heatmap
```{r}
ht_combined <- Heatmap(
  combined_matrix,

  name = "Row-scaled\nsignal",

  top_annotation = ha_col,

  left_annotation = ha_row_left,

  right_annotation = ha_row_right,

  col = viridis(100),

  cluster_columns = FALSE,

  cluster_rows = TRUE,
  
  row_split = row_type,

  show_row_names = TRUE,

  show_column_names = TRUE,
  
  column_names_rot = 90,

  row_names_gp = grid::gpar(fontsize = 8),

  column_names_gp = grid::gpar(fontsize = 9),

  column_title =
    "Integrated cortex TF activity & receptor abundance"
)
```

Save
```{r}
pdf(
  "Integrated_TF_receptor_heatmap.pdf",
  width = 9,
  height = 12
)

draw(
  ht_combined,
  heatmap_legend_side = "right",
  annotation_legend_side = "right"
)

dev.off()
```

 ══════════════════════════════════════════════════════════════════════════════
 Specific TF violin plots
 ══════════════════════════════════════════════════════════════════════════════
 Plotting function
```{r}
plot_tf_violin <- function(tf_scores,
                           tf_list = c("Esr1"),
                           brain_region = "CORT",
                           sex_label = "F") {
  # Filter selected TFs
  df <- tf_scores %>%
    filter(source %in% tf_list) %>%
    mutate(
      group = factor(group,
                     levels = c("control", "stress"),
                     labels = c("Control", "Stress")),
      TF = factor(source,
                  levels = tf_list),
      Group_num = as.numeric(group)
    )
  # ----- Statistics -----
  sig_list <- lapply(levels(df$TF), function(tf) {
    df_tf <- df %>%
      filter(TF == tf)
    val1 <- df_tf$score[df_tf$group == "Control"]
    val2 <- df_tf$score[df_tf$group == "Stress"]
    if(length(val1) > 1 & length(val2) > 1) {
      test <- wilcox.test(
        val1,
        val2,
        exact = FALSE
      )
      p <- test$p.value
      sig <- ifelse(
        p < 0.001, "***",
        ifelse(
          p < 0.01, "**",
          ifelse(
            p < 0.05, "*", "ns"
          )
        )
      )
      p_label <- ifelse(
        p < 0.001,
        paste0(sig),
        paste0(sig)
      )
      return(
        data.frame(
          TF = tf,
          x_start = 1,
          x_end = 2,
          label = p_label,
          sig = sig,
          p_value = p
        )
      )
    }
    return(NULL)
  })

  sig_df <- bind_rows(sig_list) %>%
  mutate(
    group1 = x_start,
    group2 = x_end
  )
  
  # Calculate an annotation position separately for each TF
tf_y_positions <- df %>%
  group_by(TF) %>%
  summarise(
    y_min = min(score, na.rm = TRUE),
    y_max = max(score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y_range = y_max - y_min,
    y_range = if_else(y_range > 0, y_range, 1),
    
    # Position bracket slightly above the highest observation
    y.position = y_max + 0.08 * y_range
  )

sig_df <- sig_df %>%
  left_join(tf_y_positions, by = "TF")

   # ----- Base Plot -----
  p <- ggplot(
    df,
    aes(
      x = Group_num,
      y = score,
      fill = group
    )
  ) +
    geom_violin(
      trim = FALSE,
      alpha = 0.7
    ) +
    geom_boxplot(
      width = 0.1,
      outlier.shape = NA,
      alpha = 0.5,
      color = "black"
    ) +
    geom_jitter(
      width = 0.15,
      alpha = 0.7,
      size = 2,
      color = "black"
    ) +
    facet_wrap(
      ~ TF,
      scales = "free_y"
    ) +
    scale_x_continuous(
      breaks = 1:2,
      labels = levels(df$group)
    ) +
    scale_fill_manual(
      values = c(
        "Control" = "#4D4D4D",
        "Stress"  = "#FE8A13"
      )
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0.05, 0.16))
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        hjust = 0.5
      ),
      axis.text.x = element_text(
        size = 16,
        face = "bold"
      ),
      axis.text.y = element_text(size = 10),
      strip.text = element_text(
        size = 14,
        face = "bold"
      ),
      axis.title.y = element_text(
        size = 16,
        face = "bold"
      ),
      legend.position = "none"
    ) +
    labs(
      x = NULL,
      y = "Activity Score",
      title = paste(
        brain_region,
        "- Estrogen Receptor Activity"
        )
      )
 # ----- Add significance brackets and stars -----
if (!is.null(sig_df) && nrow(sig_df) > 0) {
  
  p <- p +
    ggpubr::stat_pvalue_manual(
      sig_df,
      label = "sig",
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      bracket.size = 0.8,
      size = 7,
      vjust = -0.1,
      color = "black",
      hide.ns = FALSE,
      inherit.aes = FALSE
    )
    }
  return(p)
}
```
 
Run the plotting function
```{r}
p_esr <- plot_tf_violin(
  tf_scores = tf_per_sample_grouped,
  tf_list = c("Esr1"),
  brain_region = "CORT"
)

p_esr
```

Save output as a pdf
```{r}
ggsave(
  "CORT_ESR_TF_activity_violin.pdf",
  p_esr,
  width = 5,
  height = 7
)
```


