---
title: "TF_activity_inference"
author: "Amelia Slusser"
date: "2026-05-19"
output: html_document
---
################################################################################
Tools: decoupleR (DoRothEA regulons)
Input: global proteomics from female mice
Output: Active TFs
################################################################################
 
==============================================================================
0. INSTALL AND LOAD PACKAGES
==============================================================================

```{r}
library(tidyverse)
library(decoupleR)   
library(OmnipathR)   
library(readxl)
library(readr)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(ggplot2)
```

==============================================================================
1. LOAD AND PREPARE PROTEOMICS DATA
==============================================================================
 --- What your input file should look like ---
 A tab- or comma-separated file with at minimum:
  - A column of mouse gene symbols (e.g. "Gene.names" from MaxQuant)
   - A column of quantitative values per sample
     (LFQ intensity, iBAQ, TMT ratio, or fold-change vs. control)

 Example structure (MaxQuant proteinGroups.txt):
   Gene.names | LFQ.intensity.sample1 | LFQ.intensity.sample2 | ...
   Camk2a     | 4.2e8                 | 3.9e8                 | ...

 Adjust the path and column names below to match your actual file.
```{r}
df_raw <- as.data.frame(read_xlsx("data.xlsx"))
```

 --- Create row names ---
```{r}
rownames(df_raw) <- df_raw[,1]
df_raw <- df_raw[,-1]
```
 
# --- Convert to matrix ---
 Rows = genes, columns = samples
```{r}
cort_mat <- as.matrix(df_raw)
```
 
 Quick sanity check: should print gene symbols and sample names
```{r}
cat("Matrix dimensions:", dim(cort_mat), "\n")
cat("Sample names:", colnames(cort_mat), "\n")
cat("First few gene symbols:", head(rownames(cort_mat)), "\n")
```

 ==============================================================================
 2. DOROTHEA: INFER DIFFERENTIAL TF ACTIVITY
 ==============================================================================
   Per-sample: run decoupleR on all 12 samples, then test whether scores
       differ between groups. Gives richer statistics and per-sample variance

 --- Load DoRothEA mouse regulons ---
 get_collectri() is the recommended up-to-date regulon source in decoupleR.
 It supersedes the older DoRothEA v2 but uses the same confidence-tier logic.
 organism = 9606 is human; use 10090 for mouse.

```{r}
regulons <- get_collectri(organism = "10090", split_complexes = FALSE)
```

 Preview regulon structure:
 source = TF gene symbol, target = regulated gene, mor = mode of regulation (+1 activated, -1 repressed)
```{r}
head(regulons)
```

Export regulon table (only need this once)
```{r}
write_tsv(regulons, "DoRothEA_regulons_all.tsv")
```

 --- Per-sample scores then group comparison ---
Define which columns belong to which group 
```{r}
control_cols <- c("CORT_1L",	"CORT_1N",	"CORT_1R",	"CORT_3L",	"CORT_3N",	"CORT_3R")
 
stress_cols  <- c("CORT_5",	"CORT_6",	"CORT_7",	"CORT_11",	"CORT_12",	"CORT_13")
 
all_sample_cols <- c(control_cols, stress_cols)
```

Run all samples together using MLM
```{r}
tf_per_sample <- run_mlm(
  mat     = cort_mat,
  net     = regulons,
  .source = "source",
  .target = "target",
  .mor    = "mor",
  sparse  = FALSE,
  center  = FALSE,
  na.rm   = FALSE,
  minsize = 5
)
```

 Add group labels to per-sample scores
```{r}
group_labels <- tibble(
  condition = c(control_cols, stress_cols),
  group     = c(rep("control", length(control_cols)),
                rep("stress",  length(stress_cols)))
)
 
tf_per_sample_grouped <- tf_per_sample %>%
  left_join(group_labels, by = "condition")
```
 
 Test TF activity difference between groups (Welch t-test per TF)
```{r}
tf_group_test <- tf_per_sample_grouped %>%
  group_by(source) %>%
  summarise(
    mean_ctrl       = mean(score[group == "control"], na.rm = TRUE),
    mean_stress     = mean(score[group == "stress"],  na.rm = TRUE),
    sd_ctrl         = sd(score[group == "control"],   na.rm = TRUE),
    sd_stress       = sd(score[group == "stress"],    na.rm = TRUE),
    p_value         = tryCatch(
      t.test(score[group == "stress"],
             score[group == "control"])$p.value,
      error = function(e) NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    p_adj        = p.adjust(p_value, method = "BH"),
    delta_score  = mean_stress - mean_ctrl,
    direction    = if_else(delta_score > 0,
                           "more_active_in_stress", "less_active_in_stress")
  ) %>%
  filter(!is.na(p_value)) %>%
  arrange(p_adj)
 
sig_tf_group <- tf_group_test %>%
  filter(p_adj < 0.05, abs(delta_score) > 0.5) 

sig_tf_group <- sig_tf_group %>%
  rename(TF = source)
 
cat("\nDifferentially active TFs (per-sample method):", nrow(sig_tf_group), "\n")

write_tsv(tf_per_sample_grouped, "TF_per_sample_grouped.tsv")
write_tsv(tf_group_test, "TF_activity_per_group.tsv")
write_tsv(sig_tf_group, "Sig_only_TF_activity_per_group.tsv")
```

 ==============================================================================
 3. VISUALIZE DIFFERENTIAL TF ACTIVITY
 ==============================================================================
 
 --- Volcano-style plot of TF activity scores ---
```{r}
ggplot(tf_group_test %>% filter(!is.na(p_adj)),
       aes(x = delta_score, y = -log10(p_adj),
           color = p_adj < 0.05 & abs(delta_score) > 0.5)) +
  geom_point(alpha = 0.7, size = 2) +
  geom_hline(yintercept = -log10(0.05), linetype = "dashed", alpha = 0.5) +
  geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", alpha = 0.5) +
  ggrepel::geom_text_repel(
    data = . %>% filter(p_adj < 0.05, abs(delta_score) > 0.5),
    aes(label = source), size = 3, max.overlaps = 20
  ) +
  scale_color_manual(values = c("FALSE" = "gray70", "TRUE" = "#1D9E75"),
                     guide = "none") +
  labs(title  = "Differential TF activity: stress vs. control",
       x      = "Delta activity score (stress - control)",
       y      = "-log10(adjusted p-value)") +
  theme_minimal(base_size = 12)
 
ggsave("TF_activity_volcano.pdf", width = 8, height = 6)
```
 
  --- Heatmap of top TF activity scores per sample (Fig. S6B & C) ---
```{r}
# ── Setup ──────────────────────────────────────────────────────────────

sample_order <- c(control_cols, stress_cols)

group_vector <- c(rep("Control", length(control_cols)),
                  rep("Stress",  length(stress_cols)))
names(group_vector) <- sample_order

# Only keep samples that are actually present in tf_per_sample
sample_order_present <- intersect(sample_order, unique(tf_per_sample$condition))
group_vector_present <- group_vector[sample_order_present]

ha_col <- HeatmapAnnotation(
  Group = factor(
    group_vector_present,
    levels = c("Control", "Stress")
  ),
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress  = "#FE8A13"
    )
  ),
  border = TRUE
)

# Helper function to safely build heatmap matrix from MLM output
make_tf_heatmap_matrix <- function(tf_names, tf_per_sample, sample_order_present) {
  
  heatmap_data <- tf_per_sample %>%
    filter(source %in% tf_names) %>%
    select(source, sample = condition, score) %>%
    pivot_wider(
      names_from = sample,
      values_from = score
    ) %>%
    column_to_rownames("source") %>%
    as.matrix()
  
  # Keep only expected samples, preserving order
  heatmap_data <- heatmap_data[
    ,
    intersect(sample_order_present, colnames(heatmap_data)),
    drop = FALSE
  ]
  
  # Remove rows with all NA values
  heatmap_data <- heatmap_data[rowSums(!is.na(heatmap_data)) > 0, , drop = FALSE]
  
  # Row-scale
  heatmap_data_scaled <- t(scale(t(heatmap_data)))
  
  # Replace NaN values caused by zero variance rows
  heatmap_data_scaled[is.nan(heatmap_data_scaled)] <- 0
  
  return(heatmap_data_scaled)
}

# ══════════════════════════════════════════════════════════════════════════════
# HEATMAP 1 - Top 50 TFs by absolute delta score
# ══════════════════════════════════════════════════════════════════════════════

group_tf_names <- tf_group_test %>%
  slice_max(abs(delta_score), n = 50) %>%
  pull(source)

heatmap_data_scaled <- make_tf_heatmap_matrix(
  tf_names = group_tf_names,
  tf_per_sample = tf_per_sample,
  sample_order_present = sample_order_present
)

# Stop early with a useful message if no TFs are available
if (nrow(heatmap_data_scaled) == 0) {
  stop("No TFs available for Heatmap 1. Check whether tf_per_sample$source matches tf_group_test$source.")
}

sig_df <- tf_group_test %>%
  filter(source %in% rownames(heatmap_data_scaled)) %>%
  distinct(source, p_adj, delta_score) %>%
  mutate(
    sig_label = case_when(
      p_adj < 0.05 ~ "*",
      TRUE ~ ""
    )
  )

sig_df <- sig_df[
  match(rownames(heatmap_data_scaled), sig_df$source),
]

p_col_fun <- colorRamp2(
  c(1, 0.05, 0.01, 0),
  c("#F0F0F0", "#FDBBA1", "#FB6A4A", "#EF3B2C")
)

ha_row <- rowAnnotation(
  
  Adj_FDR = anno_simple(
    sig_df$p_adj,
    col = p_col_fun,
    pch = sig_df$sig_label,
    pt_gp = grid::gpar(
      fontsize = 9,
      fontface = "bold",
      col = "black"
    ),
    width = unit(3, "mm"),
    border = TRUE
  ),
  
  "Delta activity score\n(stress−ctrl)" = anno_barplot(
    sig_df$delta_score,
    gp = grid::gpar(
      fill = ifelse(sig_df$delta_score > 0,
                    "#FE8A13",
                    "#4D4D4D"),
      col = NA
    ),
    axis_param = list(
      gp = grid::gpar(fontsize = 7)
    ),
    width = unit(2.5, "cm")
  ),
  
  show_annotation_name = c(Adj_FDR = FALSE)
)

ht1 <- Heatmap(
  heatmap_data_scaled,
  name               = "TF Activity\n(row-scaled)",
  top_annotation     = ha_col,
  right_annotation   = ha_row,
  col                = viridis(100),
  width              = unit(8, "cm"),
  cluster_columns    = FALSE,
  cluster_rows       = TRUE,
  show_row_names     = TRUE,
  show_column_names  = TRUE,
  row_names_gp       = grid::gpar(fontsize = 9),
  column_names_gp    = grid::gpar(fontsize = 9),
  column_title       = "TF Activity"
)

pdf("TF_activity_heatmap_all_MLM.pdf", width = 8, height = 8)
draw(ht1, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

# ══════════════════════════════════════════════════════════════════════════════
# HEATMAP 2 - Significant TFs only
# ══════════════════════════════════════════════════════════════════════════════

group_tf_names_sig <- sig_tf_group %>%
  slice_max(abs(delta_score), n = 50) %>%
  pull(TF)

# Only make this heatmap if significant TFs exist
if (length(group_tf_names_sig) == 0) {
  
  message("No significant TFs found using p_adj < 0.05 and abs(delta_score) > 0.5. Skipping significant-only heatmap.")
  
} else {
  
  heatmap_data_scaled_sig <- make_tf_heatmap_matrix(
    tf_names = group_tf_names_sig,
    tf_per_sample = tf_per_sample,
    sample_order_present = sample_order_present
  )
  
  if (nrow(heatmap_data_scaled_sig) == 0) {
    stop("No TFs available for Heatmap 2 after filtering significant TFs.")
  }
  
  delta_scores_sig <- sig_tf_group %>%
    filter(TF %in% rownames(heatmap_data_scaled_sig)) %>%
    arrange(match(TF, rownames(heatmap_data_scaled_sig))) %>%
    pull(delta_score)
  
  ha_row_sig <- rowAnnotation(
    "Delta activity score\n(stress−ctrl)" = anno_barplot(
      delta_scores_sig,
      gp = grid::gpar(
        fill = ifelse(delta_scores_sig > 0,
                      "#FE8A13",
                      "#4D4D4D"),
        col = NA
      ),
      axis_param = list(
        gp = grid::gpar(fontsize = 7)
      ),
      width = unit(2.5, "cm")
    )
  )
  
  ht2 <- Heatmap(
    heatmap_data_scaled_sig,
    name               = "TF Activity\n(row-scaled)",
    top_annotation     = ha_col,
    right_annotation   = ha_row_sig,
    col                = viridis(100),
    width              = unit(8, "cm"),
    cluster_columns    = FALSE,
    cluster_rows       = TRUE,
    show_row_names     = TRUE,
    show_column_names  = TRUE,
    row_names_gp       = grid::gpar(fontsize = 9),
    column_names_gp    = grid::gpar(fontsize = 9),
    column_title       = "TF Activity, significant by group comparison"
  )
  
  pdf("TF_activity_heatmap_sig_MLM.pdf", width = 10, height = 8)
  draw(ht2, heatmap_legend_side = "right", annotation_legend_side = "right")
  dev.off()
}
```

Export the heat map data
```{r}
# Convert scaled heatmap matrix to a data frame
# and preserve TF names in the first column
heatmap_df <- heatmap_data_scaled %>%
  as.data.frame() %>%
  tibble::rownames_to_column(var = "TF")

# Export to Excel
writexl::write_xlsx(
  heatmap_df,
  path = "TF_activity_heatmap_data_scaled.xlsx"
)
```

 ==============================================================================
 4. EXPLAIN WHICH PROTEINS DROVE EACH TF INFERENCE
 ==============================================================================
 For every significant TF, show which detected proteins contributed to its
 inferred activity score, and how that contribution differs between groups.
```{r}
explain_tf_drivers <- function(tf_name, regulons, pfc_mat,
                               ctrl_cols, stress_cols) {
  # Targets of this TF in the regulon
  tf_targets <- regulons %>%
    filter(source == tf_name) %>%
    select(target, mor)
 
  # Targets that appear in the data
  detected <- tf_targets %>%
    filter(target %in% rownames(pfc_mat)) %>%
    mutate(
      mean_ctrl   = rowMeans(pfc_mat[target, ctrl_cols,   drop = FALSE],
                             na.rm = TRUE),
      mean_stress = rowMeans(pfc_mat[target, stress_cols, drop = FALSE],
                             na.rm = TRUE),
      log2FC      = mean_stress - mean_ctrl,
      # Weighted contribution: direction of regulation × fold-change
      # Positive = supports TF being more active in stress
      # Negative = supports TF being less active in stress
      weighted_contribution = mor * log2FC
    ) %>%
    arrange(desc(abs(weighted_contribution)))
 
  detected
}
```

```{r}
# Run for all TFs and save a combined driver table
all_drivers <- map_dfr(sig_tf_group$TF, function(tf) {
  explain_tf_drivers(tf, regulons, pfc_mat, control_cols, stress_cols) %>%
    mutate(
      inferred_TF   = tf,
      TF_direction  = sig_tf_group$direction[sig_tf_group$TF == tf],
      delta_score   = sig_tf_group$delta_score[sig_tf_group$TF == tf]
    )
}, .progress = TRUE)

# Flag the most influential drivers: top contributors per TF
key_drivers <- all_drivers %>%
  group_by(inferred_TF) %>%
  slice_max(abs(weighted_contribution), n = 10) %>%
  ungroup() %>%
  arrange(inferred_TF, desc(abs(weighted_contribution)))

write_tsv(all_drivers, "TF_driver_proteins_all_group.tsv")
write_tsv(key_drivers, "TF_driver_proteins_top10_group.tsv")

cat("\nDriver tables saved.\n")
cat("Columns in driver table:\n")
cat("  target                = detected protein that contributed\n")
cat("  mor                   = mode of regulation (+1 activated, -1 repressed)\n")
cat("  mean_ctrl/mean_stress = log2 abundance in each group\n")
cat("  log2FC                = stress - control log2 abundance difference\n")
cat("  weighted_contribution = mor x log2FC (positive = supports TF being\n")
cat("                          more active in stress)\n")
cat("  inferred_TF           = which TF this protein helped infer\n")
cat("  TF_direction          = more/less_active_in_stress\n")
cat("  delta_score           = delta score\n")
```

 ══════════════════════════════════════════════════════════════════════════════
 5. Specific TF violin plots (Fig. 6B & C)
 ══════════════════════════════════════════════════════════════════════════════
 Plotting function
```{r}
plot_tf_violin <- function(tf_scores,
                           tf_list = c("Esr1"),
                           brain_region = "CORT",
                           sex_label = "F") {
  # Filter selected TFs
  df <- tf_scores %>%
    filter(source %in% tf_list) %>%
    mutate(
      group = factor(group,
                     levels = c("control", "stress"),
                     labels = c("Control", "Stress")),
      TF = factor(source,
                  levels = tf_list),
      Group_num = as.numeric(group)
    )
  # ----- Statistics -----
  sig_list <- lapply(levels(df$TF), function(tf) {
    df_tf <- df %>%
      filter(TF == tf)
    val1 <- df_tf$score[df_tf$group == "Control"]
    val2 <- df_tf$score[df_tf$group == "Stress"]
    if(length(val1) > 1 & length(val2) > 1) {
      test <- wilcox.test(
        val1,
        val2,
        exact = FALSE
      )
      p <- test$p.value
      sig <- ifelse(
        p < 0.001, "***",
        ifelse(
          p < 0.01, "**",
          ifelse(
            p < 0.05, "*", "ns"
          )
        )
      )
      p_label <- ifelse(
        p < 0.001,
        paste0(sig),
        paste0(sig)
      )
      return(
        data.frame(
          TF = tf,
          x_start = 1,
          x_end = 2,
          label = p_label,
          sig = sig,
          p_value = p
        )
      )
    }
    return(NULL)
  })

  sig_df <- bind_rows(sig_list) %>%
  mutate(
    group1 = x_start,
    group2 = x_end
  )
  
  # Calculate an annotation position separately for each TF
tf_y_positions <- df %>%
  group_by(TF) %>%
  summarise(
    y_min = min(score, na.rm = TRUE),
    y_max = max(score, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    y_range = y_max - y_min,
    y_range = if_else(y_range > 0, y_range, 1),
    
    # Position bracket slightly above the highest observation
    y.position = y_max + 0.08 * y_range
  )

sig_df <- sig_df %>%
  left_join(tf_y_positions, by = "TF")

   # ----- Base Plot -----
  p <- ggplot(
    df,
    aes(
      x = Group_num,
      y = score,
      fill = group
    )
  ) +
    geom_violin(
      trim = FALSE,
      alpha = 0.7
    ) +
    geom_boxplot(
      width = 0.1,
      outlier.shape = NA,
      alpha = 0.5,
      color = "black"
    ) +
    geom_jitter(
      width = 0.15,
      alpha = 0.7,
      size = 2,
      color = "black"
    ) +
    facet_wrap(
      ~ TF,
      scales = "free_y"
    ) +
    scale_x_continuous(
      breaks = 1:2,
      labels = levels(df$group)
    ) +
    scale_fill_manual(
      values = c(
        "Control" = "#4D4D4D",
        "Stress"  = "#FE8A13"
      )
    ) +
    scale_y_continuous(
      expand = expansion(mult = c(0.05, 0.16))
    ) +
    theme_bw() +
    theme(
      plot.title = element_text(
        size = 16,
        face = "bold",
        hjust = 0.5
      ),
      axis.text.x = element_text(
        size = 16,
        face = "bold"
      ),
      axis.text.y = element_text(size = 10),
      strip.text = element_text(
        size = 14,
        face = "bold"
      ),
      axis.title.y = element_text(
        size = 16,
        face = "bold"
      ),
      legend.position = "none"
    ) +
    labs(
      x = NULL,
      y = "Activity Score",
      title = paste(
        brain_region,
        "- Estrogen Receptor Activity"
        )
      )
 # ----- Add significance brackets and stars -----
if (!is.null(sig_df) && nrow(sig_df) > 0) {
  
  p <- p +
    ggpubr::stat_pvalue_manual(
      sig_df,
      label = "sig",
      xmin = "group1",
      xmax = "group2",
      y.position = "y.position",
      tip.length = 0.02,
      bracket.size = 0.8,
      size = 7,
      vjust = -0.1,
      color = "black",
      hide.ns = FALSE,
      inherit.aes = FALSE
    )
    }
  return(p)
}
```
 
Run the plotting function
```{r}
p_esr <- plot_tf_violin(
  tf_scores = tf_per_sample_grouped,
  tf_list = c("Esr1"),
  brain_region = "CORT"
)

p_esr
```

Save output as a pdf
```{r}
ggsave(
  "ESR_TF_activity_violin.pdf",
  p_esr,
  width = 5,
  height = 7
)
```

 ══════════════════════════════════════════════════════════════════════════════
 6. Bubble plot visualization of KEGG estrogen signaling pathway proteins (Fig. S6D & E)
 ══════════════════════════════════════════════════════════════════════════════
 KEGG estrogen signaling annotation was performed on detected proteins using the STRING database
```{r}
library(tidyverse)
library(readxl)
library(writexl)
```

################################## Hippocampus ##################################
Load data (Hippocampus)
```{r}
intensity_df <- read_excel("f_hpc_imputed_values.xlsx")
pval_df <- read_excel("f_hpc_pca_weighted_drivers.xlsx")
```

Prepare data to plot (Hippocampus)
```{r}
# Choose proteins to plot
proteins_to_plot <- c(
  "Akt1", "Atf2", "Ctsd", "Gnai2", "Gnaq", "Hspa1b", "Hspa8", "Hsp90aa1", "Hsp90ab1", "Hsp90b1", "Kras", "Map2k1", "Map2k2", "Pik3r2", "Plcb1")

# Define group replicate columns
control_cols <- c(
  "HPC_1R",
  "HPC_1L",
  "HPC_1N",
  "HPC_3R",
  "HPC_3L",
  "HPC_3N"
)

stress_cols <- c(
  "HPC_5",
  "HPC_6",
  "HPC_7",
  "HPC_11",
  "HPC_12",
  "HPC_13"
)

# Prepare intensity data
plot_intensity_df <- intensity_df %>%
  filter(Gene %in% proteins_to_plot) %>%
  rowwise() %>%
  mutate(
    Control_Avg = mean(c_across(all_of(control_cols)), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(
    Gene,
    Control_Avg,
    all_of(stress_cols)
  ) %>%
  pivot_longer(
    cols = all_of(stress_cols),
    names_to = "Stress_Replicate",
    values_to = "Stress_Intensity"
  ) %>%
  mutate(
    Replicate_log2FC = Stress_Intensity - Control_Avg
  )

# Adjusted p-vals
plot_pval_df <- pval_df %>%
  filter(Gene %in% proteins_to_plot) %>%
  select(
    Gene,
    adj.P.Val
  ) %>%
  mutate(
    neg_log10_adj_p = -log10(adj.P.Val)
  )

# Join intensity and p-value data + define significance
bubble_df <- plot_intensity_df %>%
  left_join(plot_pval_df, by = "Gene") %>%
  mutate(
    Significance = case_when(
      adj.P.Val < 0.05 & Replicate_log2FC > 0  ~ "Significant Up",
      adj.P.Val < 0.05 & Replicate_log2FC < -0 ~ "Significant Down",
      TRUE ~ "Not significant"
    ),
    Significant = if_else(Significance == "Not significant", "No", "Yes"),
    Gene = factor(Gene, levels = rev(proteins_to_plot)),
    Stress_Replicate = factor(Stress_Replicate, levels = stress_cols)
  )

# Save plot data
write_xlsx(
  bubble_df,
  path = "HPC_estrogen_df.xlsx"
)
```

Plot (Hippocampus)
```{r}
bubble_plot <- ggplot(
  bubble_df,
  aes(
    x = Stress_Replicate,
    y = Gene
  )
) +
  
  # All bubbles
  geom_point(
    aes(
      size = neg_log10_adj_p,
      fill = Replicate_log2FC
    ),
    shape = 21,
    color = "black",
    stroke = 0.3,
    alpha = 0.85
  ) +
  
  # Significant bubbles overlaid with thicker black outline
  geom_point(
    data = bubble_df %>%
      filter(Significant == "Yes"),
    aes(
      size = neg_log10_adj_p,
      fill = Replicate_log2FC
    ),
    shape = 21,
    color = "black",
    stroke = 1.4,
    alpha = 1
  ) +
  
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Replicate Log2FC\nvs Control Avg"
  ) +
  
  scale_size_continuous(
    name = expression(-Log[10]("adj. p-value")),
    range = c(2, 9)
  ) +
  
  # Force adjusted p-value legend above color legend
  guides(
    size = guide_legend(order = 1),
    fill = guide_colorbar(order = 2)
  ) +
  
  labs(
    x = "Stress replicate",
    y = NULL,
    title = "HPC detected estrogen signaling pathway proteins (KEGG: mmu04915)",
    caption = "Thick black outline = significant proteins: adj. p-value < 0.05 and group |Log2FC| > 2"
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0),
    legend.position = "right"
  )

bubble_plot
```

Save the plot (Hippocampus)
```{r}
ggsave(
  filename = "female_HPC_ES_protein_bubble_plot.pdf",
  plot = bubble_plot,
  width = 9.5,
  height = 6
)
```

################################## Cortex ##################################
Load data (Cortex)
```{r}
intensity_df <- read_excel("f_cort_imputed_values.xlsx")
pval_df <- read_excel("f_cort_pca_weighted_drivers.xlsx")
```

Prepare data to plot (Cortex)
```{r}
# Choose proteins to plot
proteins_to_plot <- c(
  "Adcy9", "Atf2", "Ctsd", "Fkbp5", "Gnai1", "Gnai2", "Gnao1", "Gnaq", "Hspa1a", "Hspa1b", "Hspa8", "Hsp90aa1", "Hsp90ab1", "Hsp90b1", "Kras", "Map2k2", "Nras", "Plcb1")

# Define group replicate columns
control_cols <- c(
  "CORT_1R",
  "CORT_1L",
  "CORT_1N",
  "CORT_3R",
  "CORT_3L",
  "CORT_3N"
)

stress_cols <- c(
  "CORT_5",
  "CORT_6",
  "CORT_7",
  "CORT_11",
  "CORT_12",
  "CORT_13"
)

# Prepare intensity data
plot_intensity_df <- intensity_df %>%
  filter(Gene %in% proteins_to_plot) %>%
  rowwise() %>%
  mutate(
    Control_Avg = mean(c_across(all_of(control_cols)), na.rm = TRUE)
  ) %>%
  ungroup() %>%
  select(
    Gene,
    Control_Avg,
    all_of(stress_cols)
  ) %>%
  pivot_longer(
    cols = all_of(stress_cols),
    names_to = "Stress_Replicate",
    values_to = "Stress_Intensity"
  ) %>%
  mutate(
    Replicate_log2FC = Stress_Intensity - Control_Avg
  )

# Adjusted p-vals
plot_pval_df <- pval_df %>%
  filter(Gene %in% proteins_to_plot) %>%
  select(
    Gene,
    adj.P.Val
  ) %>%
  mutate(
    neg_log10_adj_p = -log10(adj.P.Val)
  )

# Join intensity and p-value data + define significance
bubble_df <- plot_intensity_df %>%
  left_join(plot_pval_df, by = "Gene") %>%
  mutate(
    Significance = case_when(
      adj.P.Val < 0.05 & Replicate_log2FC > 0  ~ "Significant Up",
      adj.P.Val < 0.05 & Replicate_log2FC < -0 ~ "Significant Down",
      TRUE ~ "Not significant"
    ),
    Significant = if_else(Significance == "Not significant", "No", "Yes"),
    Gene = factor(Gene, levels = rev(proteins_to_plot)),
    Stress_Replicate = factor(Stress_Replicate, levels = stress_cols)
  )

# Save plot data
write_xlsx(
  bubble_df,
  path = "CORT_estrogen_df.xlsx"
)
```

Plot
```{r}
bubble_plot <- ggplot(
  bubble_df,
  aes(
    x = Stress_Replicate,
    y = Gene
  )
) +
  
  # All bubbles
  geom_point(
    aes(
      size = neg_log10_adj_p,
      fill = Replicate_log2FC
    ),
    shape = 21,
    color = "black",
    stroke = 0.3,
    alpha = 0.85
  ) +
  
  # Significant bubbles overlaid with thicker black outline
  geom_point(
    data = bubble_df %>%
      filter(Significant == "Yes"),
    aes(
      size = neg_log10_adj_p,
      fill = Replicate_log2FC
    ),
    shape = 21,
    color = "black",
    stroke = 1.4,
    alpha = 1
  ) +
  
  scale_fill_gradient2(
    low = "#2166AC",
    mid = "white",
    high = "#B2182B",
    midpoint = 0,
    name = "Replicate Log2FC\nvs Control Avg"
  ) +
  
  scale_size_continuous(
    name = expression(-Log[10]("adj. p-value")),
    range = c(2, 9)
  ) +
  
  labs(
    x = "Stress replicate",
    y = NULL,
    title = "CORT detected estrogen signaling pathway proteins (KEGG: mmu04915)",
    caption = "Thick black outline = significant proteins: adj. p-value < 0.05 and group |Log2FC| > 2"
  ) +
  
  theme_bw(base_size = 13) +
  theme(
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.caption = element_text(hjust = 0),
    legend.position = "right"
  )

bubble_plot
```

Save the plot (Cortex)
```{r}
ggsave(
  filename = "female_CORT_ES_protein_bubble_plot.pdf",
  plot = bubble_plot,
  width = 9.5,
  height = 7
)
```
