---
title: "brain_proteomics_visualizations"
author: "Amelia Slusser"
date: "2026-02-23"
output: html_document
---
################################################################################
Male Hippocampus
################################################################################

-------------- Statistical results dataframe -------------- 
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("m_hpc_imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("m_hpc_pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "m_hpc_stats_with_reps.xlsx"
)
```

-------------- Volcano plot (Fig. 1D) --------------
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("m_hpc_pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "M HPC"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "plain"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "male_hpc_volcano_plot.pdf",
  plot = volcano_plot,
  width = 7,
  height = 5,
  units = "in"
)
```


-------------- Box and whisker plots for significant stress-related proteins (Fig. 1H) --------------
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
library(ggnewscale)
```

Load in data and choose proteins
```{r}
# Read in values spreadsheet
intensity_df <- read_excel("m_hpc_imputed_values.xlsx")

# Load statistics spreadsheet
stats_df <- read_excel("m_hpc_filtered_results_w_function.xlsx")

# Choose genes to plot
genes_to_plot <- c(
  "Hspb1",
  "Cpe",
  "App",
  "Hyou1",
  "Txnrd1",
  "Syn1",
  "Syp",
  "Dlg4",
  "Baiap2",
  "Gap43"
)

# Define sample columns
sample_cols <- grep("^HPC_", colnames(intensity_df), value = TRUE)

# Make alphabetical gene order
gene_order <- sort(genes_to_plot)

plot_df <- intensity_df %>%
  filter(Gene %in% genes_to_plot) %>%
  mutate(
    Gene = factor(Gene, levels = gene_order)
  ) %>%
  pivot_longer(
    cols = all_of(sample_cols),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs|Stress") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```

Create significance brackets
```{r}
# Create significance brackets from stats dataframe
bracket_df <- stats_df %>%
  dplyr::filter(Gene %in% genes_to_plot) %>%
  dplyr::mutate(
    Gene = factor(Gene, levels = gene_order),
    
    sig_label = dplyr::case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.05  ~ "*",
      TRUE              ~ "ns"
    )
  ) %>%
  dplyr::select(Gene, adj.P.Val, sig_label)

# Set y-positions for brackets
y_positions <- plot_df %>%
  dplyr::group_by(Gene) %>%
  dplyr::summarise(
    y.position = max(Intensity, na.rm = TRUE) * 1.15,
    .groups = "drop"
  )

# Join y-positions and set bracket positions
bracket_df <- bracket_df %>%
  dplyr::left_join(y_positions, by = "Gene") %>%
  dplyr::mutate(
    group1 = "Control",
    group2 = "Stress",
    x_num = match(as.character(Gene), gene_order),
    xmin = x_num - 0.18,
    xmax = x_num + 0.18
  )
```

Plot
```{r}
protein_boxplot <- ggplot(plot_df, aes(x = Gene, y = Intensity)) +
  
  # Light boxplot fill
  geom_boxplot(
    aes(fill = Group),
    position = position_dodge(width = 0.75),
    width = 0.6,
    outlier.shape = NA,
    color = "black",
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    ),
    name = "Group"
  ) +
  
  # Start a second fill scale for the replicate dots
  ggnewscale::new_scale_fill() +
  
  # Dark replicate dots with black border
  geom_jitter(
    aes(fill = Group),
    shape = 21,
    color = "black",
    stroke = 0.35,
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 2.5,
    alpha = 0.95,
    show.legend = FALSE
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  # Significance brackets and stars
  stat_pvalue_manual(
    bracket_df,
    label = "sig_label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.6,
    size = 5,
    hide.ns = FALSE
  ) +
  
  labs(
    title = "M HPC",
    x = "Gene",
    y = "Protein intensity",
    fill = "Group"
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.18))
  ) +
  
  coord_cartesian(clip = "off") +
  
  theme_classic(base_size = 14) +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "right",
    plot.margin = margin(10, 20, 10, 10)
  )

protein_boxplot
```

Save plot
```{r}
ggsave(
  filename = "male_hpc_selected_prot.pdf",
  plot = protein_boxplot,
  width = 7,
  height = 5,
  units = "in"
)
```

################################################################################
Male Cortex
################################################################################

-------------- Statistical results dataframe -------------- 
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("m_cort_imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("m_cort_pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "m_cort_stats_with_reps.xlsx"
)
```

-------------- Volcano plot (Fig. 1F) --------------
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("m_cort_pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "M CORT"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "plain"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "male_cort_volcano_plot.pdf",
  plot = volcano_plot,
  width = 7,
  height = 5,
  units = "in"
)
```

################################################################################
Female Hippocampus
################################################################################

-------------- Statistical results dataframe -------------- 
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("f_hpc_imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("f_hpc_pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "f_hpc_stats_with_reps.xlsx"
)
```

-------------- Volcano plot (Fig. 1E) -------------- 
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("f_hpc_pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "F HPC"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "plain"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "female_hpc_volcano_plot_big.pdf",
  plot = volcano_plot,
  width = 7,
  height = 5,
  units = "in"
)
```

-------------- Box and whisker plots for statistically significant stress-related proteins (Fig. 1I) --------------
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
library(ggnewscale)
```

Load in data and choose proteins
```{r}
# Read in values spreadsheet
intensity_df <- read_excel("f_hpc_imputed_values.xlsx")

# Load statistics spreadsheet
stats_df <- read_excel("f_hpc_filtered_results_w_function.xlsx")

# Choose genes to plot
genes_to_plot <- c(
  "Apod",
  "Hsp90b1",
  "Hmox2",
  "Vcp",
  "Gad1",
  "Syn1",
  "Map2",
  "S100b",
  "Gria1",
  "Ldha"
)

# Define sample columns
sample_cols <- grep("^HPC_", colnames(intensity_df), value = TRUE)

# Make alphabetical gene order
gene_order <- sort(genes_to_plot)

plot_df <- intensity_df %>%
  filter(Gene %in% genes_to_plot) %>%
  mutate(
    Gene = factor(Gene, levels = gene_order)
  ) %>%
  pivot_longer(
    cols = all_of(sample_cols),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs|Stress") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```

Create significance brackets
```{r}
# Create significance brackets from stats dataframe
bracket_df <- stats_df %>%
  dplyr::filter(Gene %in% genes_to_plot) %>%
  dplyr::mutate(
    Gene = factor(Gene, levels = gene_order),
    
    sig_label = dplyr::case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.05  ~ "*",
      TRUE              ~ "ns"
    )
  ) %>%
  dplyr::select(Gene, adj.P.Val, sig_label)

# Set y-positions for brackets
y_positions <- plot_df %>%
  dplyr::group_by(Gene) %>%
  dplyr::summarise(
    y.position = max(Intensity, na.rm = TRUE) * 1.15,
    .groups = "drop"
  )

# Join y-positions and set bracket positions
bracket_df <- bracket_df %>%
  dplyr::left_join(y_positions, by = "Gene") %>%
  dplyr::mutate(
    group1 = "Control",
    group2 = "Stress",
    x_num = match(as.character(Gene), gene_order),
    xmin = x_num - 0.18,
    xmax = x_num + 0.18
  )
```

Plot
```{r}
protein_boxplot <- ggplot(plot_df, aes(x = Gene, y = Intensity)) +
  
  # Light boxplot fill
  geom_boxplot(
    aes(fill = Group),
    position = position_dodge(width = 0.75),
    width = 0.6,
    outlier.shape = NA,
    color = "black",
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    ),
    name = "Group"
  ) +
  
  # Start a second fill scale for the replicate dots
  ggnewscale::new_scale_fill() +
  
  # Dark replicate dots with black border
  geom_jitter(
    aes(fill = Group),
    shape = 21,
    color = "black",
    stroke = 0.35,
    position = position_jitterdodge(
      jitter.width = 0.12,
      dodge.width = 0.75
    ),
    size = 2.5,
    alpha = 0.95,
    show.legend = FALSE
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  # Significance brackets and stars
  stat_pvalue_manual(
    bracket_df,
    label = "sig_label",
    xmin = "xmin",
    xmax = "xmax",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.6,
    size = 5,
    hide.ns = FALSE
  ) +
  
  labs(
    title = "F HPC",
    x = "Gene",
    y = "Protein intensity",
    fill = "Group"
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.18))
  ) +
  
  coord_cartesian(clip = "off") +
  
  theme_classic(base_size = 14) +
  
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    legend.position = "right",
    plot.margin = margin(10, 20, 10, 10)
  )

protein_boxplot
```

Save plot
```{r}
ggsave(
  filename = "female_hpc_selected_prot.pdf",
  plot = protein_boxplot,
  width = 7,
  height = 5,
  units = "in"
)
```

################################################################################
Female Cortex
################################################################################

-------------- Statistical results dataframe -------------- 
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("f_cort_imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("f_cort_pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "f_cort_stats_with_reps.xlsx"
)
```

Load packages
```{r}
library(readxl)
library(dplyr)
library(writexl)
```

Create selected proteins spread sheet (if needed)
```{r}
all_proteins <- read_excel("f_cort_imputed_values.xlsx")
sig_proteins <- read_excel("f_cort_results_w_function.xlsx")

# Filter only significant proteins from imputed values
filtered_proteins <- all_proteins %>%
  filter(Protein %in% sig_proteins$Protein)

# Write new file
write_xlsx(filtered_proteins,
           "f_pfc_sig.xlsx")
```

-------------- Volcano plot (Fig. 1G) -------------- 
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("f_cort_pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "F CORT"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "plain"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "female_cort_volcano_plot.pdf",
  plot = volcano_plot,
  width = 7,
  height = 5,
  units = "in"
)
```
