---
title: "metaproteomics_clean_up"
author: "Amelia Slusser"
date: "2026-04-29"
output: html_document
---
################################################################################
Statistical results df
################################################################################
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "stats_with_reps.xlsx"
)
```

################################################################################
Volcano plot
################################################################################
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "F LI"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "female_li_volcano_plot.pdf",
  plot = volcano_plot,
  width = 7,
  height = 5,
  units = "in"
)
```

################################################################################
GH family heat map
################################################################################
```{r}
library(dplyr)
library(tibble)
library(readr)
library(readxl)
library(writexl)
```

Load in data frames
```{r}
results_df <- read.delim("female_li_8_2_filtered_hybrid_results_w_function.tsv")
imputed_df <- read.delim("imputed_values.tsv")
unique_df <- read.delim("female_li_8_2_ctab_unique.tsv")
```

Filter out the unique protein rows
```{r}
results_filtered <- results_df[!(results_df$Protein %in% unique_df$Protein), ]
imputed_filtered <- imputed_df[!(imputed_df$Protein %in% unique_df$Protein), ]
```

Can also add status column and classify proteins based on unique vs shared
```{r}
results_df$status <- ifelse(results_df$Protein %in% unique_df$Protein,
                            "unique",
                            "shared")
```

Save filtered data frames
```{r}
write_tsv(results_filtered, file = "female_li_8_2_filtered_hybrid_results_w_function_shared.tsv")
write_tsv(imputed_filtered, file = "imputed_values_shared.tsv")
```

Load in excel file
```{r}
results_excel <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "microbe")
```

Add status column to excel sheet
```{r}
results_excel$status <- ifelse(results_excel$Protein %in% unique_df$Protein,
                     "unique",
                     "shared")
```

Write file back to excel
```{r}
write_xlsx(list(microbe_updated = results_excel), "updated_microbe.xlsx")
```

Plot select data as a heatmap with replicates per group
Load in packages
```{r}
library(readxl)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)
library(gridtext)
library(stringr)
library(scales)
```

Load in data from excel spread sheets
```{r}
rep_df <- read_excel("imputed_values_female.xlsx")
metadat_df <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "complx_carb_ferm_selected")
```

Filter imputed values to only proteins of interest for plotting
```{r}
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)
# Match order between sheets to align metadata (safer ordering)
filtered_rep_df <- filtered_rep_df %>%
  slice(match(metadat_df$Protein, Protein))
```

Save sig with reps file
```{r}
write_xlsx(filtered_rep_df, "f_li_gh_sig_reps.xlsx")
```

Create dataframe that will be plotted as a heatmap
```{r}
heatmap_data <- filtered_rep_df %>%
  select(starts_with("LI_")) %>%
  as.data.frame()

# Row-wise z-score scaling
heatmap_scaled <- t(scale(t(heatmap_data)))

# Assign row names
rownames(heatmap_scaled) <- metadat_df$name
```

Order data by GH family
```{r}
# Manually define GH family order
gh_order <- c(
  "GH1",
  "GH2",
  "GH3",
  "GH13",
  "GH29",
  "GH30",
  "GH31",
  "GH32",
  "GH36",
  "GH38",
  "GH42",
  "GH43",
  "GH51"
)
# Convert to a factor
metadat_df$`GH family` <- factor(
  metadat_df$`GH family`,
  levels = gh_order
)
# Order heatmap rows
order_idx <- order(metadat_df$`GH family`)

heatmap_scaled <- heatmap_scaled[order_idx, ]
metadat_df <- metadat_df[order_idx, ]
```

Create italicized labels
```{r}
formatted_labels <- sapply(metadat_df$name, function(x) {

  parts <- strsplit(x, " - ")[[1]]

  species <- parts[1]

  remainder <- ifelse(length(parts) > 1,
                      paste(parts[-1], collapse = " - "),
                      "")

  paste0("<i>", species, "</i> - ", remainder)
})
```

Annotate columns
```{r}
group <- c(rep("Control", 4), rep("Stress", 4))

ha_col <- HeatmapAnnotation(
  Group = group,
  
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress = "#FE8A13"
    )
  ),
  
  border = TRUE
)
```

GH family coloring
```{r}
gh_families <- levels(droplevels(metadat_df$`GH family`))

set.seed(1)

gh_colors <- structure(
  sample(rainbow(length(gh_families), s = 0.7, v = 0.85)),
  names = as.character(gh_families)
)

gh_legend <- Legend(
  title = "GH Family",
  labels = names(gh_colors),
  legend_gp = gpar(fill = gh_colors),
  ncol = 1
)
```

Row annotation
```{r}
ha_row <- rowAnnotation(
  GH = metadat_df$`GH family`,

  col = list(GH = gh_colors),

  border = TRUE, 
  
  show_annotation_name = FALSE,

  show_legend = FALSE,

  annotation_width = unit(3, "mm")
)
```

Plot the heatmap using viridis
```{r}
ht <- Heatmap(
  heatmap_scaled,

  width = unit(5, "cm"),
  
  name = "Z-score",

  col = viridis(100),

  top_annotation = ha_col,
  left_annotation = ha_row,

  cluster_rows = FALSE,
  cluster_columns = TRUE,

  row_split = metadat_df$`GH family`,

  row_title = NULL,

  show_row_names = TRUE,

  row_labels = gt_render(formatted_labels),

  row_names_gp = gpar(fontsize = 10),

  column_names_gp = gpar(fontsize = 10),

  border = TRUE
)
```

Draw heatmap
```{r}
draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(gh_legend)
)
```

Save heat map as a PDF
```{r}
pdf(
  "GH_family_heatmap_female_li.pdf",
  width = 12,
  height = 6
)

draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(gh_legend)
)

dev.off()
```

################################################################################
Arabinose isomerase visualization
################################################################################
```{r}
library(readxl)
library(tidyverse)
library(ggnewscale)
library(ggpubr)
```

Replicate data
```{r}
# Create data frame: one row per arabinose isomerase protein
arab_iso_wide <- data.frame(
  Protein = paste0("Arab_iso_", 1:4),
  
  LI_Con_1 = c(26.59014836, 22.86847361, 25.23974988, 22.86847361),
  LI_Con_2 = c(24.79358207, 24.47970457, 22.85565239, 21.33416101),
  LI_Con_3 = c(24.4231736,  22.78263954, 26.44463167, 24.22227051),
  LI_Con_4 = c(25.25789375, 23.64587049, 26.58626664, 23.34719595),
  
  LI_Strs_1 = c(16.66645673, 15.73025679, 18.50922405, 11.03458458),
  LI_Strs_2 = c(16.56972194, 16.60577156, 19.67905462, 13.64283085),
  LI_Strs_3 = c(17.08305329, 23.01987789, 21.02644166, 12.81229335),
  LI_Strs_4 = c(24.74865467, 15.79208467, 18.56204975, 12.69349201)
)

# Convert to long format
arab_iso_long <- arab_iso_wide %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Log2_Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress")),
    Protein = factor(Protein, levels = paste0("Arab_iso_", 1:4))
  )
```

Connected mean summary lines
```{r}
arab_iso_summary <- arab_iso_long %>%
  group_by(Protein, Group) %>%
  summarise(
    Mean_Log2_Intensity = mean(Log2_Intensity, na.rm = TRUE),
    .groups = "drop"
  )
```

Plot
```{r}
arab_iso_connected_plot <- ggplot() +
  
  # Replicate dots: original darker colors
  geom_jitter(
    data = arab_iso_long,
    aes(
      x = Group,
      y = Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    width = 0.16,
    size = 5,
    stroke = 0.4,
    alpha = 0.95
  ) +
  
    scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # Faint connected protein-level mean lines
  geom_line(
    data = arab_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      group = Protein
    ),
    color = "gray45",
    linewidth = 0.7,
    alpha = 0.5
  ) +
  
  # Mean points: lighter colors and drawn last
  geom_point(
    data = arab_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    size = 6.5,
    stroke = 0.7,
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  labs(
    title = "L-arabinose isomerase LI F",
    x = NULL,
    y = "Log2 intensity"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 25
    ),
    axis.text.x = element_text(
      color = "black",
      face = "bold",
      size = 23
    ),
    axis.text.y = element_text(
      color = "black",
      size = 23
    ),
    axis.title.y = element_text(
      color = "black",
      face = "bold",
      size = 25
    ),
    legend.position = "none"
  )

arab_iso_connected_plot
```

Save plot
```{r}
ggsave(
  filename = "arabinose_isomerase_plot.pdf",
  plot = arab_iso_connected_plot,
  width = 6,
  height = 6
)
```

################################################################################
Xylose isomerase visualization (bar graph)
################################################################################
```{r}
library(readxl)
library(tidyverse)
library(ggnewscale)
library(ggpubr)
```

Create data frame
```{r}
xylose_iso_wide <- data.frame(
  Protein = c(
    "A. muris",
    "A. chartisolvens"
  ),
  
  LI_Con_1 = c(22.76354405, 10.10437185),
  LI_Con_2 = c(22.85565239, 12.84485678),
  LI_Con_3 = c(21.9862123,  8.472315183),
  LI_Con_4 = c(21.98477853, 10.01343091),
  
  LI_Strs_1 = c(15.67163703, 22.76231867),
  LI_Strs_2 = c(16.64836058, 22.83457254),
  LI_Strs_3 = c(19.46889062, 23.24084031),
  LI_Strs_4 = c(16.59677043, 23.26018572)
)

# Convert to long format
xylose_iso_long <- xylose_iso_wide %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Log2_Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress")),
    Protein = factor(
      Protein,
      levels = c(
        "A. muris",
        "A. chartisolvens"
      )
    )
  )
```

Create a sig bracket dataframe
```{r}
bracket_df <- xylose_iso_long %>%
  group_by(Protein) %>%
  summarise(
    y.position = max(Log2_Intensity, na.rm = TRUE) + 1,
    .groups = "drop"
  ) %>%
  mutate(
    group1 = "Control",
    group2 = "Stress",
    label = c("**", "***")
  )
```

Plot
```{r}
xylose_barplot <- ggplot(
  xylose_iso_long,
  aes(x = Group, y = Log2_Intensity, fill = Group)
) +
  stat_summary(
    fun = mean,
    geom = "bar",
    width = 0.65,
    color = "black",
    linewidth = 0.6,
    alpha = 0.85
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.18,
    linewidth = 0.6,
    color = "black"
  ) +
  
  # Control replicate dots: darker gray with black border
  geom_jitter(
    data = xylose_iso_long %>% filter(Group == "Control"),
    aes(x = Group, y = Log2_Intensity),
    inherit.aes = FALSE,
    shape = 21,
    fill = "#4D4D4D",
    color = "black",
    stroke = 0.4,
    width = 0.08,
    size = 4.5,
    alpha = 0.95
  ) +
  
  # Stress replicate dots: darker orange with black border
  geom_jitter(
    data = xylose_iso_long %>% filter(Group == "Stress"),
    aes(x = Group, y = Log2_Intensity),
    inherit.aes = FALSE,
    shape = 21,
    fill = "#FE8A13",
    color = "black",
    stroke = 0.4,
    width = 0.08,
    size = 4.5,
    alpha = 0.95
  ) +
  
    stat_pvalue_manual(
    bracket_df,
    label = "label",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.6,
    size = 8,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(
    ~ Protein,
    scales = "free_y",
    nrow = 1
  ) +
  scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  labs(
    x = NULL,
    y = "Log2 intensity",
    title = "Xylose isomerase LI F"
  ) +
  theme_classic(base_size = 20) +
  theme(
    strip.background = element_rect(
      fill = "gray90",
      color = "black",
      linewidth = 0.5
    ),
    strip.text = element_text(
      face = "italic",
      color = "black",
      size = 18
    ),
    axis.text.x = element_text(
      color = "black",
      face = "bold",
      size = 23,
      hjust = 1,
      angle = 45
    ),
    axis.text.y = element_text(color = "black", size = 23),
    axis.title.y = element_text(color = "black", face = "bold", size = 25),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 25
    ),
    legend.position = "none",
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.6
    )
  )

xylose_barplot
```

Save plot
```{r}
ggsave(
  filename = "xylose_isomerase_plot.pdf",
  plot = xylose_barplot,
  width = 6,
  height = 6
)
```

################################################################################
Species-annotated volcano plot
################################################################################
```{r}
library(dplyr)
library(ggplot2)
library(stringr)
```

Load in the data
```{r}
vol_df <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "m2_plot")
```

Plot
```{r}
# Prepare dataframe
vol_plot_df <- vol_df %>%
  mutate(
    
    significant = ifelse(
      adj.P.Val < 0.05 & abs(logFC) > 2,
      "Significant",
      "Not Significant"
    ),
    
    # Assign plotting groups
    plot_group = case_when(
      Microbe == "Petralouisia muris" ~ "Petralouisia muris",
      str_detect(Microbe, regex("^Roseburia", ignore_case = TRUE)) ~ "Roseburia",
      str_detect(Microbe, regex("^Blautia", ignore_case = TRUE)) ~ "Blautia",
      str_detect(Microbe, regex("^Dorea", ignore_case = TRUE)) ~ "Dorea",
      str_detect(Microbe, regex("^Turicibacter", ignore_case = TRUE)) ~ "Turicibacter",
      significant == "Significant" ~ "Other significant",
      TRUE ~ "Not significant"
    ),
    
    neg_log10_p = -log10(P.Value)
  )

# Define plotting order
vol_plot_df$plot_group <- factor(
  vol_plot_df$plot_group,
  levels = c(
    "Not significant",
    "Other significant",
    "Petralouisia muris",
    "Roseburia",
    "Blautia",
    "Dorea",
    "Turicibacter"
  )
)

# IMPORTANT:
# Plot gray points first and highlighted taxa last
vol_plot_df <- vol_plot_df %>%
  arrange(plot_group)

# Color palette
microbe_colors <- c(
  "Not significant" = "grey85",
  "Other significant" = "grey40",
  
  "Petralouisia muris" = "#E76F51",   # coral-red
  "Roseburia" = "#2A9D8F",      # teal
  "Blautia" = "#3A86FF",        # vivid blue
  "Dorea" = "#F4A261",          # warm orange
  "Turicibacter" = "#9B5DE5"    # purple
)

# Volcano plot
volcano_plot <- ggplot(
  vol_plot_df,
  aes(
    x = logFC,
    y = neg_log10_p,
    color = plot_group
  )
) +
  
  # points
  geom_point(
    size = 2.5,
    alpha = 0.92,
    stroke = 0.3
  ) +
  
  # threshold lines
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    linewidth = 0.8,
    color = "black"
  ) +
  
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    linewidth = 0.8,
    color = "black"
  ) +
  
  # colors
  scale_color_manual(
    values = microbe_colors,
    
    breaks = c(
      "Petralouisia muris",
      "Roseburia",
      "Blautia",
      "Dorea",
      "Turicibacter",
      "Other significant",
      "Not significant"
    )
  ) +
  
  # labels
  labs(
    x = expression(Log[2]~Fold~Change),
    y = expression(-Log[10]~italic(P-value)),
    color = NULL
  ) +
  
  # axis scaling
  scale_y_continuous(
    expand = expansion(mult = c(0.02, 0.05))
  ) +
  
  scale_x_continuous(
    expand = expansion(mult = c(0.03, 0.03))
  ) +
  
  # publication theme
  theme_classic(base_size = 18) +
  
  theme(
    
    axis.text = element_text(
      size = 15,
      color = "black"
    ),
    
    axis.title = element_text(
      size = 18,
      face = "bold"
    ),
    
    legend.position = "right",
    
    legend.text = element_text(
      size = 13
    ),
    
    legend.key = element_blank(),
    
    legend.spacing.y = unit(0.2, "cm"),
    
    plot.margin = margin(
      t = 10,
      r = 15,
      b = 10,
      l = 10
    )
  )

# Display plot
volcano_plot
```

Save plot
```{r}
ggsave(
  "microbial_volcano_plot.pdf",
  volcano_plot,
  width = 8,
  height = 6,
  dpi = 600
)
```

################################################################################
Bray-Curtis dissimilarity plot - Functional redundancy analysis
################################################################################
```{r}
library(tidyverse)
library(vegan)        # Bray-Curtis, Jaccard
library(ggplot2)
library(ggVennDiagram)       # or ggVennDiagram
library(patchwork)    # multi-panel figures
library(RColorBrewer)
library(readxl)
library(ape)
```

Load in data
```{r}
norm_df <- read_excel("f_li_all_microbe_function_norm_vals.xlsx", sheet ="normalized_values")
# Define groups
con_samples <- c("Con_1", "Con_5", "Con_6", "Con_4")
str_samples <- c("Str_1", "Str_2", "Str_3", "Str_5")
# Convert to matrix
mat <- norm_df %>%
  select(-Protein, -Gene, -Functional_bin)
mat <- as.matrix(mat)
rownames(mat) <- norm_df$Protein
# Transpose
mat_t <- t(mat)
```

```{r}
# ── 0. Load & define groups ─────────────────────────────────
library(readxl); library(tidyverse); library(vegan)
library(ggVennDiagram); library(patchwork); library(ape)

df <- read_excel("f_li_all_microbe_function_norm_vals.xlsx",
                 sheet = "Carbohydrate Metabolism_GH")

con_samples <- c("Con_1", "Con_5", "Con_6", "Con_4")
str_samples <- c("Str_1", "Str_2", "Str_3", "Str_5")

# ── 1. Transformations ──────────────────────────────────────
df_linear <- df %>%
  mutate(across(all_of(c(con_samples, str_samples)),
                ~ ifelse(. > 0, 2^., 0)))

df_binary <- df %>%
  mutate(across(all_of(c(con_samples, str_samples)),
                ~ ifelse(. > 0, 1, 0)))


# ============================================================
# SECTION 1: WHOLE-COMMUNITY LEVEL
# ============================================================

# ── 1A. PCoA on Bray-Curtis dissimilarity ───────────────────
bc_mat <- df_linear %>%
  group_by(Function) %>%
  summarise(across(all_of(c(con_samples, str_samples)), sum)) %>%
  column_to_rownames("Function") %>%
  t()

bc_dist <- vegdist(bc_mat, method = "bray")

# Run PCoA (ape)
pcoa_res  <- pcoa(bc_dist)
pcoa_vals <- pcoa_res$vectors[, 1:2] %>%
  as.data.frame() %>%
  rownames_to_column("Sample") %>%
  rename(PC1 = Axis.1, PC2 = Axis.2) %>%
  mutate(Group = ifelse(Sample %in% con_samples, "Control", "Stress"))

# Percent variance explained
pct <- round(pcoa_res$values$Relative_eig[1:2] * 100, 1)

p_pcoa <- ggplot(pcoa_vals, aes(x = PC1, y = PC2,
                                color = Group, fill = Group,
                                label = Sample)) +
  stat_ellipse(geom = "polygon", alpha = 0.15, linetype = 2, level = 0.95) +
  geom_point(size = 4, shape = 21, color = "white", stroke = 0.5,
             aes(fill = Group)) +
  ggrepel::geom_text_repel(size = 3, show.legend = FALSE) +
  scale_color_manual(values = c("Control" = "#4393C3", "Stress" = "#D6604D")) +
  scale_fill_manual( values = c("Control" = "#4393C3", "Stress" = "#D6604D")) +
  labs(title = "Whole-community PCoA (Bray-Curtis)",
       x = paste0("PC1 [", pct[1], "% variance]"),
       y = paste0("PC2 [", pct[2], "% variance]")) +
  theme_classic() +
  theme(legend.position = "right")


# ── 1B. Jaccard overlap — Venn diagram ──────────────────────
con_functions <- df_binary %>%
  filter(rowSums(across(all_of(con_samples))) > 0) %>%
  pull(Function) %>% unique()

str_functions <- df_binary %>%
  filter(rowSums(across(all_of(str_samples))) > 0) %>%
  pull(Function) %>% unique()

venn_list <- list(Control = con_functions, Stress = str_functions)

p_venn <- ggVennDiagram(venn_list,
                        label = "count",
                        label_alpha = 0,
                        edge_size = 0.5) +
  scale_fill_gradient(low = "#FFFFFF", high = "#4393C3",
                      name = "Function\ncount") +
  scale_color_manual(values = c("Control" = "#4393C3",
                                "Stress"  = "#D6604D")) +
  labs(title = "Shared vs. Unique Functions (Presence/Absence)") +
  theme(legend.position = "right")


# ============================================================
# SECTION 2 & 3: USER-SELECTED FUNCTIONAL BINS
# ============================================================

# ▶▶ EDIT THIS VECTOR to choose which bins appear in the
#    barbell (FRS) plot and the heatmap. ◀◀
BINS_TO_PLOT <- c(
  "GH1", "GH13", "GH130", "GH2", "GH29", "GH3", "GH31", "GH32", "GH33", "GH36", "GH38", "GH4", "GH42", "GH43", "GH51", "GH53", "GH78", "GH92", "GH94",
  "Glycan/glycogen/starch metabolism",
  "Polysaccharide degradation",
  "Sugar utalization"
  # add or remove bin names here
)


# ── 2A. Per-bin Bray-Curtis (barbell) ───────────────────────
bins <- unique(df$Functional_bin)

bin_bc <- map_dfr(bins, function(b) {
  mat <- df_linear %>%
    filter(Functional_bin == b) %>%
    group_by(Function) %>%
    summarise(across(all_of(c(con_samples, str_samples)), sum)) %>%
    column_to_rownames("Function") %>%
    t()

  if (nrow(mat) < 1 || ncol(mat) < 2) return(NULL)

  dist_mat <- as.matrix(vegdist(mat, method = "bray"))
  between  <- dist_mat[con_samples, str_samples]
  tibble(Functional_bin = b,
         BC_between = mean(between, na.rm = TRUE))
}) %>%
  filter(Functional_bin %in% BINS_TO_PLOT)   # ← filter here

p_bin_bc <- ggplot(bin_bc %>% arrange(BC_between),
                   aes(x = BC_between,
                       y = reorder(Functional_bin, BC_between))) +
  geom_segment(aes(x = 0, xend = BC_between,
                   yend = Functional_bin), color = "grey70") +
  geom_point(size = 3, color = "#D6604D") +
  labs(title = "Between-group Bray-Curtis per Functional Bin",
       x = "Mean Bray-Curtis (Between Groups)", y = NULL) +
  theme_classic()


# ── 2B. Functional Redundancy Score (FRS) ───────────────────
total_taxa <- n_distinct(df$Microbe)

frs_df <- df_binary %>%
  pivot_longer(all_of(c(con_samples, str_samples)),
               names_to = "Sample", values_to = "Present") %>%
  mutate(Group = ifelse(Sample %in% con_samples, "Control", "Stress")) %>%
  group_by(Functional_bin, Group, Microbe) %>%
  summarise(detected = max(Present), .groups = "drop") %>%
  group_by(Functional_bin, Group) %>%
  summarise(
    n_taxa_with_function = sum(detected),
    FRS = n_taxa_with_function / total_taxa,
    .groups = "drop"
  ) %>%
  filter(Functional_bin %in% BINS_TO_PLOT)   # ← filter here

p_frs <- ggplot(frs_df,
                aes(x = FRS, y = reorder(Functional_bin, FRS),
                    color = Group, shape = Group)) +
  geom_line(aes(group = Functional_bin), color = "grey70") +
  geom_point(size = 3.5) +
  scale_color_manual(values = c("Control" = "#4393C3",
                                "Stress"  = "#D6604D")) +
  labs(title = "Functional Redundancy Score per Bin",
       x = "FRS (proportion of taxa)", y = NULL) +
  theme_classic()


# ============================================================
# SECTION 3: ABUNDANCE HEATMAP (bin × group)
# ============================================================

heatmap_df <- df_linear %>%
  filter(Functional_bin %in% BINS_TO_PLOT) %>%   # ← filter here
  pivot_longer(all_of(c(con_samples, str_samples)),
               names_to = "Sample", values_to = "Abundance") %>%
  mutate(Group = ifelse(Sample %in% con_samples, "Control", "Stress")) %>%
  group_by(Functional_bin, Group) %>%
  summarise(Mean_Abundance = mean(Abundance), .groups = "drop")

p_heat <- ggplot(heatmap_df,
                 aes(x = Group, y = Functional_bin,
                     fill = log1p(Mean_Abundance))) +
  geom_tile(color = "white") +
  scale_fill_distiller(palette = "YlOrRd", direction = 1,
                       name = "log(mean\nabundance)") +
  labs(title = "Mean Abundance by Functional Bin", x = NULL, y = NULL) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 0))


# ============================================================
# SECTION 4: COMPOSE FINAL FIGURE
# ============================================================

(p_venn | p_pcoa) /
(p_frs  | p_heat) +
  plot_annotation(tag_levels = "A")
```

Bray-Curtis
```{r}
bray_dist <- vegdist(mat_t, method = "bray")
```

Set groups
```{r}
groups_bc <- factor(c("Control","Control","Control", "Control", "Stress","Stress", "Stress", "Stress"))
```

PERMANOVA
```{r}
permanova_res <- adonis2(bray_dist ~ groups_bc)

# Extract p-value
pval <- permanova_res$`Pr(>F)`[1]
```

PCoA visualization
```{r}
pcoa <- cmdscale(bray_dist, eig = TRUE, k = 2)

# Percent variance explained
eig_percent <- round(
  100 * pcoa$eig / sum(pcoa$eig[pcoa$eig > 0]),
  1
)

# Build plotting dataframe
pcoa_df <- data.frame(
  Axis1 = pcoa$points[,1],
  Axis2 = pcoa$points[,2],
  Group = groups_bc,
  Sample = rownames(pcoa$points)
)
```

Aesthetics
```{r}
# Colors
fill_colors <- c(
  "Control" = "#4D4D4D",
  "Stress"  = "#FE8A13"
)

# Significance label
sig_label <- ifelse(
  pval < 0.001, "***",
  ifelse(
    pval < 0.01, "**",
    ifelse(
      pval < 0.05, "*", "ns"
    )
  )
)

# Text color
text_color <- ifelse(pval < 0.05, "black", "gray50")
```

Plot
```{r}
p_bray <- ggplot(
  pcoa_df,
  aes(x = Axis1, y = Axis2)
) +

  # Ellipse shading
  stat_ellipse(
    aes(group = Group, fill = Group),
    geom = "polygon",
    alpha = 0.25,
    color = NA,
    level = 0.8,
    type = "norm"
  ) +

  # Sample points
  geom_point(
    aes(fill = Group),
    shape = 21,
    size = 5,
    color = "gray50",
    alpha = 1
  ) +

  # Optional sample labels
  geom_text(
    aes(label = Sample),
    vjust = -1,
    size = 3
  ) +

  # P-value annotation
  annotate(
    "text",
    x = Inf,
    y = Inf,
    label = paste0(
      "PERMANOVA, ",
      sig_label,
      "  p = ",
      signif(pval, 3)
    ),
    hjust = 1.1,
    vjust = 1.5,
    size = 4,
    fontface = "bold",
    color = text_color
  ) +

  # Colors
  scale_fill_manual(values = fill_colors) +

  # Labels
  labs(
    title = "Female Large Intestine Metaproteome\nPCoA (Bray–Curtis)",
    x = paste0("PCoA 1 (", eig_percent[1], "%)"),
    y = paste0("PCoA 2 (", eig_percent[2], "%)")
  ) +

  # Theme
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(
      size = 14,
      face = "bold",
      hjust = 0.5
    ),
    axis.text = element_text(size = 12),
    axis.title = element_text(
      size = 12,
      face = "bold"
    ),
    legend.title = element_blank(),
    legend.text = element_text(size = 11),

    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 1
    )
  )

# View plot
p_bray

# Save plot
ggsave("bray_curtis_pcoa.pdf", p_bray,
  width = 7,
  height = 5
)
```

################################################################################
Mirrored diverging bar plot
################################################################################
```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
```

Create data frame
```{r}
# Input: one row per genus, separate columns for increased and decreased counts
df_microbe <- data.frame(
  genus = c("Petralouisia", "Blautia", "Roseburia",
            "Dorea", "Turicibacter", "Acutalibacter",
            "Acetatifactor", "Romboutsia", "Clostridium",
            "Eubacterium", "Ruminococcus", "Sporofaciens"),
  increased = c(49, 19, 14, 4, 1, 0, 5, 2, 5, 6, 7, 1),
  decreased = c(3, 25, 20, 29, 28, 22, 15, 18, 14, 12, 9, 13)
)
```

Mirror: decreased becomes negative; sort genera by net change
```{r}
df_microbe <- df_microbe %>%
  mutate(
    net = increased - decreased,
    genus = reorder(genus, net),
    decreased_mirror = -decreased
  ) %>%
  pivot_longer(
    cols = c(increased, decreased_mirror),
    names_to = "direction",
    values_to = "count"
  ) %>%
  mutate(direction = recode(direction,
    "increased" = "Increased",
    "decreased_mirror" = "Decreased"
  ))

# Export protein count data frame
write_xlsx(df_microbe, "genus_protein_count.xlsx")
```

Plot diverging bar graph
```{r}
microbe_div_bar <- ggplot(df_microbe, aes(x = count, y = genus, fill = direction)) +
  geom_col(
    width = 0.7,
    color = "black",
    linewidth = 0.4
  ) +
  geom_vline(xintercept = 0, linewidth = 0.6, color = "grey40") +
  scale_fill_manual(values = c("Increased" = "#FE8A13", "Decreased" = "#4D4D4D")) +
  scale_x_continuous(
    labels = function(x) abs(x),     # show absolute values on axis
    expand = expansion(mult = 0.05)
  ) +
  labs(
    x = "Number of differentially abundant proteins",
    y = NULL,
    fill = NULL,
    title = "Changes in microbial protein abundance - Top 12 genera by total protein count"
  ) +
  theme_minimal(base_size = 23) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "italic", size = 20),
    axis.text.x = element_text(size = 18),
    axis.title.x = element_text(face = "bold", size = 20),
    plot.title = element_text(face = "bold", size = 20)
  )
microbe_div_bar
```

Save plot
```{r}
ggsave(
  "microbial_diverging_barplot.pdf",
  microbe_div_bar,
  width = 10,
  height = 6,
  dpi = 600
)
```

################################################################################
Functional category diverging barplots
################################################################################
 ── Data ──────────────────────────────────────────────────────────────────────
```{r}
cats <- c("Carbohydrate metabolism","Lipid / SCFA metabolism",
          "Propanediol utilization","Translation / ribosome",
          "Amino acid metabolism","Transport / binding",
          "Energy / electron transfer","Cell motility / division",
          "Cofactor metabolism","Stress response / chaperones",
          "Nucleotide metabolism","Uncharacterized")

df_func <- tribble(
  ~category,                      ~genus,           ~increased, ~decreased,
  "Carbohydrate metabolism",      "Petralouisia",   15, 1,
  "Lipid / SCFA metabolism",      "Petralouisia",    4, 0,
  "Propanediol utilization",      "Petralouisia",    5, 0,
  "Translation / ribosome",       "Petralouisia",    6, 0,
  "Amino acid metabolism",        "Petralouisia",    4, 1,
  "Transport / binding",          "Petralouisia",    3, 0,
  "Energy / electron transfer",   "Petralouisia",    5, 0,
  "Cell motility / division",     "Petralouisia",    2, 0,
  "Cofactor metabolism",          "Petralouisia",    1, 0,
  "Stress response / chaperones", "Petralouisia",    0, 0,
  "Nucleotide metabolism",        "Petralouisia",    0, 0,
  "Uncharacterized",              "Petralouisia",    4, 1,

  "Carbohydrate metabolism",      "Turicibacter",    0, 6,
  "Lipid / SCFA metabolism",      "Turicibacter",    0, 1,
  "Propanediol utilization",      "Turicibacter",    0, 0,
  "Translation / ribosome",       "Turicibacter",    0, 7,
  "Amino acid metabolism",        "Turicibacter",    0, 0,
  "Transport / binding",          "Turicibacter",    0, 1,
  "Energy / electron transfer",   "Turicibacter",    0, 0,
  "Cell motility / division",     "Turicibacter",    0, 1,
  "Cofactor metabolism",          "Turicibacter",    0, 1,
  "Stress response / chaperones", "Turicibacter",    0, 1,
  "Nucleotide metabolism",        "Turicibacter",    0, 1,
  "Uncharacterized",              "Turicibacter",    1, 3,

  "Carbohydrate metabolism",      "Dorea",          1, 13,
  "Lipid / SCFA metabolism",      "Dorea",          0,  0,
  "Propanediol utilization",      "Dorea",          0,  0,
  "Translation / ribosome",       "Dorea",          0,  5,
  "Amino acid metabolism",        "Dorea",          1,  2,
  "Transport / binding",          "Dorea",          0,  0,
  "Energy / electron transfer",   "Dorea",          1,  2,
  "Cell motility / division",     "Dorea",          0,  1,
  "Cofactor metabolism",          "Dorea",          0,  0,
  "Stress response / chaperones", "Dorea",          0,  1,
  "Nucleotide metabolism",        "Dorea",          0,  0,
  "Uncharacterized",              "Dorea",          0,  1,

  "Carbohydrate metabolism",      "Blautia",        8,  9,
  "Lipid / SCFA metabolism",      "Blautia",        0,  2,
  "Propanediol utilization",      "Blautia",        0,  0,
  "Translation / ribosome",       "Blautia",        3,  6,
  "Amino acid metabolism",        "Blautia",        1,  2,
  "Transport / binding",          "Blautia",        1,  0,
  "Energy / electron transfer",   "Blautia",        1,  1,
  "Cell motility / division",     "Blautia",        0,  0,
  "Cofactor metabolism",          "Blautia",        0,  1,
  "Stress response / chaperones", "Blautia",        0,  1,
  "Nucleotide metabolism",        "Blautia",        1,  0,
  "Uncharacterized",              "Blautia",        2,  1,
  
  "Carbohydrate metabolism",      "Roseburia",        2,  7,
  "Lipid / SCFA metabolism",      "Roseburia",        1,  1,
  "Propanediol utilization",      "Roseburia",        0,  0,
  "Translation / ribosome",       "Roseburia",        4,  4,
  "Amino acid metabolism",        "Roseburia",        1,  1,
  "Transport / binding",          "Roseburia",        1,  0,
  "Energy / electron transfer",   "Roseburia",        0,  1,
  "Cell motility / division",     "Roseburia",        0,  0,
  "Cofactor metabolism",          "Roseburia",        1,  0,
  "Stress response / chaperones", "Roseburia",        0,  0,
  "Nucleotide metabolism",        "Roseburia",        1,  1,
  "Uncharacterized",              "Roseburia",        1,  0
)
```

 ── Reshape to long format, mirror decreased ──────────────────────────────────
```{r}
df_func_long <- df_func %>%
  mutate(category = factor(category, levels = rev(cats))) %>%
  pivot_longer(c(increased, decreased), names_to = "direction", values_to = "n") %>%
  mutate(
    n_mirror = if_else(direction == "decreased", -n, n),
    direction = factor(direction, levels = c("increased", "decreased"))
  )

# ── Facet label formatting ────────────────────────────────────────────────────
genus_labels <- c(
  Petralouisia = "Petralouisia",
  Turicibacter = "Turicibacter",
  Dorea        = "Dorea",
  Blautia      = "Blautia",
  Roseburia    = "Roseburia"
)

df_func_long$genus <- factor(df_func_long$genus,
  levels = c("Petralouisia","Turicibacter","Dorea","Blautia", "Roseburia"))
```

 ── Plot ──────────────────────────────────────────────────────────────────────
```{r}
pal <- c("increased" = "#FE8A13", "decreased" = "#4D4D4D")

func_plot <- ggplot(df_func_long, aes(x = n_mirror, y = category, fill = direction)) +
  geom_col(width = 0.7, colour = "black", linewidth = 0.2) +
  geom_vline(xintercept = 0, linewidth = 0.5, colour = "grey40") +
  scale_fill_manual(
    values = pal,
    labels = c("Increased in stress", "Decreased in stress")
  ) +
  scale_x_continuous(
    labels = abs,
    expand = expansion(mult = 0.05)
  ) +
  facet_wrap(~ genus, nrow = 1, labeller = labeller(genus = genus_labels)) +
  labs(
    x    = "Number of differentially abundant proteins",
    y    = NULL,
    fill = NULL,
    title    = "Functional protein abundance changes across selected gut microbiota",
    subtitle = "Stress vs. Control — Female large intestine metaproteome"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position   = "top",
    strip.text        = element_text(face = "italic", size = 11),
    panel.grid.major.y = element_blank(),
    panel.grid.minor   = element_blank(),
    axis.text.y        = element_text(size = 9),
    panel.spacing.x    = unit(1, "lines")
  )
```

 Save plot
```{r}
ggsave(
  "function_diverging_barplot.pdf",
  func_plot,
  width = 16,
  height = 8,
  dpi = 600
)
```

################################################################################
Functional redundancy score and plotting
################################################################################
# ── Compute FRS ───────────────────────────────────────────────────────────────
```{r}
frs_data <- df_func %>%
  select(category, genus, increased, decreased) %>%
  mutate(total = increased + decreased)

petra_df <- frs_data %>%
  filter(genus == "Petralouisia") %>%
  select(category, petra_up = increased, petra_total = total)

frs <- frs_data %>%
  filter(genus != "Petralouisia") %>%
  left_join(petra_df, by = "category") %>%
  mutate(
    # Directional replacement: Petra gaining where genus is losing
    replacement_k = pmin(petra_up, decreased),
    denom_k       = pmax(petra_total, total),
    frs_k         = if_else(denom_k > 0, replacement_k / denom_k, 0),
    # Category overlap (either genus has proteins here)
    both_active   = (petra_total > 0) & (total > 0)
  ) %>%
  group_by(genus) %>%
  summarise(
    FRS          = round(sum(replacement_k) / sum(denom_k), 3),
    n_categories_overlapping = sum(both_active),
    total_replacement_proteins = sum(replacement_k),
    .groups = "drop"
  ) %>%
  arrange(desc(FRS))

print(frs)
```
 
 Bubble plot to display FRS calculation
```{r}
# Per-category FRS (pre-summarise step from above, before group_by collapse)
frs_cat <- frs_data %>%
  filter(genus != "Petralouisia") %>%
  left_join(petra_df, by = "category") %>%
  mutate(
    replacement_k = pmin(petra_up, decreased),
    denom_k       = pmax(petra_total, total),
    frs_k         = if_else(denom_k > 0, replacement_k / denom_k, 0)
  ) %>%
  filter(replacement_k > 0) %>%   # only show categories with actual replacement
  mutate(
    category = factor(category, levels = rev(cats)),
    genus    = factor(genus, levels = c("Turicibacter","Dorea","Blautia", "Roseburia"))
  )

genus_pal <- c(
  "Turicibacter" = "#D85A30",
  "Dorea"        = "#7F77DD",
  "Blautia"      = "#378ADD",
  "Roseburia"    = "green4"
)

ggplot(frs_cat, aes(x = frs_k, y = category,
                    size = replacement_k, colour = genus)) +
  geom_point(alpha = 0.8) +
  scale_size_continuous(range = c(3, 12), name = "Replacement\nproteins (n)") +
  scale_colour_manual(values = genus_pal, name = NULL) +
  scale_x_continuous(limits = c(0, 1), labels = scales::percent_format()) +
  labs(
    x        = "Functional Replacement Score (per category)",
    y        = NULL,
    title    = "Functional replacement of declining genera by Petralouisia muris",
    subtitle = "Score = overlap of Petralouisia gains with genus losses, normalised by category size"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    legend.position    = "right",
    panel.grid.major.x = element_line(linewidth = 0.3, colour = "grey85"),
    panel.grid.major.y = element_blank(),
    axis.text.y        = element_text(size = 9)
  )
```
 
################################################################################
Box and whisker plots for individual proteins
################################################################################
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
```

Load in data
```{r}
protein_df <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "p_muris_pro")

# Convert from wide to long format
protein_long <- protein_df %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```

Choose proteins to plot
```{r}
# This order controls the order of the boxplots/facets
proteins_to_plot <- c(
  "WP_135987631.1",
  "WP_135989817.1",
  "WP_135989873.1",
  "WP_135989816.1",
  "WP_135989815.1",
  "WP_135989814.1",
  "WP_135987782.1",
  "WP_135987616.1"
)

plot_df <- protein_long %>%
  filter(Protein %in% proteins_to_plot) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot)
  )
```

Create ordered facet labels
```{r}
facet_order_df <- protein_df %>%
  filter(Protein %in% proteins_to_plot) %>%
  distinct(Protein, Function) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot),
    Plot_Label = Function
  ) %>%
  arrange(Protein)

plot_label_order <- facet_order_df$Plot_Label

plot_df <- plot_df %>%
  left_join(facet_order_df, by = c("Protein", "Function")) %>%
  mutate(
    Plot_Label = factor(Plot_Label, levels = plot_label_order)
  )
```

Significance labels and brackets
```{r}
sig_df <- protein_df %>%
  filter(Protein %in% proteins_to_plot) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot),
    sig_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  select(Protein, Function, adj.P.Val, sig_label)

bracket_df <- plot_df %>%
  group_by(Protein, Function, Plot_Label) %>%
  summarise(
    y.position = max(Intensity, na.rm = TRUE) + 
      0.12 * diff(range(Intensity, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(sig_df, by = c("Protein", "Function")) %>%
  mutate(
    group1 = "Control",
    group2 = "Stress",
    Plot_Label = factor(Plot_Label, levels = plot_label_order)
  )
```

Plot
```{r}
protein_boxplot <- ggplot(
  plot_df,
  aes(
    x = Group,
    y = Intensity,
    fill = Group
  )
) +
  
  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.7,
    alpha = 0.85
  ) +
  
  geom_jitter(
    aes(fill = Group),
    shape = 21,
    color = "black",
    width = 0.12,
    size = 3.5,
    stroke = 0.5,
    alpha = 0.9
  ) +
  
  stat_pvalue_manual(
    bracket_df,
    label = "sig_label",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.8,
    size = 9,
    hide.ns = FALSE,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(
    ~ Plot_Label,
    scales = "free_y",
    ncol = 2,
    
    # Automatically wrap long facet labels
    labeller = labeller(
      Plot_Label = label_wrap_gen(
        width = 32,
        multi_line = TRUE
      )
    )
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  scale_color_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  # Leave room for brackets and significance labels
  scale_y_continuous(
    expand = expansion(
      mult = c(0.05, 0.20)
    )
  ) +
  
  labs(
    x = NULL,
    y = "Log2-normalized protein intensity",
    title = paste(
      "Propanediol metabolism and associated proteins",
      "from P. muris"
    )
  ) +
  
  theme_classic(base_size = 20) +
  
  theme(
    # Main title
    plot.title = element_text(
      size = 25,
      face = "bold",
      hjust = 0.5,
      lineheight = 1.05,
      margin = margin(b = 15)
    ),
    
    # Gray facet ribbons
    strip.background = element_rect(
      fill = "grey90",
      color = "black",
      linewidth = 0.7
    ),
    
    # Wrapped facet text
    strip.text = element_text(
      color = "black",
      face = "bold",
      size = 17,
      lineheight = 1,
      margin = margin(
        t = 9,
        r = 8,
        b = 9,
        l = 8
      )
    ),
    
    # Axis text
    axis.text.x = element_text(
      color = "black",
      size = 23,
      face = "bold"
    ),
    
    axis.text.y = element_text(
      color = "black",
      size = 20
    ),
    
    # Axis title
    axis.title.y = element_text(
      color = "black",
      size = 23,
      face = "bold",
      margin = margin(r = 12)
    ),
    
    # Legend text
    legend.position = "top",
    legend.title = element_text(
      size = 23,
      face = "bold"
    ),
    legend.text = element_text(
      size = 23,
      face = "bold"
    ),
    
    # Spacing between facets
    panel.spacing = grid::unit(1.2, "lines")
  )

protein_boxplot
```

Save plots
```{r}
ggsave(
  filename = "PM_protein_boxplots2.pdf",
  width = 10,
  height = 18,
  dpi = 300
)
```

################################################################################
Selected heat map
################################################################################
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
library(RColorBrewer)
library(viridis)
library(ComplexHeatmap)
```

Load in data
```{r}
protein_df <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "p_muris_pro")
```

Define sample columns
```{r}
sample_cols <- c(
  "LI_Con_1", "LI_Con_2", "LI_Con_3", "LI_Con_4",
  "LI_Strs_1", "LI_Strs_2", "LI_Strs_3", "LI_Strs_4"
)

control_cols <- c("LI_Con_1", "LI_Con_2", "LI_Con_3", "LI_Con_4")
stress_cols  <- c("LI_Strs_1", "LI_Strs_2", "LI_Strs_3", "LI_Strs_4")
```

Filter by desired proteins
```{r}
proteins_to_plot <- c(
  "WP_135987631.1",
  "WP_135989817.1",
  "WP_135989816.1",
  "WP_135989815.1",
  "WP_135989873.1",
  "WP_135989814.1"
)

pdu_df <- protein_df %>%
  filter(Protein %in% proteins_to_plot)
```

Clean labels and add significance
```{r}
pdu_df <- pdu_df %>%
  mutate(
    Function_clean = case_when(
      str_detect(Function, regex("L-fucose/L-arabinose isomerase family protein", ignore_case = TRUE)) ~
        "L-fucose/L-arabinose isomerase",
      str_detect(Function, regex("microcompartment protein PduB", ignore_case = TRUE)) ~
        "PduB microcompartment protein",
      str_detect(Function, regex("propanediol/glycerol family dehydratase large", ignore_case = TRUE)) ~
        "PduC-like dehydratase large subunit",
      str_detect(Function, regex("propanediol/glycerol family dehydratase medium", ignore_case = TRUE)) ~
        "PduD-like dehydratase medium subunit",
      str_detect(Function, regex("diol dehydratase small", ignore_case = TRUE)) ~
        "PduE-like dehydratase small subunit",
      str_detect(Function, regex("BMC domain", ignore_case = TRUE)) ~
        "BMC domain-containing protein",
      TRUE ~ Function
    ),
    
    sig_label = case_when(
      adj.P.Val < 0.05  ~ "*",
      TRUE ~ ""
    ),
    
    mean_intensity = rowMeans(across(all_of(sample_cols)), na.rm = TRUE),
    
    row_label = paste0(Function_clean, "\n", Protein)
  )
```

Set the pathway order
```{r}
pathway_order <- c(
  "L-fucose/L-arabinose isomerase",
  "PduB microcompartment protein",
  "BMC domain-containing protein",
  "PduC-like dehydratase large subunit",
  "PduD-like dehydratase medium subunit",
  "PduE-like dehydratase small subunit"
)

pdu_df <- pdu_df %>%
  mutate(Function_clean = factor(Function_clean, levels = pathway_order)) %>%
  arrange(Function_clean)
```

Create the protein intensity matrix for plotting
```{r}
heatmap_matrix <- pdu_df %>%
  select(row_label, all_of(sample_cols)) %>%
  column_to_rownames("row_label") %>%
  as.matrix()

mode(heatmap_matrix) <- "numeric"
```

Set colors
```{r}
# --- Log2FC matrix ---
logfc_matrix <- matrix(
  pdu_df$logFC,
  ncol = 1,
  dimnames = list(rownames(heatmap_matrix), "Log2FC")
)

# --- Color scales ---

# Protein intensity: full viridis gradient
intensity_col_fun <- colorRamp2(
  seq(min(heatmap_matrix, na.rm = TRUE),
      max(heatmap_matrix, na.rm = TRUE),
      length.out = 100),
  viridis(100, option = "viridis")
)

# Log2FC red gradient anchored correctly at 0
# Handles both all-positive and mixed-sign logFC values
logfc_min <- min(pdu_df$logFC, na.rm = TRUE)
logfc_max <- max(pdu_df$logFC, na.rm = TRUE)

if (logfc_min >= 0) {
  # All positive: white at 0, deep red at max
  logfc_col_fun <- colorRamp2(
    c(0, logfc_max),
    c("white", "#CC0000")
  )
} else {
  # Mixed: symmetric diverging scale, white at 0
  abs_max <- max(abs(logfc_min), abs(logfc_max))
  logfc_col_fun <- colorRamp2(
    c(-abs_max, 0, abs_max),
    c("#0033CC", "white", "#CC0000")
  )
}

# --- Column annotation (top) ---
column_group <- data.frame(
  Group = c(
    rep("Control", length(control_cols)),
    rep("Stress", length(stress_cols))
  )
)
rownames(column_group) <- sample_cols

group_colors <- c(
  "Control" = "#4D4D4D",
  "Stress"  = "#FE8A13"
)

top_anno <- HeatmapAnnotation(
  Group = column_group$Group,
  col = list(Group = group_colors),
  annotation_name_side = "left",
  annotation_legend_param = list(
    Group = list(title = "Group")
  )
)

# --- Main intensity heatmap ---
# row_split on seq_len gives each row its own gap; row_title = NULL suppresses
# the per-row title labels that row_split would otherwise generate
main_heatmap <- Heatmap(
  heatmap_matrix,

  name = "Protein\nintensity",
  col  = intensity_col_fun,

  top_annotation = top_anno,

  cluster_rows    = FALSE,
  cluster_columns = FALSE,

  row_split   = seq_len(nrow(heatmap_matrix)),
  row_title   = NULL,
  row_gap     = unit(2, "mm"),

  show_row_names    = TRUE,
  show_column_names = TRUE,

  row_names_side   = "left",
  row_names_gp     = gpar(fontsize = 8),
  column_names_gp  = gpar(fontsize = 9),
  column_names_rot = 45,

  column_split = column_group$Group,

  column_title = "LI metaproteomic protein intensities",

  heatmap_legend_param = list(
    title         = "Protein\nintensity",
    legend_height = unit(4, "cm")
  ),

  border  = TRUE,
  rect_gp = gpar(col = "black", lwd = 0.6)
)

# --- Log2FC heatmap ---
logfc_heatmap <- Heatmap(
  logfc_matrix,

  name = "Log2\nFold Change",
  col  = logfc_col_fun,

  cluster_rows    = FALSE,
  cluster_columns = FALSE,

  row_split = seq_len(nrow(logfc_matrix)),
  row_title = NULL,
  row_gap   = unit(2, "mm"),

  show_row_names    = FALSE,
  show_column_names = TRUE,

  column_names_gp  = gpar(fontsize = 9),
  column_names_rot = 45,

  width = unit(0.5, "cm"),

  cell_fun = function(j, i, x, y, width, height, fill) {
    label <- pdu_df$sig_label[i]
    if (nchar(label) > 0) {
      grid.text(
        label,
        x = x, y = y,
        gp = gpar(fontsize = 11, fontface = "bold", col = "black")
      )
    }
  },

  heatmap_legend_param = list(
    title         = "Log2\nFold Change",
    legend_height = unit(4, "cm")
  ),

  border  = TRUE,
  rect_gp = gpar(col = "black", lwd = 0.6)
)

# --- Draw ---
draw(
  main_heatmap + logfc_heatmap,
  heatmap_legend_side    = "right",
  annotation_legend_side = "right"
)
```

Save heatmap
```{r}
pdf("pdu_heatmap.pdf", width = 8, height = 6)

draw(
  main_heatmap + logfc_heatmap,
  heatmap_legend_side    = "right",
  annotation_legend_side = "right"
)

dev.off()
```

============================================================
Relative Abundance Bar Chart: Functional Categories
Stress (Upregulated) vs Control (Downregulated) Groups
============================================================

 --- Load required libraries ---
```{r}
library(ggplot2)
library(readxl)
library(dplyr)
library(tidyr)
library(scales)
library(RColorBrewer)
library(rcartocolor)
```


 ============================================================
 1. DATA INPUT
 ============================================================

```{r}
data <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "carb_meta")
```


 ============================================================
 2. ASSIGN GROUPS
    Upregulated   -> "Stress"
    Downregulated -> "Control"
 ============================================================

```{r}
data <- data %>%
  mutate(
    Group = case_when(
      regulation == "Up"   ~ "Stress",
      regulation == "Down" ~ "Control",
      TRUE                 ~ NA_character_
    )
  ) %>%
  filter(!is.na(Group))
```

 ============================================================
 3. COUNT PROTEINS PER FUNCTIONAL CATEGORY × GROUP
 ============================================================

```{r}
counts <- data %>%
  group_by(Group, Functional_bin) %>%
  summarise(n = n(), .groups = "drop")

# Calculate relative abundance (%) within each group
counts <- counts %>%
  group_by(Group) %>%
  mutate(
    total      = sum(n),
    rel_abund  = n / total * 100
  ) %>%
  ungroup()

# Order categories by total count (descending) for a tidy y-axis
cat_order <- counts %>%
  group_by(Functional_bin) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(total_n) %>%
  pull(Functional_bin)

counts <- counts %>%
  mutate(Functional_bin = factor(Functional_bin, levels = cat_order))
```

 ============================================================
 4. COLOUR PALETTE
    Stress  = orange  (#FE8A13)
    Control = gray  (#4D4D4D)
 ============================================================

```{r}
group_colours <- c("Stress"  = "#FE8A13",
                   "Control" = "#4D4D4D")
```

 ============================================================
 5. PLOT
 ============================================================

```{r}
p <- ggplot(counts,
            aes(x    = rel_abund,
                y    = Functional_bin,
                fill = Group)) +

  # Dodged horizontal bars
  geom_col(
    position = position_dodge(width = 0.75),
    width    = 0.68,
    colour   = "white",
    linewidth = 0.25
  ) +

  # Percentage labels at bar ends
  geom_text(
    aes(label = sprintf("%.1f%%", rel_abund)),
    position  = position_dodge(width = 0.75),
    hjust     = -0.15,
    size      = 2.8,
    colour    = "grey35",
    fontface  = "plain"
  ) +

  # Colour scale
  scale_fill_manual(
    values = group_colours,
    labels = c("Downregulated proteins", "Upregulated proteins")
  ) +

  # X-axis: percentage, add a little right padding for labels
  scale_x_continuous(
    labels   = label_percent(scale = 1, accuracy = 1),
    expand   = expansion(mult = c(0, 0.18))
  ) +

  # Labels
  labs(
    title    = "Relative Abundance of Functional Categories",
    subtitle = "Female Large Intestine | 100 Downregulated proteins | 50 Upregulated proteins",
    x        = "Relative Abundance (%)",
    y        = NULL,
    fill     = "Group"
  ) +

  # Theme
  theme_minimal(base_size = 12) +
  theme(
    # Panel
    panel.grid.major.y  = element_blank(),
    panel.grid.major.x  = element_line(colour = "grey90", linewidth = 0.4),
    panel.grid.minor    = element_blank(),
    panel.border        = element_blank(),
    axis.line.x         = element_line(colour = "grey70", linewidth = 0.4),

    # Axis text
    axis.text.y         = element_text(size = 10, colour = "grey20"),
    axis.text.x         = element_text(size = 9,  colour = "grey40"),
    axis.title.x        = element_text(size = 10, colour = "grey30",
                                       margin = margin(t = 8)),

    # Title
    plot.title          = element_text(size = 14, face = "bold",
                                       colour = "grey10",
                                       margin = margin(b = 4)),
    plot.subtitle       = element_text(size = 9,  colour = "grey45",
                                       margin = margin(b = 12)),
    plot.margin         = margin(14, 20, 14, 14),

    # Legend
    legend.position     = "top",
    legend.justification = "left",
    legend.title        = element_text(size = 9, face = "bold"),
    legend.text         = element_text(size = 9),
    legend.key.size     = unit(0.55, "cm"),
    legend.margin       = margin(b = 6)
  )

print(p)
```

 ============================================================
 6. SAVE
 ============================================================

```{r}
ggsave(
  filename = "carbohydrate_relative_abundance_bar_chart.pdf",
  plot     = p,
  width    = 10,
  height   = 7,
  device   = pdf   # crisp text; swap for "pdf" if cairo unavailable
)
```

Create a relative abundance stacked barplot
 ============================================================
 Following steps 1-2 above
 3. COUNT PROTEINS & CALCULATE PROPORTIONS
 ============================================================
 
```{r}
counts <- data %>%
  group_by(Group, Functional_bin) %>%
  summarise(n = n(), .groups = "drop") %>%
  group_by(Group) %>%
  mutate(
    total = sum(n),
    prop  = n / total
  ) %>%
  ungroup()

# Desired bottom category
bottom_cat <- "Poor characterization"

# Order categories by total abundance across both groups
cat_levels_by_abundance <- counts %>%
  group_by(Functional_bin) %>%
  summarise(total_n = sum(n), .groups = "drop") %>%
  arrange(desc(total_n)) %>%
  pull(Functional_bin)

# Remove Poor characterization from abundance-based order
cat_levels_by_abundance <- setdiff(cat_levels_by_abundance, bottom_cat)

# Put Poor characterization at the bottom of the stack
# With position_fill(reverse = FALSE), this usually means putting it LAST
stack_levels <- c(cat_levels_by_abundance, bottom_cat)

counts <- counts %>%
  mutate(
    Functional_bin = factor(
      Functional_bin,
      levels = stack_levels
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```
 
 ============================================================
 4. COLOR PALETTE
    12 qualitative colors for up to 12 functional categories.
    Extend or swap as needed.
 ============================================================
 
```{r}
# Number of functional categories
n_cats <- length(levels(counts$Functional_bin))

# Base Pastel palette
base_pastel <- carto_pal(n = 12, name = "Pastel")

# Gray color from Pastel palette
pastel_gray <- base_pastel[12]

# Categories excluding Poor characterization, in stack/factor order
non_poor_levels <- setdiff(levels(counts$Functional_bin), bottom_cat)

# Number of non-gray categories
n_non_poor <- length(non_poor_levels)

# Reverse non-gray Pastel colors so larger categories get later/warmer colors
non_gray_pastel <- rev(base_pastel[-12])

# Expand palette if needed
non_poor_colours <- colorRampPalette(non_gray_pastel)(n_non_poor)

# Assign colors
cat_colours <- c(
  setNames(non_poor_colours, non_poor_levels),
  setNames(pastel_gray, bottom_cat)
)

# Match factor order exactly
cat_colours <- cat_colours[levels(counts$Functional_bin)]
```

 ============================================================
 5. PLOT
 ============================================================
 
```{r}
p <- ggplot(counts,
            aes(x    = Group,
                y    = prop,
                fill = Functional_bin)) +
 
  # Stacked bars — position_fill() ensures each bar sums to 1
  geom_col(
    position = position_fill(reverse = FALSE),
    width    = 0.6,
    colour   = "black",       # thin black border between segments (reference style)
    linewidth = 0.25
  ) +
 
  # Y-axis: 0–1 with no expansion gap at bottom (matches reference)
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.01)),
    labels = label_number(accuracy = 0.25)   # 0, 0.25, 0.50, 0.75, 1.00
  ) +
 
  # Colour scale for functional categories
  scale_fill_manual(
  values = cat_colours,
  breaks = levels(counts$Functional_bin)
  ) +
 
  # Labels
  labs(
    title = "Relative Abundance of Functional Categories Within Carbohydrate Metabolism",
    subtitle = "Female Large Intestine  |  603 Total Differentially Expressed Proteins",
    x     = NULL,
    y     = "Abundance",
    fill  = "Functional Category"
  ) +
 
  # Theme: closely mirrors the reference (classic ggplot2 box style)
  theme_classic(base_size = 13) +
  theme(
    # Full panel border
    panel.border        = element_rect(colour = "black", fill = NA,
                                       linewidth = 0.7),
    axis.line           = element_blank(),   # border replaces axis lines
 
    # Axes
    axis.text.x         = element_text(size = 13, colour = "black"),
    axis.text.y         = element_text(size = 11, colour = "black"),
    axis.title.y        = element_text(size = 13, colour = "black",
                                       margin = margin(r = 8)),
    axis.ticks          = element_line(colour = "black", linewidth = 0.4),
 
    # Title
    plot.title          = element_text(size = 13, face = "bold",
                                       colour = "grey15",
                                       margin = margin(b = 8)),
    plot.margin         = margin(12, 12, 12, 12),
 
    # Legend: right side, stacked vertically
    legend.position     = "right",
    legend.title        = element_text(size = 11, face = "bold"),
    legend.text         = element_text(size = 10),
    legend.key.size     = unit(0.5, "cm"),
    legend.key          = element_rect(colour = "black", linewidth = 0.3),
    legend.margin       = margin(l = 10)
  )
 
print(p)
```

 ============================================================
 6. SAVE
 ============================================================
 
```{r}
ggsave(
  filename = "relative_abundance_carbohydrate_stacked_bar.pdf",
  plot     = p,
  width    = 8,
  height   = 6,
  device   = pdf
)
```

#################################################################
Stacked bar chart for genera in metaprotemic data
#################################################################
```{r}
library(tidyverse)
library(readxl)
library(rcartocolor)
```

Read in data
```{r}
genus_df <- read_excel(
  "female_li_8_2_filtered_hybrid_results_w_function.xlsx",
  sheet = "genus"
)
```

Prepare data
```{r}
# Clean genus names
genus_df <- genus_df %>%
  mutate(
    Genus = if_else(is.na(Genus) | Genus == "", "Unknown", Genus)
  )

# Count total number of proteins per genus before collapsing
genus_counts <- genus_df %>%
  count(Genus, name = "genus_total_n")

# Count how many genera will be collapsed into Other
n_other_genera <- genus_counts %>%
  filter(genus_total_n < 5) %>%
  summarise(n_other = n_distinct(Genus)) %>%
  pull(n_other)

# Create a label that includes the number of genera in Other
other_label <- paste0("Other (", n_other_genera, " genera)")

# Collapse genera with fewer than 5 proteins into Other
genus_plot_df <- genus_df %>%
  left_join(genus_counts, by = "Genus") %>%
  mutate(
    Genus_plot = if_else(
      genus_total_n < 5,
      other_label,
      Genus
    )
  ) %>%
  count(Genus_plot, name = "n") %>%
  mutate(
    prop = n / sum(n)
  )

# Order genera by percent composition
genus_order <- genus_plot_df %>%
  arrange(prop) %>%
  pull(Genus_plot)

# Remove Other and Petralouisia so we can manually place them
genus_order <- genus_order %>%
  setdiff(c(other_label, "Petralouisia"))

# Desired stack order from bottom to top
genus_order <- c(
  other_label,
  genus_order,
  "Petralouisia"
)

genus_plot_df <- genus_plot_df %>%
  mutate(
    Genus_plot = factor(Genus_plot, levels = genus_order)
  )

# Export genus plot dataframe
write_xlsx(genus_plot_df, "genus_plot_df.xlsx")
```

Color palette for plot
```{r}
# Make a lighter/brighter Carto palette
set.seed(752)  # Change this number to reshuffle colors

genus_colors <- colorRampPalette(
  carto_pal(n = 12, name = "Pastel")
)(
  nlevels(genus_plot_df$Genus_plot)
)

# Randomize color order so similar colors are less likely to sit next to each other
genus_colors <- sample(genus_colors)

names(genus_colors) <- levels(genus_plot_df$Genus_plot)

# Keep Other gray so it stays visually de-emphasized
genus_colors[other_label] <- "gray85"
```

Plot
```{r}
genus_stacked_bar_percent <- ggplot(
  genus_plot_df,
  aes(
    x = "Total",
    y = prop,
    fill = Genus_plot
  )
) +
  geom_col(
    width = 0.6,
    color = "black",
    linewidth = 0.25,
    position = position_stack(reverse = TRUE)
  ) +
  coord_flip() +
  scale_fill_manual(
    values = genus_colors,
    breaks = rev(levels(genus_plot_df$Genus_plot))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = NULL,
    y = "Percent of total significant proteins",
    fill = "Genus"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, color = "black"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 16),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 20)
  )

genus_stacked_bar_percent
```

Save
```{r}
ggsave(
  filename = "genus_stacked_bar.pdf",
  plot     = genus_stacked_bar_percent,
  width    = 14,
  height   = 10,
  device   = pdf
)
```

#################################################################
P. muris proteins and metabolic grouping heatmap
#################################################################
```{r}
library(tidyverse)
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)
library(gridtext)
```

Load in data
```{r}
rep_df <- read_excel("imputed_values_female.xlsx")

metadat_df <- read_excel(
  "female_li_8_2_filtered_hybrid_results_w_function.xlsx",
  sheet = "p_muris"
)
```

Prepare data for plotting
```{r}
# Filter imputed proteins to proteins of interest
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)

# Match order between sheets to align metadata
filtered_rep_df <- filtered_rep_df %>%
  slice(match(metadat_df$Protein, Protein))

# Create dataframe that will be plotted as a heatmap
heatmap_data <- filtered_rep_df %>%
  select(starts_with("LI_")) %>%
  as.data.frame()

# Row-wise z-score scaling
heatmap_scaled <- t(scale(t(heatmap_data)))

# Assign row names using Function column
rownames(heatmap_scaled) <- metadat_df$Function

# Order alphabetically by functional category
# Convert Functional_category to alphabetically ordered factor
category_order <- sort(unique(metadat_df$Functional_category))

metadat_df$Functional_category <- factor(
  metadat_df$Functional_category,
  levels = category_order
)

# Order heatmap rows
order_idx <- order(metadat_df$Functional_category)

heatmap_scaled <- heatmap_scaled[order_idx, ]
metadat_df <- metadat_df[order_idx, ]

# Row labels
formatted_labels <- metadat_df$Function

# Column annotation
group <- c(rep("Control", 4), rep("Stress", 4))

ha_col <- HeatmapAnnotation(
  Group = group,
  
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress  = "#FE8A13"
    )
  ),
  
  border = TRUE
)
```

Export P muris proteins with replicates
```{r}
write_xlsx(filtered_rep_df, "p_muris_sig_prot.xlsx")
```

Coloring for functional categories
```{r}
functional_categories <- levels(droplevels(metadat_df$Functional_category))

set.seed(1)

category_colors <- structure(
  sample(rainbow(length(functional_categories), s = 0.7, v = 0.85)),
  names = as.character(functional_categories)
)

category_legend <- Legend(
  title = "Functional Category",
  labels = names(category_colors),
  legend_gp = gpar(fill = category_colors),
  ncol = 1
)
```

Row annotation
```{r}
ha_row <- rowAnnotation(
  Category = metadat_df$Functional_category,
  
  col = list(Category = category_colors),
  
  border = TRUE,
  
  show_annotation_name = FALSE,
  
  show_legend = FALSE,
  
  annotation_width = unit(3, "mm")
)
```

Plot heatmap
```{r}
ht <- Heatmap(
  heatmap_scaled,
  
  width = unit(5, "cm"),
  
  name = "Z-score",
  
  col = viridis(100),
  
  top_annotation = ha_col,
  left_annotation = ha_row,
  
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  
  row_split = metadat_df$Functional_category,
  
  row_title = NULL,
  
  show_row_names = TRUE,
  
  row_labels = formatted_labels,
  
  row_names_gp = gpar(fontsize = 10),
  
  column_names_gp = gpar(fontsize = 10),
  
  border = TRUE
)

# Draw heatmap
draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(category_legend)
)
```

Save heatmap
```{r}
pdf(
  "p_muris_functional_category_heatmap_female_li.pdf",
  width = 16,
  height = 8
)

draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(category_legend)
)

dev.off()
```

