---
title: "metaproteomics_visualizations"
author: "Amelia Slusser"
date: "2026-04-29"
output: html_document
---
################################################################################
Male
################################################################################
---------------------- Statistical dataframe ----------------------
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("m_li_imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("m_li_pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "m_li_stats_with_reps.xlsx"
)
```

---------------------- Volcano plot (Fig. 3E) ----------------------
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("m_li_pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "M LI"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "male_li_volcano_plot.pdf",
  plot = volcano_plot,
  width = 7,
  height = 6,
  units = "in"
)
```

---------------------- Stress response proteins (Fig. S3F & 3A) ----------------------
Plot select data as a heatmap with replicates per group
Load in packages
```{r}
library(readxl)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)
library(gridtext)
library(stringr)
library(scales)
```

Load in data from excel spread sheets
```{r}
rep_df <- read_excel("m_li_imputed_values.xlsx")
metadat_df <- read_excel("male_li_results_w_function.xlsx", sheet = "stress_regulated_selected")
```

Filter imputed values to only proteins of interest for plotting
```{r}
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)

# Match order between sheets to align metadata
filtered_rep_df <- filtered_rep_df %>%
  slice(match(metadat_df$Protein, Protein))
```

Save sig with reps file
```{r}
write_xlsx(filtered_rep_df, "m_li_sig_reps_stress_prot.xlsx")
```

Create dataframe that will be plotted as a heatmap
```{r}
heatmap_data <- filtered_rep_df %>%
  select(starts_with("LI_")) %>%
  as.data.frame()

# Row-wise z-score scaling
heatmap_scaled <- t(scale(t(heatmap_data)))

# Assign row names
rownames(heatmap_scaled) <- metadat_df$name
```

Order data by bin
```{r}
bin_order <- metadat_df %>%
  distinct(Functional_bin) %>%
  pull(Functional_bin)

metadat_df$Functional_bin <- factor(
  metadat_df$Functional_bin,
  levels = bin_order
)

# Order heatmap rows
order_idx <- order(metadat_df$Functional_bin)

heatmap_scaled <- heatmap_scaled[order_idx, ]
metadat_df <- metadat_df[order_idx, ]
```

Create italicized labels
```{r}
formatted_labels <- sapply(metadat_df$name, function(x) {

  parts <- strsplit(x, " - ")[[1]]

  species <- parts[1]

  remainder <- ifelse(length(parts) > 1,
                      paste(parts[-1], collapse = " - "),
                      "")

  paste0("<i>", species, "</i> - ", remainder)
})
```

Annotate columns
```{r}
group <- c(rep("Control", 6), rep("Stress", 6))

ha_col <- HeatmapAnnotation(
  Group = group,
  
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress = "#FE8A13"
    )
  ),
  
  border = TRUE
)
```

functional coloring
```{r}
functional_bins <- levels(droplevels(metadat_df$Functional_bin))

set.seed(1)

bin_colors <- structure(
  sample(rainbow(length(functional_bins), s = 0.7, v = 0.85)),
  names = as.character(functional_bins)
)

bin_legend <- Legend(
  title = "Functional category",
  labels = names(bin_colors),
  legend_gp = gpar(fill = bin_colors),
  ncol = 1
)
```

Row annotation
```{r}
ha_row <- rowAnnotation(
  Functional_bin = metadat_df$Functional_bin,
  
  col = list(
    Functional_bin = bin_colors
  ),
  
  border = TRUE,
  
  show_annotation_name = FALSE,
  
  show_legend = FALSE,
  
  annotation_width = unit(3, "mm")
)
```

Plot the heatmap using viridis
```{r}
ht_strs <- Heatmap(
  heatmap_scaled,
  
  width = unit(5, "cm"),
  
  name = "Z-score",
  
  col = viridis(100),
  
  top_annotation = ha_col,
  left_annotation = ha_row,
  
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  
  row_split = metadat_df$Functional_bin,
  
  row_title = NULL,
  
  show_row_names = TRUE,
  
  row_labels = gt_render(formatted_labels),
  
  row_names_gp = gpar(fontsize = 10),
  
  column_names_gp = gpar(fontsize = 10),
  
  border = TRUE
)
```

Draw heatmap
```{r}
draw(
  ht_strs,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(bin_legend)
)
```

Save heat map as a PDF
```{r}
pdf(
  "Functional_heatmap_male_li_stress_regulated.pdf",
  width = 18,
  height = 12
)

draw(
  ht_strs,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(bin_legend)
)

dev.off()
```

---------------------- Organism-specific stress response proteins (Fig. 3B-D) ----------------------
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
```

Load in data from excel spread sheets
```{r}
rep_df <- read_excel("m_li_imputed_values_male.xlsx")
metadat_df <- read_excel("male_li_results_w_function.xlsx", sheet = "stress_regulated_selected")
```

Filter imputed values to only proteins of interest for plotting
```{r}
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)

# Match order between sheets to align metadata
filtered_rep_df <- filtered_rep_df %>%
  dplyr::slice(match(metadat_df$Protein, Protein))
```

Choose proteins to plot
```{r}
# P. muris
proteins_to_plot <- c(
  "WP_135987996.1",
  "WP_135990045.1"
)

# A. rhamnosivorans
#proteins_to_plot <- c(
  #"QCP36024.1",
  #"QCP34468.1"
#)

# A. sedimenticola
#proteins_to_plot <- c(
  #"QHQ59687.1",
  #"QHQ60156.1",
  #"QHQ60147.1"
#)
```

Filter data
```{r}
plot_metadat_df <- metadat_df %>%
  filter(Protein %in% proteins_to_plot) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot)
  ) %>%
  arrange(Protein) %>%
  select(
    Protein,
    Plot_Label = Function
  )

plot_rep_df <- rep_df %>%
  filter(Protein %in% proteins_to_plot) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot)
  ) %>%
  arrange(Protein)

# Combine with the Function column
sample_cols <- names(plot_rep_df) %>%
  keep(~ str_starts(.x, "LI_"))

group_key <- tibble(
  Sample = sample_cols,
  Group = c(rep("Control", 6), rep("Stress", 6))
)

plot_df <- plot_rep_df %>%
  left_join(
    plot_metadat_df,
    by = "Protein"
  ) %>%
  pivot_longer(
    cols = all_of(sample_cols),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  left_join(
    group_key,
    by = "Sample"
  ) %>%
  mutate(
    Group = factor(Group, levels = c("Control", "Stress")),
    Plot_Label = factor(Plot_Label, levels = plot_metadat_df$Plot_Label)
  )

# Calculate Control vs Stress statistics separately for each protein
protein_stats <- plot_df %>%
  group_by(Protein, Plot_Label) %>%
  rstatix::t_test(
    Intensity ~ Group,
    comparisons = list(c("Control", "Stress")),
    var.equal = FALSE
  ) %>%
  rstatix::add_significance(
    p.col = "p",
    output.col = "p.signif"
  ) %>%
  ungroup()

# Calculate a facet-specific bracket position
stat_positions <- plot_df %>%
  group_by(Protein, Plot_Label) %>%
  summarize(
    y_max = max(Intensity, na.rm = TRUE),
    y_min = min(Intensity, na.rm = TRUE),
    y_range = y_max - y_min,
    y.position = if_else(
      y_range == 0,
      y_max + 0.5,
      y_max + 0.15 * y_range
    ),
    .groups = "drop"
  )

# Add bracket positions to the statistics table
protein_stats <- protein_stats %>%
  left_join(
    stat_positions,
    by = c("Protein", "Plot_Label")
  ) %>%
  mutate(
    group1 = factor(group1, levels = c("Control", "Stress")),
    group2 = factor(group2, levels = c("Control", "Stress"))
  )
```

Plot
```{r}
protein_boxplot <- ggplot(
  plot_df,
  aes(
    x = Group,
    y = Intensity,
    fill = Group
  )
) +
  
  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.7,
    alpha = 0.85
  ) +
  
  geom_jitter(
    shape = 21,
    color = "black",
    width = 0.12,
    size = 3.8,
    stroke = 0.5,
    alpha = 0.9
  ) +
  
  # Add significance brackets and statistical stars
  ggpubr::stat_pvalue_manual(
    protein_stats,
    label = "p.signif",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.8,
    size = 10,
    hide.ns = FALSE,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(
    ~ Plot_Label,
    scales = "free_y",
    ncol = 2,
    
    # Wrap long facet labels at approximately 35 characters
    labeller = labeller(
      Plot_Label = label_wrap_gen(
        width = 35,
        multi_line = TRUE
      )
    )
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  # Add additional space above each facet for the brackets
  scale_y_continuous(
    expand = expansion(
      mult = c(0.05, 0.22)
    )
  ) +
  
  labs(
    x = NULL,
    y = "Log2 intensity",
    fill = NULL
  ) +
  
  theme_classic(base_size = 18) +
  
  theme(
    # Facet-label boxes
    strip.background = element_rect(
      fill = "gray85",
      color = "black",
      linewidth = 0.7
    ),
    
    strip.text = element_text(
      face = "bold",
      color = "black",
      size = 17,
      lineheight = 1,
      
      # Enlarges the gray box around wrapped text
      margin = margin(
        t = 9,
        r = 8,
        b = 9,
        l = 8
      )
    ),
    
    # Axis text
    axis.text.x = element_text(
      color = "black",
      size = 17,
      face = "bold",
      margin = margin(t = 7)
    ),
    
    axis.text.y = element_text(
      color = "black",
      size = 16
    ),
    
    # Axis title
    axis.title.y = element_text(
      color = "black",
      size = 20,
      face = "bold",
      margin = margin(r = 10)
    ),
    
    # Additional spacing between facets
    panel.spacing = grid::unit(1.2, "lines"),
    
    legend.position = "none"
  )

protein_boxplot
```

Save
```{r}
ggsave(
  filename = "p._muris_selected_stress_boxplots_male_li.pdf",
  plot = protein_boxplot,
  width = 9.5,
  height = 5,
  dpi = 300
)
```

---------------------- GH family heatmap (Fig. S5A & 5A) ----------------------
Plot select data as a heatmap with replicates per group
Load in packages
```{r}
library(readxl)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)
library(gridtext)
library(stringr)
library(scales)
```

Load in data from excel spread sheets
```{r}
rep_df <- read_excel("m_li_imputed_values_male.xlsx")
metadat_df <- read_excel("male_li_results_w_function.xlsx", sheet = "complex_carb_selected")
```

Filter imputed values to only proteins of interest for plotting
```{r}
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)
# Match order between sheets to align metadata (safer ordering)
filtered_rep_df <- filtered_rep_df %>%
  slice(match(metadat_df$Protein, Protein))
```

Save sig with reps file
```{r}
write_xlsx(filtered_rep_df, "m_li_gh_sig_reps.xlsx")
```

Create dataframe that will be plotted as a heatmap
```{r}
heatmap_data <- filtered_rep_df %>%
  select(starts_with("LI_")) %>%
  as.data.frame()

# Row-wise z-score scaling
heatmap_scaled <- t(scale(t(heatmap_data)))

# Assign row names
rownames(heatmap_scaled) <- metadat_df$name
```

Order data by GH family
```{r}
# Manually define GH family order
gh_order <- c(
  "GH1",
  "GH2",
  "GH3",
  "GH13",
  "GH29",
  "GH33",
  "GH35",
  "GH36",
  "GH42",
  "GH43",
  "GH51",
  "GH74",
  "GH127",
  "GH130"
)
# Convert to a factor
metadat_df$`GH family` <- factor(
  metadat_df$`GH family`,
  levels = gh_order
)
# Order heatmap rows
order_idx <- order(metadat_df$`GH family`)

heatmap_scaled <- heatmap_scaled[order_idx, ]
metadat_df <- metadat_df[order_idx, ]
```

Create italicized labels
```{r}
formatted_labels <- sapply(metadat_df$name, function(x) {

  parts <- strsplit(x, " - ")[[1]]

  species <- parts[1]

  remainder <- ifelse(length(parts) > 1,
                      paste(parts[-1], collapse = " - "),
                      "")

  paste0("<i>", species, "</i> - ", remainder)
})
```

Annotate columns
```{r}
group <- c(rep("Control", 6), rep("Stress", 6))

ha_col <- HeatmapAnnotation(
  Group = group,
  
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress = "#FE8A13"
    )
  ),
  
  border = TRUE
)
```

GH family coloring
```{r}
gh_families <- levels(droplevels(metadat_df$`GH family`))

set.seed(1)

gh_colors <- structure(
  sample(rainbow(length(gh_families), s = 0.7, v = 0.85)),
  names = as.character(gh_families)
)

gh_legend <- Legend(
  title = "GH Family",
  labels = names(gh_colors),
  legend_gp = gpar(fill = gh_colors),
  ncol = 1
)
```

Row annotation
```{r}
ha_row <- rowAnnotation(
  GH = metadat_df$`GH family`,

  col = list(GH = gh_colors),

  border = TRUE, 
  
  show_annotation_name = FALSE,

  show_legend = FALSE,

  annotation_width = unit(3, "mm")
)
```

Plot the heatmap using viridis
```{r}
ht <- Heatmap(
  heatmap_scaled,

  width = unit(5, "cm"),
  
  name = "Z-score",

  col = viridis(100),

  top_annotation = ha_col,
  left_annotation = ha_row,

  cluster_rows = FALSE,
  cluster_columns = TRUE,

  row_split = metadat_df$`GH family`,

  row_title = NULL,

  show_row_names = TRUE,

  row_labels = gt_render(formatted_labels),

  row_names_gp = gpar(fontsize = 10),

  column_names_gp = gpar(fontsize = 10),

  border = TRUE
)
```

Draw heatmap
```{r}
draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(gh_legend)
)
```

Save heat map as a PDF
```{r}
pdf(
  "GH_family_heatmap_male_li.pdf",
  width =16,
  height = 6
)

draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(gh_legend)
)

dev.off()
```

---------------------- Arabinose isomerases (Fig. 5C) ----------------------
```{r}
library(readxl)
library(tidyverse)
library(ggnewscale)
```

```{r}
# Create data frame
arab_iso_df <- data.frame(
  Sample = c(
    "LI_2R", "LI_2L", "LI_2N", "LI_4R", "LI_4L", "LI_4N",
    "LI_8", "LI_9", "LI_10", "LI_14", "LI_15", "LI_16"
  ),
  Log2_Intensity = c(
    24.11368032, 22.82925275, 23.15363277, 23.66263724, 23.9576346, 23.59308378,
    13.75938998, 13.92129068, 14.49067997, 16.3292061, 13.79934126, 16.79112359
  )
)

# Add group information
arab_iso_df <- arab_iso_df %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress"
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )

# Plot
arab_iso_plot <- ggplot(arab_iso_df, aes(x = Group, y = Log2_Intensity, fill = Group)) +
  
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.18,
    linewidth = 0.6,
    color = "black"
  ) +
  
  geom_jitter(
    shape = 21,
    color = "black",
    width = 0.12,
    size = 4,
    stroke = 0.5,
    alpha = 0.95
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress" = "#FE8A13"
    )
  ) +
  
  labs(
    title = "L-arabinose isomerase",
    x = NULL,
    y = "log2 intensity"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "italic",
      size = 16
    ),
    axis.text.x = element_text(
      color = "black",
      size = 13
    ),
    axis.text.y = element_text(
      color = "black",
      size = 12
    ),
    axis.title.y = element_text(
      color = "black",
      size = 14
    ),
    legend.position = "none"
  )

arab_iso_plot
```

More replicate data
```{r}
# Create data frame: one row per arabinose isomerase protein
arab_iso_wide <- data.frame(
  Protein = paste0("Arab_iso_", 1:8),
  
  LI_2R = c(25.18520163, 24.94279731, 25.29151547, 21.66082115, 22.08799437, 24.02147435, 23.79678867, 26.47895119),
  LI_2L = c(22.09947004, 22.27698834, 23.40961778, 22.27698834, 22.20670323, 23.44770931, 21.90569884, 23.89622702),
  LI_2N = c(24.04184345, 22.87496804, 23.11653658, 21.65988568, 22.9292736, 22.88451935, 21.27891846, 25.62418605),
  LI_4R = c(24.32007465, 23.58292204, 24.58100694, 22.56653239, 23.15168295, 23.12491663, 22.51964181, 25.80986456),
  LI_4L = c(22.97940175, 24.1426314, 24.23329067, 21.51574623, 23.35831294, 24.18996645, 22.51846268, 26.09262615),
  LI_4N = c(24.19506548, 24.19506548, 24.23829669, 21.84505017, 23.31441275, 23.53604915, 21.93114633, 25.57948882),
  
  LI_8 = c(18.43611154, 16.18489406, 17.74539151, 11.96770412, 11.96770412, 11.96770412, 14.97787138, 15.09133125),
  LI_9 = c(15.96524372, 16.39687652, 14.94335051, 11.04504083, 11.04504083, 11.04504083, 15.62085813, 17.67807304),
  LI_10 = c(23.06424774, 16.19705485, 23.395165, 11.04641151, 11.04641151, 11.04641151, 17.60790938, 16.5130809),
  LI_14 = c(16.72150384, 16.23358471, 17.30598662, 11.77021159, 11.77021159, 11.77021159, 16.5254586, 25.61600236),
  LI_15 = c(17.08289903, 21.2699493, 17.429364, 11.50011976, 11.50011976, 11.50011976, 14.3991348, 16.45521658),
  LI_16 = c(24.48099289, 16.72720736, 17.7793825, 9.395038869, 9.395038869, 9.395038869, 25.03149178, 23.20757735)
)

# Convert to long format
arab_iso_long <- arab_iso_wide %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Log2_Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```

Connected mean summary lines
```{r}
arab_iso_summary <- arab_iso_long %>%
  group_by(Protein, Group) %>%
  summarise(
    Mean_Log2_Intensity = mean(Log2_Intensity, na.rm = TRUE),
    .groups = "drop"
  )
```

Plot
```{r}
arab_iso_connected_plot <- ggplot() +
  
  # Replicate dots: original darker colors
  geom_jitter(
    data = arab_iso_long,
    aes(
      x = Group,
      y = Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    width = 0.16,
    size = 5,
    stroke = 0.4,
    alpha = 0.95
  ) +
  
    scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # Faint connected protein-level mean lines
  geom_line(
    data = arab_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      group = Protein
    ),
    color = "gray45",
    linewidth = 0.7,
    alpha = 0.5
  ) +
  
  # Mean points: lighter colors and drawn last
  geom_point(
    data = arab_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    size = 6.5,
    stroke = 0.7,
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  labs(
    title = "L-arabinose isomerase LI M",
    x = NULL,
    y = "Log2 intensity"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 25
    ),
    axis.text.x = element_text(
      color = "black",
      face = "bold",
      size = 23
    ),
    axis.text.y = element_text(
      color = "black",
      size = 23
    ),
    axis.title.y = element_text(
      color = "black",
      face = "bold",
      size = 25
    ),
    legend.position = "none"
  )

arab_iso_connected_plot
```

Save plot
```{r}
ggsave(
  filename = "m_li_arabinose_isomerase_plot.pdf",
  plot = arab_iso_connected_plot,
  width = 6,
  height = 6
)
```

---------------------- Xylose isomerases (Fig. 5D) ----------------------
```{r}
library(readxl)
library(tidyverse)
library(ggnewscale)
```

Replicate data
```{r}
# Create data frame: one row per xylose isomerase protein
xylo_iso_wide <- data.frame(
  Protein = paste0("Xylo_iso_", 1:8),
  
  LI_2R = c(21.76556329, 23.14371343),
  LI_2L = c(21.0097721, 23.13050296),
  LI_2N = c(21.0152900, 23.66535397),
  LI_4R = c(22.21457453, 23.9050938),
  LI_4L = c(21.52527622, 23.38993302),
  LI_4N = c(20.23979984, 22.13690788),
  
  LI_8 = c(11.96770412, 11.96770412),
  LI_9 = c(11.04504083, 11.04504083),
  LI_10 = c(11.04641151, 11.04641151),
  LI_14 = c(11.77021159, 11.77021159),
  LI_15 = c(11.50011976, 11.50011976),
  LI_16 = c(9.395038869, 9.395038869)
)

# Convert to long format
xylo_iso_long <- xylo_iso_wide %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Log2_Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```

Connected mean summary lines
```{r}
xylo_iso_summary <- xylo_iso_long %>%
  group_by(Protein, Group) %>%
  summarise(
    Mean_Log2_Intensity = mean(Log2_Intensity, na.rm = TRUE),
    .groups = "drop"
  )
```

Plot
```{r}
xylo_iso_connected_plot <- ggplot() +
  
  # Replicate dots: original darker colors
  geom_jitter(
    data = xylo_iso_long,
    aes(
      x = Group,
      y = Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    width = 0.16,
    size = 5,
    stroke = 0.4,
    alpha = 0.95
  ) +
  
    scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # Faint connected protein-level mean lines
  geom_line(
    data = xylo_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      group = Protein
    ),
    color = "gray45",
    linewidth = 0.7,
    alpha = 0.5
  ) +
  
  # Mean points: lighter colors and drawn last
  geom_point(
    data = xylo_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    size = 6.5,
    stroke = 0.7,
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  labs(
    title = "Xylose isomerase LI M",
    x = NULL,
    y = "Log2 intensity"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 25
    ),
    axis.text.x = element_text(
      color = "black",
      face = "bold",
      size = 23
    ),
    axis.text.y = element_text(
      color = "black",
      size = 23
    ),
    axis.title.y = element_text(
      color = "black",
      face = "bold",
      size = 25
    ),
    legend.position = "none"
  )

xylo_iso_connected_plot
```

Save plot
```{r}
ggsave(
  filename = "m_li_xylose_isomerase_plot.pdf",
  plot = xylo_iso_connected_plot,
  width = 6,
  height = 6
)
```

################################################################################
Female
################################################################################
---------------------- Statistical dataframe ----------------------
```{r}
library(tidyverse)
library(writexl)
```

Load and prepare data frame
```{r}
# Replicate data frame
rep_df <- read_tsv("f_li_imputed_values.tsv", col_name = TRUE)
# Stats data frame
stats_df <- read_tsv("f_li_pca_weighted_drivers.tsv", col_name = TRUE)

# Combine dataframes by Protein
combined_df <- full_join(
  rep_df,
  stats_df,
  by = "Protein",
  suffix = c("_rep", "_stats")
)

# Export to Excel
write_xlsx(
  combined_df,
  path = "f_li_stats_with_reps.xlsx"
)
```

---------------------- Volcano plot (Fig. S4E) ----------------------
```{r}
library(readxl)
library(tidyverse)
```

Load and prepare data
```{r}
# Load data
volcano_df <- read_excel("f_li_pca_weighted_drivers.xlsx")

# Define significance categories for plotting
volcano_df <- volcano_df %>%
  mutate(
    volcano_group = case_when(
      adj.P.Val < 0.05 & logFC > 2  ~ "Significantly increased",
      adj.P.Val < 0.05 & logFC < -2 ~ "Significantly decreased",
      TRUE                          ~ "Not significant"
    ),
    volcano_group = factor(
      volcano_group,
      levels = c(
        "Significantly decreased",
        "Not significant",
        "Significantly increased"
      )
    )
  )
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = logFC, y = log10p, color = volcano_group)
) +
  
  geom_point(
    size = 2.2,
    alpha = 0.85
  ) +
  
  # Threshold lines
  geom_hline(
    yintercept = 1.3,
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  geom_vline(
    xintercept = c(-2, 2),
    linetype = "dashed",
    color = "gray80",
    linewidth = 0.6
  ) +
  
  # Colors
  scale_color_manual(
    values = c(
      "Significantly decreased" = "#4D4D4D",
      "Not significant"         = "gray80",
      "Significantly increased" = "#FE8A13"
    )
  ) +
  
  labs(
    x = "Log2 fold change",
    y = expression(-Log[10](p-value)),
    color = NULL,
    title = "F LI"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(face = "plain"),
    legend.position = "right"
  )

# View plot
volcano_plot
```

Save
```{r}
ggsave(
  filename = "female_li_volcano_plot.pdf",
  plot = volcano_plot,
  width = 7,
  height = 5,
  units = "in"
)
```

---------------------- Genus stacked bar chart (Fig. S4F) ----------------------
```{r}
library(tidyverse)
library(readxl)
library(rcartocolor)
```

Read in data
```{r}
genus_df <- read_excel(
  "female_li_8_2_filtered_hybrid_results_w_function.xlsx",
  sheet = "genus"
)
```

Prepare data
```{r}
# Clean genus names
genus_df <- genus_df %>%
  mutate(
    Genus = if_else(is.na(Genus) | Genus == "", "Unknown", Genus)
  )

# Count total number of proteins per genus before collapsing
genus_counts <- genus_df %>%
  count(Genus, name = "genus_total_n")

# Count how many genera will be collapsed into Other
n_other_genera <- genus_counts %>%
  filter(genus_total_n < 5) %>%
  summarise(n_other = n_distinct(Genus)) %>%
  pull(n_other)

# Create a label that includes the number of genera in Other
other_label <- paste0("Other (", n_other_genera, " genera)")

# Collapse genera with fewer than 5 proteins into Other
genus_plot_df <- genus_df %>%
  left_join(genus_counts, by = "Genus") %>%
  mutate(
    Genus_plot = if_else(
      genus_total_n < 5,
      other_label,
      Genus
    )
  ) %>%
  count(Genus_plot, name = "n") %>%
  mutate(
    prop = n / sum(n)
  )

# Order genera by percent composition
genus_order <- genus_plot_df %>%
  arrange(prop) %>%
  pull(Genus_plot)

# Remove Other and Petralouisia so we can manually place them
genus_order <- genus_order %>%
  setdiff(c(other_label, "Petralouisia"))

# Desired stack order from bottom to top
genus_order <- c(
  other_label,
  genus_order,
  "Petralouisia"
)

genus_plot_df <- genus_plot_df %>%
  mutate(
    Genus_plot = factor(Genus_plot, levels = genus_order)
  )

# Export genus plot dataframe
write_xlsx(genus_plot_df, "genus_plot_df.xlsx")
```

Color palette for plot
```{r}
# Make a lighter/brighter Carto palette
set.seed(752)  # Change this number to reshuffle colors

genus_colors <- colorRampPalette(
  carto_pal(n = 12, name = "Pastel")
)(
  nlevels(genus_plot_df$Genus_plot)
)

# Randomize color order so similar colors are less likely to sit next to each other
genus_colors <- sample(genus_colors)

names(genus_colors) <- levels(genus_plot_df$Genus_plot)

# Keep Other gray so it stays visually de-emphasized
genus_colors[other_label] <- "gray85"
```

Plot
```{r}
genus_stacked_bar_percent <- ggplot(
  genus_plot_df,
  aes(
    x = "Total",
    y = prop,
    fill = Genus_plot
  )
) +
  geom_col(
    width = 0.6,
    color = "black",
    linewidth = 0.25,
    position = position_stack(reverse = TRUE)
  ) +
  coord_flip() +
  scale_fill_manual(
    values = genus_colors,
    breaks = rev(levels(genus_plot_df$Genus_plot))
  ) +
  scale_y_continuous(
    labels = scales::percent_format(accuracy = 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  labs(
    x = NULL,
    y = "Percent of total significant proteins",
    fill = "Genus"
  ) +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.text.x = element_text(size = 18, color = "black"),
    axis.title.x = element_text(size = 20, color = "black"),
    legend.title = element_text(face = "bold"),
    legend.text = element_text(size = 16),
    plot.margin = margin(t = 10, r = 10, b = 10, l = 20)
  )

genus_stacked_bar_percent
```

Save
```{r}
ggsave(
  filename = "genus_stacked_bar.pdf",
  plot     = genus_stacked_bar_percent,
  width    = 14,
  height   = 10,
  device   = pdf
)
```

---------------------- Top 12 genera by protein count (Fig. 4A) ----------------------
```{r}
library(dplyr)
library(ggplot2)
library(stringr)
library(tidyr)
```

Create data frame
```{r}
# Input: one row per genus, separate columns for increased and decreased counts
df_microbe <- data.frame(
  genus = c("Petralouisia", "Blautia", "Roseburia",
            "Dorea", "Turicibacter", "Acutalibacter",
            "Acetatifactor", "Romboutsia", "Clostridium",
            "Eubacterium", "Ruminococcus", "Sporofaciens"),
  increased = c(49, 19, 14, 4, 1, 0, 5, 2, 5, 6, 7, 1),
  decreased = c(3, 25, 20, 29, 28, 22, 15, 18, 14, 12, 9, 13)
)
```

Mirror: decreased becomes negative; sort genera by net change
```{r}
df_microbe <- df_microbe %>%
  mutate(
    net = increased - decreased,
    genus = reorder(genus, net),
    decreased_mirror = -decreased
  ) %>%
  pivot_longer(
    cols = c(increased, decreased_mirror),
    names_to = "direction",
    values_to = "count"
  ) %>%
  mutate(direction = recode(direction,
    "increased" = "Increased",
    "decreased_mirror" = "Decreased"
  ))

# Export protein count data frame
write_xlsx(df_microbe, "genus_protein_count.xlsx")
```

Plot diverging bar graph
```{r}
microbe_div_bar <- ggplot(df_microbe, aes(x = count, y = genus, fill = direction)) +
  geom_col(
    width = 0.7,
    color = "black",
    linewidth = 0.4
  ) +
  geom_vline(xintercept = 0, linewidth = 0.6, color = "grey40") +
  scale_fill_manual(values = c("Increased" = "#FE8A13", "Decreased" = "#4D4D4D")) +
  scale_x_continuous(
    labels = function(x) abs(x),     # show absolute values on axis
    expand = expansion(mult = 0.05)
  ) +
  labs(
    x = "Number of differentially abundant proteins",
    y = NULL,
    fill = NULL,
    title = "Changes in microbial protein abundance - Top 12 genera by total protein count"
  ) +
  theme_minimal(base_size = 23) +
  theme(
    legend.position = "top",
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "italic", size = 20),
    axis.text.x = element_text(size = 18),
    axis.title.x = element_text(face = "bold", size = 20),
    plot.title = element_text(face = "bold", size = 20)
  )
microbe_div_bar
```

Save plot
```{r}
ggsave(
  "microbial_diverging_barplot.pdf",
  microbe_div_bar,
  width = 10,
  height = 6,
  dpi = 600
)
```

---------------------- P. muris proteins and metabolic groupings heatmap (Fig. 4B) ----------------------
```{r}
library(tidyverse)
library(readxl)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)
library(gridtext)
```

Load in data
```{r}
rep_df <- read_excel("imputed_values_female.xlsx")

metadat_df <- read_excel(
  "female_li_8_2_filtered_hybrid_results_w_function.xlsx",
  sheet = "p_muris"
)
```

Prepare data for plotting
```{r}
# Filter imputed proteins to proteins of interest
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)

# Match order between sheets to align metadata
filtered_rep_df <- filtered_rep_df %>%
  slice(match(metadat_df$Protein, Protein))

# Create dataframe that will be plotted as a heatmap
heatmap_data <- filtered_rep_df %>%
  select(starts_with("LI_")) %>%
  as.data.frame()

# Row-wise z-score scaling
heatmap_scaled <- t(scale(t(heatmap_data)))

# Assign row names using Function column
rownames(heatmap_scaled) <- metadat_df$Function

# Order alphabetically by functional category
# Convert Functional_category to alphabetically ordered factor
category_order <- sort(unique(metadat_df$Functional_category))

metadat_df$Functional_category <- factor(
  metadat_df$Functional_category,
  levels = category_order
)

# Order heatmap rows
order_idx <- order(metadat_df$Functional_category)

heatmap_scaled <- heatmap_scaled[order_idx, ]
metadat_df <- metadat_df[order_idx, ]

# Row labels
formatted_labels <- metadat_df$Function

# Column annotation
group <- c(rep("Control", 4), rep("Stress", 4))

ha_col <- HeatmapAnnotation(
  Group = group,
  
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress  = "#FE8A13"
    )
  ),
  
  border = TRUE
)
```

Export P muris proteins with replicates
```{r}
write_xlsx(filtered_rep_df, "p_muris_sig_prot.xlsx")
```

Coloring for functional categories
```{r}
functional_categories <- levels(droplevels(metadat_df$Functional_category))

set.seed(1)

category_colors <- structure(
  sample(rainbow(length(functional_categories), s = 0.7, v = 0.85)),
  names = as.character(functional_categories)
)

category_legend <- Legend(
  title = "Functional Category",
  labels = names(category_colors),
  legend_gp = gpar(fill = category_colors),
  ncol = 1
)
```

Row annotation
```{r}
ha_row <- rowAnnotation(
  Category = metadat_df$Functional_category,
  
  col = list(Category = category_colors),
  
  border = TRUE,
  
  show_annotation_name = FALSE,
  
  show_legend = FALSE,
  
  annotation_width = unit(3, "mm")
)
```

Plot heatmap
```{r}
ht <- Heatmap(
  heatmap_scaled,
  
  width = unit(5, "cm"),
  
  name = "Z-score",
  
  col = viridis(100),
  
  top_annotation = ha_col,
  left_annotation = ha_row,
  
  cluster_rows = FALSE,
  cluster_columns = TRUE,
  
  row_split = metadat_df$Functional_category,
  
  row_title = NULL,
  
  show_row_names = TRUE,
  
  row_labels = formatted_labels,
  
  row_names_gp = gpar(fontsize = 10),
  
  column_names_gp = gpar(fontsize = 10),
  
  border = TRUE
)

# Draw heatmap
draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(category_legend)
)
```

Save heatmap
```{r}
pdf(
  "p_muris_functional_category_heatmap_female_li.pdf",
  width = 16,
  height = 8
)

draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(category_legend)
)

dev.off()
```

---------------------- Box and whisker plots for P. muris Pdu proteins (Fig. 4D) ----------------------
```{r}
library(tidyverse)
library(readxl)
library(ggpubr)
```

Load in data
```{r}
protein_df <- read_excel("female_li_8_2_filtered_hybrid_results_w_function.xlsx", sheet = "p_muris_pro")

# Convert from wide to long format
protein_long <- protein_df %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress"))
  )
```

Choose proteins to plot
```{r}
# This order controls the order of the boxplots/facets
proteins_to_plot <- c(
  "WP_135987631.1",
  "WP_135989817.1",
  "WP_135989873.1",
  "WP_135989816.1",
  "WP_135989815.1",
  "WP_135989814.1",
  "WP_135987782.1",
  "WP_135987616.1"
)

plot_df <- protein_long %>%
  filter(Protein %in% proteins_to_plot) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot)
  )
```

Create ordered facet labels
```{r}
facet_order_df <- protein_df %>%
  filter(Protein %in% proteins_to_plot) %>%
  distinct(Protein, Function) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot),
    Plot_Label = Function
  ) %>%
  arrange(Protein)

plot_label_order <- facet_order_df$Plot_Label

plot_df <- plot_df %>%
  left_join(facet_order_df, by = c("Protein", "Function")) %>%
  mutate(
    Plot_Label = factor(Plot_Label, levels = plot_label_order)
  )
```

Significance labels and brackets
```{r}
sig_df <- protein_df %>%
  filter(Protein %in% proteins_to_plot) %>%
  mutate(
    Protein = factor(Protein, levels = proteins_to_plot),
    sig_label = case_when(
      adj.P.Val < 0.001 ~ "***",
      adj.P.Val < 0.01  ~ "**",
      adj.P.Val < 0.05  ~ "*",
      TRUE ~ "ns"
    )
  ) %>%
  select(Protein, Function, adj.P.Val, sig_label)

bracket_df <- plot_df %>%
  group_by(Protein, Function, Plot_Label) %>%
  summarise(
    y.position = max(Intensity, na.rm = TRUE) + 
      0.12 * diff(range(Intensity, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  left_join(sig_df, by = c("Protein", "Function")) %>%
  mutate(
    group1 = "Control",
    group2 = "Stress",
    Plot_Label = factor(Plot_Label, levels = plot_label_order)
  )
```

Plot
```{r}
protein_boxplot <- ggplot(
  plot_df,
  aes(
    x = Group,
    y = Intensity,
    fill = Group
  )
) +
  
  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    color = "black",
    linewidth = 0.7,
    alpha = 0.85
  ) +
  
  geom_jitter(
    aes(fill = Group),
    shape = 21,
    color = "black",
    width = 0.12,
    size = 3.5,
    stroke = 0.5,
    alpha = 0.9
  ) +
  
  stat_pvalue_manual(
    bracket_df,
    label = "sig_label",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.8,
    size = 9,
    hide.ns = FALSE,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(
    ~ Plot_Label,
    scales = "free_y",
    ncol = 2,
    
    # Automatically wrap long facet labels
    labeller = labeller(
      Plot_Label = label_wrap_gen(
        width = 32,
        multi_line = TRUE
      )
    )
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  scale_color_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  # Leave room for brackets and significance labels
  scale_y_continuous(
    expand = expansion(
      mult = c(0.05, 0.20)
    )
  ) +
  
  labs(
    x = NULL,
    y = "Log2-normalized protein intensity",
    title = paste(
      "Propanediol metabolism and associated proteins",
      "from P. muris"
    )
  ) +
  
  theme_classic(base_size = 20) +
  
  theme(
    # Main title
    plot.title = element_text(
      size = 25,
      face = "bold",
      hjust = 0.5,
      lineheight = 1.05,
      margin = margin(b = 15)
    ),
    
    # Gray facet ribbons
    strip.background = element_rect(
      fill = "grey90",
      color = "black",
      linewidth = 0.7
    ),
    
    # Wrapped facet text
    strip.text = element_text(
      color = "black",
      face = "bold",
      size = 17,
      lineheight = 1,
      margin = margin(
        t = 9,
        r = 8,
        b = 9,
        l = 8
      )
    ),
    
    # Axis text
    axis.text.x = element_text(
      color = "black",
      size = 23,
      face = "bold"
    ),
    
    axis.text.y = element_text(
      color = "black",
      size = 20
    ),
    
    # Axis title
    axis.title.y = element_text(
      color = "black",
      size = 23,
      face = "bold",
      margin = margin(r = 12)
    ),
    
    # Legend text
    legend.position = "top",
    legend.title = element_text(
      size = 23,
      face = "bold"
    ),
    legend.text = element_text(
      size = 23,
      face = "bold"
    ),
    
    # Spacing between facets
    panel.spacing = grid::unit(1.2, "lines")
  )

protein_boxplot
```

Save plots
```{r}
ggsave(
  filename = "PM_protein_boxplots2.pdf",
  width = 10,
  height = 18,
  dpi = 300
)
```

---------------------- GH family heatmap (Fig. S5B & 5E) ----------------------
Plot select data as a heatmap with replicates per group
Load in packages
```{r}
library(readxl)
library(dplyr)
library(ComplexHeatmap)
library(circlize)
library(viridis)
library(grid)
library(gridtext)
library(stringr)
library(scales)
```

Load in data from excel spread sheets
```{r}
rep_df <- read_excel("f_li_imputed_values.xlsx")
metadat_df <- read_excel("female_li_results_w_function.xlsx", sheet = "complx_carb_ferm_selected")
```

Filter imputed values to only proteins of interest for plotting
```{r}
filtered_rep_df <- rep_df %>%
  filter(Protein %in% metadat_df$Protein)
# Match order between sheets to align metadata (safer ordering)
filtered_rep_df <- filtered_rep_df %>%
  slice(match(metadat_df$Protein, Protein))
```

Create dataframe that will be plotted as a heatmap
```{r}
heatmap_data <- filtered_rep_df %>%
  select(starts_with("LI_")) %>%
  as.data.frame()

# Row-wise z-score scaling
heatmap_scaled <- t(scale(t(heatmap_data)))

# Assign row names
rownames(heatmap_scaled) <- metadat_df$name
```

Order data by GH family
```{r}
# Manually define GH family order
gh_order <- c(
  "GH1",
  "GH2",
  "GH3",
  "GH13",
  "GH29",
  "GH30",
  "GH31",
  "GH32",
  "GH36",
  "GH38",
  "GH42",
  "GH43",
  "GH51"
)
# Convert to a factor
metadat_df$`GH family` <- factor(
  metadat_df$`GH family`,
  levels = gh_order
)
# Order heatmap rows
order_idx <- order(metadat_df$`GH family`)

heatmap_scaled <- heatmap_scaled[order_idx, ]
metadat_df <- metadat_df[order_idx, ]
```

Create italicized labels
```{r}
formatted_labels <- sapply(metadat_df$name, function(x) {

  parts <- strsplit(x, " - ")[[1]]

  species <- parts[1]

  remainder <- ifelse(length(parts) > 1,
                      paste(parts[-1], collapse = " - "),
                      "")

  paste0("<i>", species, "</i> - ", remainder)
})
```

Annotate columns
```{r}
group <- c(rep("Control", 4), rep("Stress", 4))

ha_col <- HeatmapAnnotation(
  Group = group,
  
  col = list(
    Group = c(
      Control = "#4D4D4D",
      Stress = "#FE8A13"
    )
  ),
  
  border = TRUE
)
```

GH family coloring
```{r}
gh_families <- levels(droplevels(metadat_df$`GH family`))

set.seed(1)

gh_colors <- structure(
  sample(rainbow(length(gh_families), s = 0.7, v = 0.85)),
  names = as.character(gh_families)
)

gh_legend <- Legend(
  title = "GH Family",
  labels = names(gh_colors),
  legend_gp = gpar(fill = gh_colors),
  ncol = 1
)
```

Row annotation
```{r}
ha_row <- rowAnnotation(
  GH = metadat_df$`GH family`,

  col = list(GH = gh_colors),

  border = TRUE, 
  
  show_annotation_name = FALSE,

  show_legend = FALSE,

  annotation_width = unit(3, "mm")
)
```

Plot the heatmap using viridis
```{r}
ht <- Heatmap(
  heatmap_scaled,

  width = unit(5, "cm"),
  
  name = "Z-score",

  col = viridis(100),

  top_annotation = ha_col,
  left_annotation = ha_row,

  cluster_rows = FALSE,
  cluster_columns = TRUE,

  row_split = metadat_df$`GH family`,

  row_title = NULL,

  show_row_names = TRUE,

  row_labels = gt_render(formatted_labels),

  row_names_gp = gpar(fontsize = 10),

  column_names_gp = gpar(fontsize = 10),

  border = TRUE
)
```

Draw heatmap
```{r}
draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(gh_legend)
)
```

Save heat map as a PDF
```{r}
pdf(
  "GH_family_heatmap_female_li.pdf",
  width = 12,
  height = 6
)

draw(
  ht,
  heatmap_legend_side = "left",
  annotation_legend_side = "left",
  annotation_legend_list = list(gh_legend)
)

dev.off()
```

---------------------- Arabinose isomerases (Fig. 5F) ----------------------
```{r}
library(readxl)
library(tidyverse)
library(ggnewscale)
library(ggpubr)
```

Replicate data
```{r}
# Create data frame: one row per arabinose isomerase protein
arab_iso_wide <- data.frame(
  Protein = paste0("Arab_iso_", 1:4),
  
  LI_1R = c(26.59014836, 22.86847361, 25.23974988, 22.86847361),
  LI_3R = c(24.79358207, 24.47970457, 22.85565239, 21.33416101),
  LI_3L = c(24.4231736,  22.78263954, 26.44463167, 24.22227051),
  LI_3N = c(25.25789375, 23.64587049, 26.58626664, 23.34719595),
  
  LI_5 = c(16.66645673, 15.73025679, 18.50922405, 11.03458458),
  LI_6 = c(16.56972194, 16.60577156, 19.67905462, 13.64283085),
  LI_7 = c(17.08305329, 23.01987789, 21.02644166, 12.81229335),
  LI_12 = c(24.74865467, 15.79208467, 18.56204975, 12.69349201)
)

# Convert to long format
arab_iso_long <- arab_iso_wide %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Log2_Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress")),
    Protein = factor(Protein, levels = paste0("Arab_iso_", 1:4))
  )
```

Connected mean summary lines
```{r}
arab_iso_summary <- arab_iso_long %>%
  group_by(Protein, Group) %>%
  summarise(
    Mean_Log2_Intensity = mean(Log2_Intensity, na.rm = TRUE),
    .groups = "drop"
  )
```

Plot
```{r}
arab_iso_connected_plot <- ggplot() +
  
  # Replicate dots: original darker colors
  geom_jitter(
    data = arab_iso_long,
    aes(
      x = Group,
      y = Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    width = 0.16,
    size = 5,
    stroke = 0.4,
    alpha = 0.95
  ) +
  
    scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  
  ggnewscale::new_scale_fill() +
  
  # Faint connected protein-level mean lines
  geom_line(
    data = arab_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      group = Protein
    ),
    color = "gray45",
    linewidth = 0.7,
    alpha = 0.5
  ) +
  
  # Mean points: lighter colors and drawn last
  geom_point(
    data = arab_iso_summary,
    aes(
      x = Group,
      y = Mean_Log2_Intensity,
      fill = Group
    ),
    shape = 21,
    color = "black",
    size = 6.5,
    stroke = 0.7,
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  labs(
    title = "L-arabinose isomerase LI F",
    x = NULL,
    y = "Log2 intensity"
  ) +
  
  theme_classic(base_size = 20) +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 25
    ),
    axis.text.x = element_text(
      color = "black",
      face = "bold",
      size = 23
    ),
    axis.text.y = element_text(
      color = "black",
      size = 23
    ),
    axis.title.y = element_text(
      color = "black",
      face = "bold",
      size = 25
    ),
    legend.position = "none"
  )

arab_iso_connected_plot
```

Save plot
```{r}
ggsave(
  filename = "f_li_arabinose_isomerase_plot.pdf",
  plot = arab_iso_connected_plot,
  width = 6,
  height = 6
)
```

---------------------- Xylose isomerases (Fig. 5G) ----------------------
```{r}
library(readxl)
library(tidyverse)
library(ggnewscale)
library(ggpubr)
```

Create data frame
```{r}
xylose_iso_wide <- data.frame(
  Protein = c(
    "A. muris",
    "A. chartisolvens"
  ),
  
  LI_1R = c(22.76354405, 10.10437185),
  LI_3R = c(22.85565239, 12.84485678),
  LI_3L = c(21.9862123,  8.472315183),
  LI_3N = c(21.98477853, 10.01343091),
  
  LI_5 = c(15.67163703, 22.76231867),
  LI_6 = c(16.64836058, 22.83457254),
  LI_7 = c(19.46889062, 23.24084031),
  LI_12 = c(16.59677043, 23.26018572)
)

# Convert to long format
xylose_iso_long <- xylose_iso_wide %>%
  pivot_longer(
    cols = starts_with("LI_"),
    names_to = "Sample",
    values_to = "Log2_Intensity"
  ) %>%
  mutate(
    Group = case_when(
      str_detect(Sample, "Con") ~ "Control",
      str_detect(Sample, "Strs") ~ "Stress",
      TRUE ~ NA_character_
    ),
    Group = factor(Group, levels = c("Control", "Stress")),
    Protein = factor(
      Protein,
      levels = c(
        "A. muris",
        "A. chartisolvens"
      )
    )
  )
```

Create a sig bracket dataframe
```{r}
bracket_df <- xylose_iso_long %>%
  group_by(Protein) %>%
  summarise(
    y.position = max(Log2_Intensity, na.rm = TRUE) + 1,
    .groups = "drop"
  ) %>%
  mutate(
    group1 = "Control",
    group2 = "Stress",
    label = c("**", "***")
  )
```

Plot
```{r}
xylose_barplot <- ggplot(
  xylose_iso_long,
  aes(x = Group, y = Log2_Intensity, fill = Group)
) +
  stat_summary(
    fun = mean,
    geom = "bar",
    width = 0.65,
    color = "black",
    linewidth = 0.6,
    alpha = 0.85
  ) +
  stat_summary(
    fun.data = mean_se,
    geom = "errorbar",
    width = 0.18,
    linewidth = 0.6,
    color = "black"
  ) +
  
  # Control replicate dots: darker gray with black border
  geom_jitter(
    data = xylose_iso_long %>% filter(Group == "Control"),
    aes(x = Group, y = Log2_Intensity),
    inherit.aes = FALSE,
    shape = 21,
    fill = "#4D4D4D",
    color = "black",
    stroke = 0.4,
    width = 0.08,
    size = 4.5,
    alpha = 0.95
  ) +
  
  # Stress replicate dots: darker orange with black border
  geom_jitter(
    data = xylose_iso_long %>% filter(Group == "Stress"),
    aes(x = Group, y = Log2_Intensity),
    inherit.aes = FALSE,
    shape = 21,
    fill = "#FE8A13",
    color = "black",
    stroke = 0.4,
    width = 0.08,
    size = 4.5,
    alpha = 0.95
  ) +
  
    stat_pvalue_manual(
    bracket_df,
    label = "label",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.6,
    size = 8,
    inherit.aes = FALSE
  ) +
  
  facet_wrap(
    ~ Protein,
    scales = "free_y",
    nrow = 1
  ) +
  scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  labs(
    x = NULL,
    y = "Log2 intensity",
    title = "Xylose isomerase LI F"
  ) +
  theme_classic(base_size = 20) +
  theme(
    strip.background = element_rect(
      fill = "gray90",
      color = "black",
      linewidth = 0.5
    ),
    strip.text = element_text(
      face = "italic",
      color = "black",
      size = 18
    ),
    axis.text.x = element_text(
      color = "black",
      face = "bold",
      size = 23,
      hjust = 1,
      angle = 45
    ),
    axis.text.y = element_text(color = "black", size = 23),
    axis.title.y = element_text(color = "black", face = "bold", size = 25),
    plot.title = element_text(
      face = "bold",
      hjust = 0.5,
      size = 25
    ),
    legend.position = "none",
    panel.border = element_rect(
      color = "black",
      fill = NA,
      linewidth = 0.6
    )
  )

xylose_barplot
```

Save plot
```{r}
ggsave(
  filename = "f_li_xylose_isomerase_plot.pdf",
  plot = xylose_barplot,
  width = 6,
  height = 6
)
```
