---
title: "metabolomics_workup_and_analysis_NEG"
author: "Amelia Slusser"
date: "2026-06-29"
output: html_document
---
#############################################################################
Volcano plot with selected metabolites highlighted
#############################################################################
```{r}
library(tidyverse)
library(readxl)
library(ggrepel)
library(writexl)
```

Load metabolomics data
```{r}
metab_df <- read_excel("serum_f_neg_compounds.xlsx", sheet = "log2")

# Clean column names slightly while preserving meaningful names
colnames(metab_df) <- colnames(metab_df) %>%
  str_squish()

# Define sample columns
control_cols <- colnames(metab_df) %>%
  keep(~ str_detect(.x, "SE_CON"))

stress_cols <- colnames(metab_df) %>%
  keep(~ str_detect(.x, "SE_STR"))

blank_cols <- colnames(metab_df) %>%
  keep(~ str_detect(.x, "BLNK"))

qc_cols <- colnames(metab_df) %>%
  keep(~ str_detect(.x, "QC"))

# Check that the correct columns were found
control_cols
stress_cols
blank_cols
qc_cols
```

Define metabolites of interest
```{r}
glucuronides <- c(
  "p-Cresol glucuronide",
  "5-Hydroxy-6-methoxyindoleglucuronide",
  "Acetaminophen glucuronide",
  "1-Salicylateglucuronide",
  "Pyrogallol-2-O-glucuronide",
  "Indoxyl-beta-D-glucuronide",
  "Cholicacidglucuronide",
  "Benzoylglucuronide(Benzoicacid)",
  "Estroneglucuronide",
  "3-Methoxy-4-hydroxyphenylglycolglucuronide",
  "4-Hydroxy-5-(3′,4′-dihydroxyphenyl)-valeric acid-O-glucuronide",
  "Nicotine glucuronide",
  "3-indole carboxylic acid glucuronide",
  "Tyramineglucuronide",
  "D-Glucuronic acid",
  "[Similar to: β-D-Glucopyranuronic acid; ΔMass: 208.1209 Da]",
  "β-D-Glucopyranuronic acid",
  "[Similar to: β-D-Glucopyranuronic acid; ΔMass: 194.1053 Da]",
  "1-(α-D-Glucopyranosyluronosyl)-3-[(2S)-1-methyl-5-oxo-2-pyrrolidinyl]pyridinium",
  "8-Hydroxy-2-quinolinyl β-D-glucopyranosiduronic acid",
  "5-(2-Carboxyethyl)-2-hydroxyphenyl β-D-glucopyranosiduronic acid",
  "4-[(E)-2-Carboxyvinyl]-2-methoxyphenyl β-D-glucopyranosiduronic acid"
) %>%
  unique()
```

Calculate statistics
```{r}
# Make sure sample columns are numeric
sample_cols <- c(control_cols, stress_cols)

metab_df <- metab_df %>%
  mutate(
    across(
      all_of(sample_cols),
      ~ as.numeric(.x)
    )
  )

# Calculate the statistics
volcano_df <- metab_df %>%
  rowwise() %>%
  mutate(
    Control_Mean = mean(c_across(all_of(control_cols)), na.rm = TRUE),
    Stress_Mean  = mean(c_across(all_of(stress_cols)), na.rm = TRUE),
    
    log2FC = Stress_Mean - Control_Mean,
    
    p_value = tryCatch(
      {
        control_values <- c_across(all_of(control_cols))
        stress_values  <- c_across(all_of(stress_cols))
        
        control_values <- control_values[!is.na(control_values)]
        stress_values  <- stress_values[!is.na(stress_values)]
        
        if (
          length(control_values) >= 2 &&
          length(stress_values) >= 2 &&
          sd(control_values) > 0 &&
          sd(stress_values) > 0
        ) {
          t.test(
            x = stress_values,
            y = control_values
          )$p.value
        } else {
          NA_real_
        }
      },
      error = function(e) NA_real_
    )
  ) %>%
  ungroup() %>%
  mutate(
    adj_p_value = p.adjust(p_value, method = "BH"),
    
    # prevents Inf if any p-value is exactly 0
    neg_log10_p = -log10(pmax(p_value, .Machine$double.xmin))
  )
```

Classify metabolites for plotting
```{r}
p_cutoff <- 0.05
log2FC_cutoff <- 0.5

volcano_df <- volcano_df %>%
  mutate(
    Metabolite_Group = case_when(
      Name %in% glucuronides ~ "Glucuronide",
      p_value < p_cutoff & abs(log2FC) > log2FC_cutoff ~ "Significant",
      TRUE ~ "Not significant"
    ),
    Metabolite_Group = factor(
      Metabolite_Group,
      levels = c(
        "Not significant",
        "Significant",
        "Glucuronide"
      )
    )
  )
```

Check which compounds were actually selected
```{r}
unmatched_glucuronides <- setdiff(glucuronides, volcano_df$Name)
unmatched_aglycones <- setdiff(aglycones, volcano_df$Name)

unmatched_glucuronides
unmatched_aglycones
```

Export annotated data frame
```{r}
write_xlsx(volcano_df, "f_serum_annotated.xlsx")
```

Plot
```{r}
volcano_plot <- ggplot(
  volcano_df,
  aes(x = log2FC, y = neg_log10_p)
) +
  
  # Background metabolites: no outline
  geom_point(
    data = volcano_df %>%
      filter(Metabolite_Group %in% c("Not significant", "Significant")),
    aes(color = Metabolite_Group),
    shape = 16,
    size = 3,
    alpha = 0.85
  ) +
  
  # Glucuronides: filled with black outline
  geom_point(
    data = volcano_df %>%
      filter(Metabolite_Group == "Glucuronide"),
    aes(fill = Metabolite_Group),
    shape = 21,
    color = "black",
    stroke = 0.5,
    size = 5,
    alpha = 1
  ) +
  
  # Raw p-value cutoff line
  geom_hline(
    yintercept = -log10(p_cutoff),
    linetype = "dashed",
    linewidth = 0.5,
    color = "gray40"
  ) +
  
  # log2FC cutoff lines
  geom_vline(
    xintercept = c(-log2FC_cutoff, log2FC_cutoff),
    linetype = "dashed",
    linewidth = 0.5,
    color = "gray40"
  ) +
  
  scale_color_manual(
    values = c(
      "Not significant" = "#D9D9D9",
      "Significant" = "#7F7F7F"
    ),
    name = NULL
  ) +
  
  scale_fill_manual(
    values = c(
      #"Aglycone" = "#E0B0FF",
      "Glucuronide" = "#98FBCB"
    ),
    name = NULL
  ) +
  
 labs(
  x = expression(
    bold(Log[2]~Fold~Change~"(Stress / Control)")
  ),
  y = expression(
    bold(-Log[10]~"p-value")
  ),
  title = "SE F"
) +

theme_classic(base_size = 23) +

theme(
  plot.title = element_text(
    hjust = 0.5,
    face = "bold",
    size = 25
  ),
  axis.title.x = element_text(
    face = "bold",
    size = 23
  ),
  axis.title.y = element_text(
    face = "bold",
    size = 23
  ),
  legend.position = "right"
)

volcano_plot
```

Save the plot
```{r}
ggsave(
  filename = "f_serum_meta_glucuronides.pdf",
  plot = volcano_plot,
  width = 10,
  height = 6,
  units = "in"
)
```

#############################################################################
Estrogen search analysis
#############################################################################

```{r}
library(tidyverse)
library(readxl)
library(openxlsx)
library(ggpubr)
library(ggnewscale)
```

Load in feature table
```{r}
feature_df <- read_excel("serum_f_neg_compounds.xlsx")
```

Define estrogen/metabolite exact-mass search list
This list focuses on negative-mode [M-H]- masses, which I would prioritize for estrogen sulfates, glucuronides, and phenolic estrogens.
```{r}
estrogen_targets_neg <- tribble(
  ~Target_Name, ~Class, ~Formula, ~Adduct, ~Target_mz,
  
  # Free estrogens
  "Estrone", "Free estrogen", "C18H22O2", "[M-H]-", 269.154703,
  "17beta-Estradiol", "Free estrogen", "C18H24O2", "[M-H]-", 271.170354,
  "Estriol", "Free estrogen / hydroxyestrogen isobar", "C18H24O3", "[M-H]-", 287.165268,
  
  # Hydroxylated estrogens
  "Hydroxyestrone isomer", "Hydroxy estrogen", "C18H22O3", "[M-H]-", 285.149618,
  "Hydroxyestradiol isomer", "Hydroxy estrogen / estriol isobar", "C18H24O3", "[M-H]-", 287.165268,
  
  # Methoxy estrogens
  "Methoxyestrone isomer", "Methoxy estrogen", "C19H24O3", "[M-H]-", 299.165268,
  "Methoxyestradiol isomer", "Methoxy estrogen", "C19H26O3", "[M-H]-", 301.180918,
  
  # Sulfated estrogens
  "Estrone sulfate", "Sulfated estrogen", "C18H22O5S", "[M-H]-", 349.111519,
  "Estradiol sulfate", "Sulfated estrogen", "C18H24O5S", "[M-H]-", 351.127169,
  "Estriol sulfate", "Sulfated estrogen", "C18H24O6S", "[M-H]-", 367.122083,
  
  # Glucuronidated estrogens
  "Estrone glucuronide", "Glucuronidated estrogen", "C24H30O8", "[M-H]-", 445.186791,
  "Estradiol glucuronide", "Glucuronidated estrogen", "C24H32O8", "[M-H]-", 447.202442,
  "Estriol glucuronide", "Glucuronidated estrogen", "C24H32O9", "[M-H]-", 463.197356
)
```

Set ppm tolerance within 5ppm
```{r}
ppm_tol <- 5

estrogen_mass_hits <- feature_df %>%
  mutate(
    mz_observed = as.numeric(`m/z`),
    rt_min = as.numeric(`RT [min]`)
  ) %>%
  crossing(estrogen_targets_neg) %>%
  mutate(
    ppm_error = ((mz_observed - Target_mz) / Target_mz) * 1e6,
    abs_ppm_error = abs(ppm_error)
  ) %>%
  filter(abs_ppm_error <= ppm_tol) %>%
  arrange(Target_Name, abs_ppm_error)
```

Check number of hits
```{r}
estrogen_mass_hits %>%
  count(Target_Name, Class, Adduct, sort = TRUE)
```

View main candidate table
```{r}
main_candidate_table <- estrogen_mass_hits %>%
  select(
    any_of(c(
      "Name",
      "Calc. MW",
      "m/z",
      "RT [min]",
      "MS2",
      "Reference Ion",
      "Annot. Source: Predicted Compositions",
      "Annot. Source: mzCloud Search",
      "Annot. Source: Metabolika Search",
      "Annot. Source: ChemSpider Search",
      "Annot. Source: MassList Search",
      "Mass List Match: Endogenous Metabolites database 4400 compounds",
      "Mass List Match: HMDB Feces metabolites (v5. 6790 cpds)",
      "Peak Rating (Max.)",
      "Peak Rating QC (Max.)"
    )),
    Target_Name,
    Class,
    Target_mz,
    mz_observed,
    ppm_error,
    abs_ppm_error,
    rt_min
  ) %>%
  arrange(Target_Name, abs_ppm_error)

main_candidate_table
```

Identify all per-sample are columns
```{r}
area_cols <- names(estrogen_mass_hits) %>%
  str_subset("^Area: .*\\.raw")

area_cols
```

Convert candidate hits to long format
```{r}
estrogen_long <- estrogen_mass_hits %>%
  select(
    any_of(c(
      "Name",
      "Calc. MW",
      "m/z",
      "RT [min]",
      "MS2",
      "Reference Ion",
      "Peak Rating (Max.)",
      "Peak Rating QC (Max.)"
    )),
    Target_Name,
    Class,
    Target_mz,
    mz_observed,
    ppm_error,
    abs_ppm_error,
    rt_min,
    all_of(area_cols)
  ) %>%
  pivot_longer(
    cols = all_of(area_cols),
    names_to = "Area_Column",
    values_to = "Area"
  ) %>%
  mutate(
    Raw_File = Area_Column %>%
      str_remove("^Area: ") %>%
      str_remove(" \\(.*\\)$"),
    
    Sample_Type = case_when(
      str_detect(Raw_File, "^BLNK") ~ "Blank",
      str_detect(Raw_File, "^QC") ~ "QC",
      TRUE ~ "Sample"
    ),
    
    Group = case_when(
      str_detect(Raw_File, "_CON_") ~ "Control",
      str_detect(Raw_File, "_STR_") ~ "Stress",
      TRUE ~ NA_character_
    ),
    
    Ion_Mode = case_when(
      str_detect(Raw_File, "_N\\.raw$") ~ "Negative",
      str_detect(Raw_File, "_P\\.raw$") ~ "Positive",
      TRUE ~ NA_character_
    ),
    
    Area = as.numeric(Area),
    log2_Area = log2(Area + 1),
    
    Has_MS2 = !is.na(MS2) & MS2 != "" & !str_detect(tolower(MS2), "no")
  )

# Check
glimpse(estrogen_long)

estrogen_long %>%
  count(Sample_Type, Group, Ion_Mode)
```

Summarize blank, QC, and sample signal
```{r}
# Key filtering step
estrogen_qc_summary <- estrogen_long %>%
  group_by(
    Name,
    Target_Name,
    Class,
    Target_mz,
    mz_observed,
    ppm_error,
    abs_ppm_error,
    rt_min,
    MS2,
    Has_MS2,
    Sample_Type
  ) %>%
  summarise(
    n_nonmissing = sum(!is.na(Area)),
    mean_area = mean(Area, na.rm = TRUE),
    median_area = median(Area, na.rm = TRUE),
    sd_area = sd(Area, na.rm = TRUE),
    cv_percent = 100 * sd_area / mean_area,
    .groups = "drop"
  )

# Convert to wide summary table
estrogen_qc_wide <- estrogen_qc_summary %>%
  pivot_wider(
    names_from = Sample_Type,
    values_from = c(mean_area, median_area, sd_area, cv_percent, n_nonmissing),
    names_sep = "_"
  ) %>%
  mutate(
    sample_to_blank_ratio = case_when(
      is.na(mean_area_Blank) | mean_area_Blank == 0 ~ Inf,
      TRUE ~ mean_area_Sample / mean_area_Blank
    ),
    
    qc_to_blank_ratio = case_when(
      is.na(mean_area_Blank) | mean_area_Blank == 0 ~ Inf,
      TRUE ~ mean_area_QC / mean_area_Blank
    ),
    
    Candidate_Flag = case_when(
      is.na(mean_area_Sample) ~ "No sample signal",
      is.na(mean_area_Blank) | mean_area_Blank == 0 ~ "Good: no blank signal detected",
      sample_to_blank_ratio >= 5 & qc_to_blank_ratio >= 5 ~ "Good: sample/QC >5x blank",
      sample_to_blank_ratio >= 3 & qc_to_blank_ratio >= 3 ~ "Moderate: sample/QC >3x blank",
      TRUE ~ "Caution: possible blank/background issue"
    )
  ) %>%
  arrange(Target_Name, abs_ppm_error)

# View candidate summary
estrogen_qc_wide %>%
  select(
    Name,
    Target_Name,
    Class,
    mz_observed,
    Target_mz,
    ppm_error,
    rt_min,
    MS2,
    Has_MS2,
    mean_area_Blank,
    mean_area_QC,
    mean_area_Sample,
    sample_to_blank_ratio,
    qc_to_blank_ratio,
    cv_percent_QC,
    Candidate_Flag
  )
```

Prioritized candidate list
```{r}
estrogen_priority_hits <- estrogen_qc_wide %>%
  filter(
    abs_ppm_error <= 5,
    !is.na(mean_area_Sample),
    is.na(mean_area_Blank) | mean_area_Blank == 0 | sample_to_blank_ratio >= 3
  ) %>%
  arrange(
    desc(Has_MS2),
    abs_ppm_error,
    cv_percent_QC
  )
estrogen_priority_hits

# Stricter list
estrogen_priority_hits_qc_filtered <- estrogen_priority_hits %>%
  filter(
    is.na(cv_percent_QC) | cv_percent_QC <= 30
  )
estrogen_priority_hits_qc_filtered
```

Prepare plotting data
```{r}
plot_df <- estrogen_long %>%
  unite(
    Candidate_ID,
    Target_Name,
    mz_observed,
    rt_min,
    remove = FALSE,
    sep = " | "
  ) %>%
  filter(
    Sample_Type == "Sample",
    Target_Name == "Estrone glucuronide",
    Candidate_ID %in% (
      estrogen_priority_hits %>%
        filter(Target_Name == "Estrone glucuronide") %>%
        unite(
          Candidate_ID,
          Target_Name,
          mz_observed,
          rt_min,
          remove = FALSE,
          sep = " | "
        ) %>%
        pull(Candidate_ID)
    )
  ) %>%
  mutate(
    Group = factor(Group, levels = c("Control", "Stress")),
    
    # This keeps the actual panels separate
    Candidate_Panel = Candidate_ID,
    
    # This is what will show in the gray facet strip
    Feature_Label = "Estrone glucuronide candidate feature"
  )
```

Statistics
```{r}
estrogen_stats <- plot_df %>%
  group_by(
    Candidate_Panel,
    Feature_Label,
    Target_Name,
    Class,
    mz_observed,
    Target_mz,
    ppm_error,
    rt_min
  ) %>%
  summarise(
    n_control = sum(Group == "Control" & !is.na(log2_Area)),
    n_stress  = sum(Group == "Stress" & !is.na(log2_Area)),
    
    control_mean = mean(log2_Area[Group == "Control"], na.rm = TRUE),
    stress_mean  = mean(log2_Area[Group == "Stress"], na.rm = TRUE),
    
    log2FC_stress_vs_control = stress_mean - control_mean,
    
    p_value = ifelse(
      n_control >= 2 & n_stress >= 2,
      wilcox.test(log2_Area ~ Group)$p.value,
      NA_real_
    ),
    .groups = "drop"
  ) %>%
  mutate(
    sig_label = case_when(
      is.na(p_value) ~ "not tested",
      p_value < 0.001 ~ "***",
      p_value < 0.01  ~ "**",
      p_value < 0.05  ~ "*",
      TRUE ~ "ns"
    ),
    
    label = case_when(
      is.na(p_value) ~ "not tested",
      TRUE ~ paste0(
        sig_label,
        "\n",
        "p = ", signif(p_value, 3)
      )
    )
  )
```

Plot
```{r}
# Create brackets and positioning
bracket_df <- estrogen_stats %>%
  select(
    Candidate_Panel,
    Feature_Label,
    p_value,
    sig_label,
    label
  ) %>%
  left_join(
    plot_df %>%
      group_by(Candidate_Panel, Feature_Label) %>%
      summarise(
        y_max = max(log2_Area, na.rm = TRUE),
        y_min = min(log2_Area, na.rm = TRUE),
        .groups = "drop"
      ),
    by = c("Candidate_Panel", "Feature_Label")
  ) %>%
  mutate(
    y_range = y_max - y_min,
    y.position = ifelse(
      y_range == 0,
      y_max + 0.5,
      y_max + 0.15 * y_range
    ),
    group1 = "Control",
    group2 = "Stress"
  )

# Plot
# Make a labeller so every separate panel displays the same clean label
facet_labels <- setNames(
  rep("Estrone glucuronide candidate feature", length(unique(plot_df$Candidate_Panel))),
  unique(plot_df$Candidate_Panel)
)

estrogen_candidate_plot_brackets <- ggplot(
  plot_df,
  aes(x = Group, y = log2_Area, fill = Group)
) +
  
  geom_boxplot(
    width = 0.6,
    outlier.shape = NA,
    color = "black",
    alpha = 0.85
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#BDBDBD",
      "Stress"  = "#FDBA74"
    )
  ) +
  
  ggnewscale::new_scale_fill() +
  
  geom_jitter(
    aes(fill = Group),
    shape = 21,
    color = "black",
    stroke = 0.35,
    width = 0.12,
    size = 2.4,
    alpha = 0.95,
    show.legend = FALSE
  ) +
  
  scale_fill_manual(
    values = c(
      "Control" = "#4D4D4D",
      "Stress"  = "#FE8A13"
    )
  ) +
  
  stat_pvalue_manual(
    bracket_df,
    label = "label",
    xmin = "group1",
    xmax = "group2",
    y.position = "y.position",
    tip.length = 0.01,
    bracket.size = 0.5,
    size = 3.2,
    hide.ns = FALSE
  ) +
  
  facet_wrap(
    ~ Candidate_Panel,
    scales = "free_y",
    labeller = labeller(Candidate_Panel = facet_labels)
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0.05, 0.18))
  ) +
  
  coord_cartesian(clip = "off") +
  
  theme_classic() +
  
  labs(
    x = NULL,
    y = "Log2(area + 1)",
    title = "Exact-mass estrogen/metabolite candidate features"
  ) +
  
  theme(
    legend.position = "none",
    strip.text = element_text(
      size = 9,
      color = "black",
      face = "bold"
    ),
    strip.background = element_rect(
      fill = "#D9D9D9",
      color = "black",
      linewidth = 0.5
    ),
    plot.title = element_text(hjust = 0.5),
    plot.margin = margin(10, 20, 10, 10)
  )

estrogen_candidate_plot_brackets
```

Save the data that goes into the plot
```{r}
plot_df_export <- plot_df %>%
  select(
    Raw_File,
    Group,
    Target_Name,
    Name,
    mz_observed,
    rt_min,
    Area,
    log2_Area,
    Has_MS2
  )

write.xlsx(
  plot_df_export,
  file = "Estrone_glucuronide_boxplot_data.xlsx",
  overwrite = TRUE
)
```

Save the plot
```{r}
ggsave(
  filename = "f_serum_estrone_gluc.pdf",
  plot = estrogen_candidate_plot_brackets,
  width = 8,
  height = 6,
  units = "in"
)
```